---
title: "AgeSex"
format: html
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Setup


```{r}
#| message: false
library(tidyverse)
library(sf)
library(corrr)
library(car)
library(spdep)
library(zeallot)
library(rstatix)
library(GGally)
library(ggpubr)
library(rcompanion)
source("demographic-functions.R")
```

```{r}
process <- function(df, var) {
  df %>%
    st_drop_geometry() %>%
    group_by_at(c("geoid", "district", var)) %>%
    summarise(value = sum(value)) %>%
    pivot_wider(names_from = all_of(var), values_from = value) %>% 
    ungroup()
}
variables <- c("agegroup", "education", "hisp")
layers <- c("age_sex_dist_sf", "edu_dist_sf", "race_dist_sf")

c(abq_age, abq_edu, abq_race) %<-%
  map2(
    layers,
    variables,
    \(layer, variable) st_read("data/abq_demographics.gpkg", layer = layer) %>%
      process(variable) %>% 
      filter(district != "Los Ranchos")
  )
```

# Age Levels

```{r}
under_20_cols <- names(abq_age)[c(3:5, 11, 16)]
dist_levels <- unique(abq_age$district) %>% 
  sort()
abq_age <- abq_age %>% 
  mutate(district = factor(district, levels = dist_levels),
         Total = rowSums(across(where(is.numeric)), na.rm = T),
         Under.20 = rowSums(across(all_of(under_20_cols)), na.rm = T),
         Under.20.pct = Under.20 / Total)

summary(abq_age)
```

```{r}
abq_age %>% 
  ggpubr::ggboxplot("district", "Under.20.pct", fill = "district")
```

```{r}
result <- aov(Under.20.pct ~ district, data = abq_age)
summary(result)
```

```{r}
(result_tk <- TukeyHSD(result))
```

District 3-District 2, District 4-District 3, District 6-District 3, District 7-District 3,
District 8-District 3, District 8-District 3

```{r}
c(district = , Under.20.pct = ) %<-%
  abq_age

bartlett.test(Under.20.pct, district)
```

The Elfving transformation is qnorm((rank(x) - pi/8)/(N - pi/4 + 1)), where x is the original data, N is the number of observations, and pi is the constant pi. – 
Sal Mangiafico


```{r}
glm(formula = Under.20.pct ~ district, family = "quasibinomial",
    data = abq_age) %>% summary()
```

```{r}
abq_age %>% 
  shapiro_test(Under.20.pct)
```

```{r}
abq_age %>%
  shapiro_test(Total)
```

```{r}
ks.test(abq_age$Under.20.pct, "pnorm")
```

```{r}
plotNormalHistogram(abq_age$Under.20.pct)
```

```{r}
ggqqplot(abq_age$Under.20.pct)
```

```{r}
abq_age$Under.20.pct.NST <- blom(abq_age$Under.20.pct)
plotNormalHistogram(abq_age$Under.20.pct.NST)
```


```{r}
ggqqplot(abq_age$Over.54.pct.NST)
```

```{r}
abq_age %>% 
  shapiro_test(Under.20.pct.NST)
```

```{r}
abq_age %>% 
  ggpubr::ggboxplot("district", "Under.20.pct.NST", fill = "district")
```

```{r}
result <- aov(Under.20.pct.NST ~ district, data = abq_age)
summary(result)
```

```{r}
(result_tk <- TukeyHSD(result))
```

District 3-District 2, District 4-District 3, District 6-District 3, District 7-District 3,
District 8-District 3, District 8-District 3

```{r}
c(district = , Under.20.pct.NST = ) %<-%
  abq_age

bartlett.test(Under.20.pct.NST, district)
```

```{r}
summary(abq_age$Under.20.pct.NST)
```

```{r}
cols_gt_55 <- names(abq_age)[12:15]
abq_age <- abq_age %>% 
  mutate(Over.54 = rowSums(across(all_of(cols_gt_55)), na.rm = T),
         Over.54.pct = Over.54 / Total)
```

```{r}
plotNormalHistogram(abq_age$Over.54.pct)
```

```{r}
shapiro.test(abq_age$Over.54.pct)
```

```{r}
abq_age$Over.54.pct.NST <- blom(abq_age$Over.54.pct)
plotNormalHistogram(abq_age$Over.54.pct.NST)
```


```{r}
shapiro_test(abq_age$Over.54.pct.NST)
```

```{r}
abq_age %>% 
  ggpubr::ggboxplot("district", "Over.54.pct.NST", fill = "district")
```

```{r}
result <- aov(Over.54.pct.NST ~ district, data = abq_age)
summary(result)
```

```{r}
(result_tk <- TukeyHSD(result))
```

```{r}
abq_age$From.18.To.34.pct.NST <- blom(abq_age$From.18.To.34.pct)
plotNormalHistogram(abq_age$From.18.To.34.pct)
shapiro_test(abq_age$From.18.To.34.pct)
```

```{r}
result <- aov(Over.54.pct.NST ~ district, data = abq_age)
summary(result)
```

# Education

```{r}
abq_edu <- st_read("data/abq_demographics.gpkg",
                   layer = "edu_dist_sf")

edu_levels <- unique(abq_edu$education)
abq_edu <- abq_edu %>% 
  process("education") %>% 
  filter(district != "Los Ranchos")
```

```{r}
names(abq_edu)
```

```{r}
col_degree <- names(abq_edu)[c(4, 5, 7, 9)]
no_degree <- names(abq_edu)[c(6, 8)]
abq_edu <- abq_edu %>% 
  mutate(district = factor(district, levels = dist_levels),
         Total = rowSums(across(where(is.numeric))),
         College.Degree = rowSums(across(all_of(col_degree)), na.rm = T),
         College.Degree.pct = Advanced.Degree / Total,
         No.Degree = rowSums(across(all_of(no_degree)), na.rm = T),
         No.Degree.pct = No.Degree / Total)
abq_edu
```

```{r}
shapiro_test(abq_edu$Advanced.Degree.pct)
```


```{r}
abq_edu$Advanced.Degree.pct.NST <- blom(abq_edu$Advanced.Degree.pct)
plotNormalHistogram(abq_edu$Advanced.Degree.pct.NST)
shapiro_test(abq_edu$Advanced.Degree.pct.NST)
```

```{r}
abq_edu$No.Degree.pct.NST <- blom(abq_edu$No.Degree.pct)
plotNormalHistogram(abq_edu$No.Degree.pct.NST)
shapiro_test(abq_edu$No.Degree.pct.NST)
```


```{r}
abq_edu %>% 
  ggpubr::ggboxplot("district", "Advanced.Degree.pct.NST", fill = "district")
```

```{r}
abq_edu %>% 
  ggpubr::ggboxplot("district", "No.Degree.pct.NST", fill = "district")
```

```{r}
result <- aov(Advanced.Degree.pct.NST ~ district, data = abq_edu)
summary(result)
```


```{r}
(result_tk <- TukeyHSD(result))
```

```{r}
result <- aov(No.Degree.pct.NST ~ district, data = abq_edu)
summary(result)
```

```{r}
(result_tk <- TukeyHSD(result))
```



