---
title: "Demographics"
format: 
  html:
    code-fold: false
    code-tools: true
    toc: true
    toc-depth: 2
exec:
  freeze: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

In this article, I would like to look at the city of Albuquerque to compare variation among different parts of the city. To this end, I will break down the city by council districts, of which there are nine, and consider economic factors such as income, housing values, poverty levels and income equality. For this exercise, I will obtain the demographic data from `census.gov` using `tidycensus`. I will get the council districts from the county website, and the census data will need to be apportioned to each district. Many census tracts cross district lines, so data must be appropriately assigned. 

Throughout, I hope to emphasize best practices, such as writing functions and utilizing the mapping functionality of `purrr` to write clear, concise, non-repetitive and easily understandable code. I will also demonstrate the value of visualization in the data cleaning process.

I will use the following libraries:

```{r}
#| message: false
library(tidycensus)
library(tidyverse)
library(ggspatial)
library(sf)
library(units)
library(crsuggest)
library(gt)
library(ggtext)
library(patchwork)
library(janitor)
library(nngeo)
options(tigris_use_cache = TRUE)
```

# City Council Districts

I will start by obtaining the boundaries for the city council districts. The Albuquerque website provides a wealth of data, although it is not necessarily well-structured. In many of the data sets, the relevant data is contained in tables embedded in columns of `html` tables. I'll get to that when I incorporate crime statistics later. In this case, it is useless, as the website only provides councilors' names as a variable, not district numbers, and the names provided are for councilors from several years ago!

Fortunately, the Bernalillo County website provides what's needed, as well as providing the shape for The Village of Los Ranchos de Albuquerque. Los Ranchos is sort of the Beverly Hills of Albuquerque, in the sense that it is a separate municipal enclave (almost) entirely surrounded by Albuquerque, and where many the largest properties and mansions in the area are located.

The data portal for Bernalillo County is not very friendly for programmatic downloading, though. My first effort to download and unzip the file containing the city council boundaries failed to produce a usable file. Adding the argument `method = "curl"` allowed me to download the files, however.

```{r}
#| eval: false
download.file("https://www.berncoclerk.gov/wp-content/uploads/2025/07/BERNCO_CLERKJuly2025.gdb_.zip", "data/bernalillo.zip", method = "curl")
unzip("data/bernalillo.zip", overwrite = T)
```

I will read the file with `sf::st_read`, which will result in an `sf` object. The only variable I'm interested in is the district number, so I'll just grab that, fix the variable name, and make the content more descriptive for easy mapping. The `geometry` is sticky, so I don't need to specify it.

```{r}
council_dists_bc <- st_read("data/BC_CityCouncil/ABQ_CityCouncils.shp") %>%
  select(district = DISTRICTNU) %>%
  mutate(district = paste("District", district))
council_dists_bc
```

I'll do the same for Los Ranchos.

```{r}
#| eval: false
download.file("https://www.berncoclerk.gov/wp-content/uploads/2023/02/LosRanchos.zip", "data/test.zip", method = "curl")
unzip("data/LosRanchos.zip")
```

```{r}
los_ranchos <- st_read("data/LosRanchos/LosRanchos.shp") %>%
  select(district = Name)
```


:::{.callout-tip title="Finding a Coordinte Reference System" collapse="true"}

The `crsuggest` package has a useful function to determine an appropriate coordinate reference system. 

```{r}
suggest_crs(council_dists_bc) %>% head(2)
```

:::

Now I will bind them and transform the coordinate reference system to a local one in meters.

```{r}
council_dists <- rbind(council_dists_bc, los_ranchos) %>%
  st_transform(6528)
council_dists
```

I'll plot the district boundaries with `ggplot`, adding a base map with `ggspatial::annotation_map_tile`. I'll be doing the same map several times, so I'll make a function to do so.

```{r}
dist_plot <- function(df) {
  ggplot(df) +
  annotation_map_tile(
    type = "osm",
    cachedir = "~/.cache/maps/",
    zoomin = -1,
    alpha = 0.7
  ) +
  geom_sf() +
  theme_void() +
  geom_sf(aes(color = district), fill = NA, linewidth = 1) +
  geom_sf_text(aes(label = district)) +
  scale_color_viridis_d() +
  theme_void() +
  labs(title = "Albuquerque Council Districts")
}
```

```{r}
dist_plot(council_dists)
```

One thing to note from this map: there are numerous areas of unincorporated Bernalillo County which lie inside Albuquerque's borders but do not pertain to any city district. A closer look reveals some of these holes.

```{r}
dist_plot(council_dists %>% filter(district %in% c("District 2", "District 4", "District 5", "Los Ranchos")))
```

This will become relevant later as we map the census data onto the districts, as some of the data will lie outside city boundaries. On the other hand, I think it makes sense to fill in the holes in the middle of districts, to minimize the loss of data. I'll use the convenient `st_remove_holes()` function from the `nngeo` package.

```{r}
council_dists <- st_remove_holes(council_dists)
```

We can see that the holes are gone.

```{r}
dist_plot(council_dists)
```

```{r}
dist_plot(council_dists %>% filter(district %in% c("District 2", "District 4", "District 5", "Los Ranchos")))
```


#  Basic Demographics

I want to look at some basic demographic and economic information, to see how they compare between council districts. I will use `tidycensus` to download data from `census.gov`, using tables from 2023's five-year American Community Survey. I want tract level data for my study area. I also want to compare to state-wide and country-wide data, so I'll grab those at a less detailed level.

I will start with looking at some basic demographic characteristics (age, sex, race) to see how they compare to each other, and to the state and country as a whole. 

The function to download the data is `get_acs()`, which takes a series of pretty self-explanatory arguments. For the state and country statistics, I will use the PEP summary tables.

## Age and Sex

```{r}
get_tables_tract <- function(table, labels) {
  get_acs(
  geography = "tract",
  state = "NM",
  county = "Bernalillo",
  table = table,
  year = 2023,
  geometry = T,
  cache_table = T
  ) %>%
    clean_tables(labels)
}

get_tables_state <- function(table, labels) {
  get_acs(
    geography = "state",
    state = "NM",
    table = table,
    year = 2023,
    geometry = T,
    cache_table = T
  ) %>% 
    clean_tables(labels)
}

get_tables_us <- function(table, labels) {
  get_acs(
    geography = "us",
    table = table,
    year = 2023,
    geometry = T,
    cache_table = T
  ) %>% 
    clean_tables(labels)
}

v23 <- load_variables(2023, "acs5")

get_labels <- function(table) {
  v23 %>% 
  filter(str_detect(name, table)) %>% 
  select(1:2) %>% 
  mutate(
    label = str_replace(label, "Estimate!!Total:!!", ""),
    label = str_replace(label, "Estimate!!", ""),
    label = str_replace(label, ":!!", "_"),
    label = str_replace(label, ":$", "")
  ) %>% 
  rename(variable = name)
}

clean_tables <- function(df, labels) {
  df %>% 
    select(-2) %>% 
    clean_names() %>% 
    left_join(labels) %>% 
    st_transform(6528) %>% 
    mutate(area = st_area(.))
}
```

```{r}
#| message: false
age_sex_labels <- get_labels("B01001A")

age_sex_nm <- get_tables_state("B01001A", age_sex_labels) %>% 
  st_drop_geometry() %>% 
  filter(str_detect(label, "years")) %>% 
  separate(label, c("sex", "agegroup"), sep = "_") %>% 
  select(sex, agegroup, value = estimate) %>% 
  as_tibble()

age_sex_us <- get_tables_us("B01001A", age_sex_labels) %>% 
    st_drop_geometry() %>% 
  filter(str_detect(label, "years")) %>% 
  separate(label, c("sex", "agegroup"), sep = "_") %>% 
  select(sex, agegroup, value = estimate) %>% 
  as_tibble()
```


:::{.callout-tip title="Explanation" collapse="true"}

:::

For Bernalillo County, this more difficult, since the summary tables aren't available at the tract level. I need tract level data in order to divide the area by the council districts. I will use the ACS 2023 tables. One problem to solve is that variables are returned with variable names which are alphanumeric codes and aren't descriptive. I will use the `get_variables()` function, which provides labels for the variables. The descriptions are still pretty ugly, so I will clean them up. I'll need to do this a few times, so I'll write a helper function.

Now I can download the data and join the labels. I'll filter out the summary rows, and separate the sex and agegroup into different variables. I also need to calculate the area of each tract, so I can apportion the data properly among council districts. I can do that with `st_area`.




```{r}
age_sex_bern <- get_tables_tract("B01001A", age_sex_labels) %>% 
  filter(str_detect(label, "years")) %>% 
  separate(label, c("sex", "agegroup"), sep = "_") %>% 
  select(geoid, sex, agegroup, value = estimate, area)
```

Now I need to split the data by district. I'll make a helper function for this, since I'll do this multiple times. Since all of the variables are counts (extensive), I will split the based on the areal proportion of each district within a given tract.

```{r}  
split_dists <- function(df, vars) {
  st_intersection(df, council_dists) %>% 
  mutate(
    area_split = st_area(.),
    area_pct = area_split / area,
    value = as.numeric(round(value * area_pct, 0))
  ) %>% 
  st_drop_geometry() %>% 
  select(all_of(vars)) %>% 
  arrange(district) %>% 
  as_tibble()
}
```

```{r}
age_sex_dist <- age_sex_bern %>% 
  split_dists(c("district", "sex", "agegroup", "value"))
age_sex_dist
```

```{r}
age_sex_plot <- function(df) {
  df %>% 
  group_by(sex, agegroup) %>% 
  summarise(value = sum(value)) %>% 
  mutate(value = ifelse(sex == "Male", -value, value)) %>% 
  ggplot(aes(x = value, y = agegroup, fill = sex)) +
  geom_col()
}
```

```{r}
age_levels <- age_sex_dist %>% 
  select(agegroup) %>% 
  unique() %>% 
  pull()

age_sex_dist <- age_sex_dist %>% 
      mutate(agegroup = factor(agegroup, age_levels))
```

```{r}
#| message: false
p1 <- age_sex_us %>% 
      mutate(agegroup = factor(agegroup, age_levels)) %>% 
      age_sex_plot()
p2 <- age_sex_nm %>% 
      mutate(agegroup = factor(agegroup, age_levels)) %>% 
      age_sex_plot()
p1 + p2
```

```{r}
#| message: false
age_sex_dist %>% 
  filter(district != "Los Ranchos") %>% 
  group_by(district,sex, agegroup) %>% 
  summarise(value = sum(value)) %>% 
  mutate(value = ifelse(sex == "Male", -value, value)) %>% 
  ggplot(aes(x = value, y = agegroup, fill = sex)) +
  geom_col() +
  facet_wrap(~district)
```

The small relative number of children from 15-19 years old, compared to both the country and the state as a whole, is striking. I want to confirm that nothing significant was lost by looking at the data before the split.

Apart from minor differences explained by tract areas lying outside district lines, this seems to confirm the data. This deserves further inquiry.

## Race

```{r}
#| collapse: false

race_labels <- get_labels("B03002")

race_us <- get_tables_us("B03002", race_labels) %>% 
  st_drop_geometry() %>% 
  filter(str_detect(label, "_")) %>% 
  separate(label, c("hisp", "race"), sep = "_") %>% 
  select(hisp, race, value = estimate) %>% 
  mutate(hisp = if_else(hisp == "Not Hispanic or Latino", "Non-Hispanic", "Hispanic"),
         race = str_replace(race, " alone$", ""),
         race = if_else(str_detect(race, "Two"), "Two or more races", race),
         race = if_else(str_detect(race, "Black"), "Black", race),
         ) %>% 
  group_by(hisp, race) %>% 
  summarise(value = sum(value))

race_nm <- get_tables_state("B03002", race_labels) %>% 
  st_drop_geometry() %>% 
  filter(str_detect(label, "_")) %>% 
  separate(label, c("hisp", "race"), sep = "_") %>% 
  select(hisp, race, value = estimate) %>% 
  mutate(hisp = if_else(hisp == "Not Hispanic or Latino", "Non-Hispanic", "Hispanic"),
         race = str_replace(race, " alone$", ""),
         race = if_else(str_detect(race, "Two"), "Two or more races", race),
         race = if_else(str_detect(race, "Black"), "Black", race),
         ) %>% 
  group_by(hisp, race) %>% 
  summarise(value = sum(value))
```

```{r}
race_bern <- get_tables_tract("B03002", race_labels) %>% 
  filter(str_detect(label, "_")) %>% 
  separate(label, c("hisp", "race"), sep = "_") %>% 
  select(geoid, hisp, race, value = estimate) %>% 
  st_transform(6528) %>% 
  mutate(area = st_area(.)) %>% 
  mutate(hisp = if_else(hisp == "Not Hispanic or Latino", "Non-Hispanic", "Hispanic"),
         race = str_replace(race, " alone$", ""),
         race = if_else(str_detect(race, "Two"), "Two or more races", race),
         race = if_else(str_detect(race, "Black"), "Black", race),
         ) %>% 
  group_by(hisp, race) %>% 
  summarise(value = sum(value),
            area = first(area))
race_bern
```


```{r}
race_dist <- race_bern %>% 
  split_dists(c("district", "hisp", "race", "value"))
race_dist
```

```{r}
race_plot <- function(df) {
  df %>% 
  group_by(race) %>% 
  summarise(value = sum(value)) %>% 
  mutate(value = ifelse(hisp == "Hispanic", -value, value)) %>% 
  ggplot(aes(x = value, y = hisp, fill = race)) +
  geom_col()
}

race_us$region <- "US"
race_nm$region <- "New Mexico"
race_bern_plt <- race_bern %>% st_drop_geometry() %>% select(-area)
race_bern_plt$region <- "Bernalillo"
race_compare <- rbind(race_bern_plt, race_us, race_nm)

race_compare
```

```{r}
race_compare %>% 
  filter(race != "Some other race" & 
           race != "Native Hawaiian and Other Pacific Islander") %>% 
  ggplot(aes(x = factor(region, level = c("US", "New Mexico", "Bernalillo")), 
           y = value, 
           fill = race)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d() +
  labs(x = "Region",
       y = "Race",
       title = "Racial Comparison",
       subtitle = "Bernalillo County, State and Country",
       caption = "Source: census.gov, 2023",
       fill = "Selected Races")
```

```{r}
race_dist %>% 
  filter(race != "Some other race" & 
           race != "Native Hawaiian and Other Pacific Islander") %>% 
  ggplot(aes(x = district, 
             y = value, 
             fill = race)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Region",
       y = "Race",
       title = "Racial Comparison",
       subtitle = "Bernalillo County, State and Country",
       caption = "Source: census.gov, 2023",
       fill = "Selected Races")
```

The proportions seem so similar in this plot. Let's look at actual values.

```{r}
race_dist %>% 
  filter(race != "Some other race" & 
           race != "Native Hawaiian and Other Pacific Islander") %>% 
  ggplot(aes(x = district, 
             y = value, 
             fill = race)) +
  geom_col() +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Region",
       y = "Race",
       title = "Racial Comparison",
       subtitle = "Bernalillo County, State and Country",
       caption = "Source: census.gov, 2023",
       fill = "Selected Races")
```

```{r}
ggplot(race_compare, 
       aes(x = factor(region, level = c("US", "New Mexico", "Bernalillo")), 
           y = value, fill = hisp)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d() +
  labs(x = "Region",
       y = "Race",
       title = "Hispanic Comparison",
       subtitle = "Bernalillo County, State and Country",
       caption = "Source: census.gov")
```

```{r}
ggplot(race_dist %>% 
  filter(race != "Some other race" & race != "Native Hawaiian and Other Pacific Islander"), 
       aes(x = district, 
           y = value, fill = hisp)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Region",
       y = "Race",
       title = "Hispanic Comparison",
       subtitle = "Bernalillo County, State and Country",
       caption = "Source: census.gov")
```

## Education

```{r}
edu_labels <- get_labels("B15002")

edu_us <- get_tables_us("B15002", edu_labels) %>% 
  st_drop_geometry() %>% 
  filter(str_detect(label, "_")) %>% 
  separate(label, c("sex", "education"), sep = "_") %>% 
  select(sex, education, value = estimate)

edu_nm <- get_tables_state("B15002", edu_labels) %>% 
  st_drop_geometry() %>% 
  filter(str_detect(label, "_")) %>% 
  separate(label, c("sex", "education"), sep = "_") %>% 
  select(geoid, sex, education, value = estimate)

edu_bern <- get_tables_tract("B15002", edu_labels) %>% 
  filter(str_detect(label, "_")) %>% 
  separate(label, c("sex", "education"), sep = "_") %>% 
  select(geoid, sex, education, value = estimate) %>% 
  st_transform(6528) %>% 
  mutate(area = st_area(.))

edu_dist <- edu_bern %>% 
  split_dists(c("district", "sex", "education", "value"))
```


```{r}
edu_plot <- function(df) {
  df %>% 
  group_by(sex, education) %>% 
  summarise(value = sum(value)) %>% 
  mutate(value = ifelse(sex == "Male", -value, value)) %>% 
  ggplot(aes(x = value, y = education, fill = sex)) +
  geom_col()
}
```

```{r}
edu_levels <- edu_us %>% 
  select(education) %>% 
  unique() %>% 
  pull()

edu_levels <- c("No HS Diploma", "High school graduate", "Some college", edu_levels[-(1:11)])

simplify_education <- function(df) {
  df %>% 
    mutate(
    education = case_when(
      education == "No schooling completed" |
        education == "Nursery to 4th grade" |
        education == "5th and 6th grade" |
        education == "7th and 8th grade" |
        education == "9th grade" |
        education == "10th grade" |
        education == "11th grade" |
        education == "12th grade, no diploma"
        ~ "No HS Diploma",
      education == "Some college, 1 or more years, no degree" |
        education == "Some college, less than 1 year"
        ~ "Some college",
      education == "High school graduate (includes equivalency)"
        ~ "High school graduate",
      TRUE ~ education
      )
    ) %>%
    {
      if ("district" %in% names(.)) group_by(., district, sex, education) 
      else group_by(., sex, education)
    } %>% 
    summarise(value = sum(value)) %>% 
   
  mutate(education = factor(education, edu_levels))
}

edu_us <- edu_us %>% simplify_education()
edu_nm <- edu_nm %>% simplify_education()
edu_dist <- edu_dist %>% simplify_education()
```

```{r}
p1 <- edu_us %>% edu_plot() +
  theme(axis.title.x = element_blank(),
        legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x = element_blank()
        )
p2 <- edu_nm %>% edu_plot() +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank())
p1 + p2
```

# Economic variables


```{r}
vars <- c(
  med_house_value = "B25077_001",
  income_hh = "B19001_001",
  population = "B01003_001",
  gini = "B19083_001",
  med_age = "B01002_001",
  in_poverty = "B17001A_002"
)
```

```{r}
#| eval: false
bern_data <- get_acs(
  geography = "tract",
  state = "NM",
  county = "Bernalillo",
  variables = vars,
  output = "wide",
  year = 2023,
  geometry = T,
  cache_table = T
)

st_write(bern_data, "data/demographics.gpkg", layer = "county", append = F)

nm_data <- get_acs(
  geography = "county",
  state = "NM",
  variables = vars,
  output = "wide",
  year = 2023,
  geometry = T,
  cache_table = T
)

st_write(nm_data, "data/demographics.gpkg", layer = "state", append = F)

us_data <- get_acs(
  geography = "state",
  variables = vars,
  output = "wide",
  year = 2023,
  geometry = T,
  cache_table = T
)

st_write(us_data, "data/demographics.gpkg", layer = "country", append = F)
```

I would like to compare education levels as well. The data for this is broken down into 25 variables, far more than I need, so I will group them into four. I'm also going to drop the margin of error information. Notice that, in this case, I am requesting an entire table rather than specific variables, by using the `table` argument rather than the `variable` argument.

```{r}
#| eval: false
bern_edu <- get_acs(
  geography = "tract",
  state = "NM",
  county = "Bernalillo",
  table = "B15003",
  output = "wide",
  year = 2023,
  geometry = T,
  cache_table = T
) %>%
  select(-2) %>%
  select(!ends_with("M")) %>%
  mutate(
    no_hs_deg = rowSums(across("B15003_002E":"B15003_016E")),
    hs_some_college = rowSums(across("B15003_017E":"B15003_020E")),
    bach_assoc_deg = rowSums(across("B15003_021E":"B15003_022E")),
    grad_deg = rowSums(across("B15003_023E":"B15003_025E"))
  ) %>%
  select(-c("B15003_001E":"B15003_025E"))

st_write(bern_edu, "data/demographics.gpkg", layer = "education", append = F)
```

```{r}
bernalillo_edu <-
  st_read("data/demographics.gpkg", layer = "education")
bernalillo_demographics <-
  st_read("data/demographics.gpkg", layer = "county")
```

Now let's look at the data.

```{r}
summary(bernalillo_demographics)
```

There is clean-up work to do here. There are some `NA`s, some areas with 0 population, an unneeded NAME column, and the ugly variable names to deal with. I also need to make sure it is in the correct CRS. First I'll fix the names, using a small function to make the code cleaner, drop the uninteresting columns and transform the coordinates. Then I do a simple join with the educational table based on the GEOID, dropping the geometry from one, and then dropping the error values.

```{r}
fix_names <- function(df) {
  df %>%
    rename_with(~ sub("E$", "", .)) %>%
    rename_with(~ sub("M$", "_moe", .))
}

bernalillo_demographics <-
  bernalillo_demographics %>%
  select(-2) %>% 
  fix_names() %>%
  st_transform(6528)


bernalillo_demographics <- bernalillo_demographics %>%
  left_join(st_drop_geometry(bernalillo_edu), by = "GEOID") %>%
  select(!ends_with("moe"))

summary(bernalillo_demographics)
```

That looks better. Now let's look at the 0 population areas.

```{r}
bernalillo_demographics %>% filter(population == 0)
```

```{r}
#| eval: false
base_plot(bernalillo_demographics %>% filter(population == 0)) +
  geom_sf_text(aes(label = GEOID))
```


One of these is way out of town. the other is the state fairgrounds. I will go ahead and drop them. For the few other `NA`s, I'll replace them with averages.

```{r}
bernalillo_demographics <- bernalillo_demographics %>%
  filter(population != 0) %>% 
  mutate(
    med_house_value = ifelse(is.na(med_house_value),
      median(med_house_value, na.rm = T),
      med_house_value
    ),
    gini = ifelse(is.na(gini),
      mean(gini, na.rm = T),
      gini
    )
  )

summary(bernalillo_demographics)
```

Most of the data seems normal, now. I am curious about the tract(s) with no household income and very small populations. Are they the same?

```{r}
bernalillo_demographics %>% 
  filter(income_hh < 100 | population < 150)
```

```{r}
#| eval: false
base_plot(bernalillo_demographics %>% filter(income_hh < 100 | population < 150)) +
  geom_sf_text(aes(label = GEOID))
```

One is again far out of town, and the other is the Petroglyph National Monument. I'll drop these as well.

```{r}
bernalillo_demographics <- 
  bernalillo_demographics %>% 
  filter(income_hh > 100 & population > 150)

summary(bernalillo_demographics)
```

Before combining this data with the council district shapes, I want to calculate the area for each tract, which I'll need later. I'll then calculate a population density for each area.

```{r}
bernalillo_demographics <- bernalillo_demographics %>% 
  mutate(
    area = st_area(.),
    pop_dens = population / area
  )
```

I would like to to calculate percentage values for the extensive variables. I'll use `purrr`'s `modify` with a lambda function to perform the calculations across all the relevant columns, then bind the new values back to the original data.

```{r}
demo_pcts <- 
  modify(bernalillo_demographics[7:11], 
       \(x) x / bernalillo_demographics$population) %>% 
  rename_with(~ sub("$", "_pct", .), .cols = everything()) %>% 
  st_drop_geometry()

bernalillo_demographics <- 
  bernalillo_demographics %>% 
  cbind(demo_pcts)

summary(bernalillo_demographics)
```


Now I want to divide the census data by council districts. Since many census tracts are split between multiple districts, areal interpolation will be necessary to assign appropriate values to each district. A number of options exist, such as `sf`'s own `st_interpolate_aw` and `aw_interpolate` from the `areal` package. Neither of these are what I really want, however, because I would like to keep the tract level data. I will instead intersect the datasets and do a manual areal interpolation.

First I'll use `st_intersection()` to split the tracts by council district. Since I'm not interested in areas outside the districts, I'll then crop the data.

```{r}
cd_demographics <- 
  st_intersection(bernalillo_demographics, council_dists) %>% 
  st_crop(st_bbox(council_dists))
cd_demographics %>% 
  arrange(GEOID) %>% 
  select(GEOID, population, no_hs_deg)
```

As you can see, we now have separate slices for tracts which span more than one district, but the data has been copied into each new row. For intensive variables like percentages or averages, this is okay, but for extensive variables like population, they must be spread across the new areas. I'll do this based on areal proportioning. First I'll calculate the areas of each new subdivision, and the percentage they represent of the former regions. This allows me to re-calculate the extensive variables, assigning the proper proportion to each.

```{r}
cd_demographics$area_split <- 
  st_area(cd_demographics) %>% 
  set_units(km^2)

cd_demographics <- cd_demographics %>% 
  group_by(GEOID) %>% 
  mutate(
    area_pct = area_split / area,
    population = round(population * area_pct, 0),
    in_poverty = round(in_poverty * area_pct, 0),
    no_hs_deg = round(no_hs_deg * area_pct, 0),
    hs_some_college = round(hs_some_college * area_pct, 0),
    bach_assoc_deg = round(bach_assoc_deg * area_pct, 0),
    grad_deg = round(grad_deg * area_pct, 0)
  ) %>% 
  arrange(GEOID)

st_write(cd_demographics, "data/cd-demographics.gpkg", append = F)

options(scipen = 100)

cd_demographics %>% 
  arrange(GEOID) %>% 
  select(GEOID, population, no_hs_deg)
```

## Visualization

Let's take a look, finally, at what we have. First, a few maps. I'll create a function to generate maps for each variable.

```{r}
base_plot_2 <- function(df, var) {
  var = df[[var]]
  ggplot(df) +
  annotation_map_tile(
    type = "osm",
    cachedir = "~/.cache/maps/",
    zoomin = -1,
    alpha = 0.6
  ) +
  geom_sf(aes(fill = as.numeric(var)), alpha = 0.4) +
  geom_sf(data = council_dists, aes(color = district), 
          fill = NA, linewidth = 0.5) +
  scale_fill_viridis_c() +
  scale_color_viridis_d() +
  guides(color = "none") +
  xlim(455000, 480000) +
  ylim(445000, 465000)
}
```


```{r}
poverty <- base_plot_2(cd_demographics, "in_poverty") +
  labs(
  title = "Poverty",
  subtitle = "By census tract",
  caption = "Source: census.gov ACS 2023",
  fill = "People\n in Poverty"
  )

household_value <- base_plot_2(cd_demographics, "med_house_value") +
  labs(
    title = "Median Household Value",
    subtitle = "By census tract",
    caption = "Source: census.gov ACS 2023",
    fill = "Med Household Value"
  )

gini <- base_plot_2(cd_demographics, "gini") +
  labs(
    title = "GINI",
    subtitle = "By census tract",
    caption = "Source: census.gov ACS 2023",
    fill = "GINI"
  )

income <- base_plot_2(cd_demographics, "income_hh") +
  labs(
    title = "Household Income",
    subtitle = "By census tract",
    caption = "Source: census.gov ACS 2023",
    fill = "Household Income"
  )

```

```{r}
(household_value + income) /
  (gini + poverty)
  
```



```{r}
cd_demographics %>% 
  st_drop_geometry() %>%
  group_by(district) %>% 
  summarise(
    population = sum(population), 
    med_house_value = median(med_house_value),
    income_hh = median(income_hh),
    med_age = median(med_age),
    in_poverty = sum(in_poverty),
    gini = mean(gini)
  ) %>% 
  select(district, population, med_house_value, income_hh, med_age, in_poverty, gini) %>%
  gt(rowname_col = "district") %>%
  cols_label(
    population = "Population",
    med_house_value = "Med House Value",
    income_hh = "HH Income",
    med_age = "Age",
    gini = "GINI"
  ) %>%
  tab_header(
    title = md("**Basic Demographics for Albuquerque**")
  ) %>%
  tab_source_note(
    source_note = md("*Source: census.gov, acs5, 2023*")
  ) %>%
  opt_stylize(style = 2) %>%
  fmt_currency(
    columns = c(med_house_value, income_hh),
    decimals = 0
  ) %>%
  fmt_number(columns = "med_age", decimals = 1) %>%
  fmt_number(columns = "gini", decimals = 3) %>%
  fmt_number(columns = "population", decimals = 0)
```

## Compare to State and Country
```{r}

```


