---
title: "Demographics"
format: 
  html:
    code-fold: false
    code-tools: true
    toc: true
    toc-depth: 2
exec:
  freeze: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

I would like to compare the demographics and some key economic data between the city council districts of Albuquerque, as well as comparing Albuquerque to the rest of the state and nation. This will involve obtaining data from `census.gov` as well as local government websites. Since census tracts do not line up with council districts, the census data will need to be interpolated into the council districts.

Throughout, I will emphasize good coding techniques, especially the creation of helper functions to avoid the significant amount of repetition involved in obtaining multiple data sets for three geographical levels.

I will use the following libraries:

```{r}
#| message: false
library(tidycensus)
library(tidyverse)
library(ggspatial)
library(sf)
library(units)
library(crsuggest)
library(gt)
library(ggtext)
library(patchwork)
library(janitor)
library(nngeo)
options(tigris_use_cache = TRUE)
```

# City Council Districts

I will start by obtaining the boundaries for the city council districts. The Albuquerque website provides a wealth of data, although it is not necessarily well-structured. In many of the data sets, the relevant data is contained in tables embedded in columns of `html` tables. I'll get to that when I incorporate crime statistics later. In this case, it is useless, as the website only provides councilors' names as a variable, not district numbers, and the names provided are for councilors from several years ago!

Fortunately, the Bernalillo County website provides what's needed, as well as providing the shape for The Village of Los Ranchos de Albuquerque. Los Ranchos is sort of the Beverly Hills of Albuquerque, in the sense that it is a separate municipal enclave (almost) entirely surrounded by Albuquerque, and where many the largest properties and mansions in the area are located.

The data portal for Bernalillo County is not very friendly for programmatic downloading, though. My first effort to download and unzip the file containing the city council boundaries failed to produce a usable file. Adding the argument `method = "curl"` allowed me to download the files, however.

```{r}
#| eval: false
download.file("https://www.berncoclerk.gov/wp-content/uploads/2025/07/BERNCO_CLERKJuly2025.gdb_.zip", "data/bernalillo.zip", method = "curl")
unzip("data/bernalillo.zip", overwrite = T)
```

I will read the file with `sf::st_read`, which will result in an `sf` object. The only variable I'm interested in is the district number, so I'll just grab that, fix the variable name, and make the content more descriptive for easy mapping. The `geometry` is sticky, so I don't need to specify it.

```{r}
council_dists_bc <- st_read("data/BC_CityCouncil/ABQ_CityCouncils.shp") %>%
  select(district = DISTRICTNU) %>%
  mutate(district = paste("District", district))
council_dists_bc
```

I'll do the same for Los Ranchos.

```{r}
#| eval: false
download.file("https://www.berncoclerk.gov/wp-content/uploads/2023/02/LosRanchos.zip", "data/test.zip", method = "curl")
unzip("data/LosRanchos.zip")
```

```{r}
los_ranchos <- st_read("data/LosRanchos/LosRanchos.shp") %>%
  select(district = Name)
```


:::{.callout-tip title="Finding a Coordinte Reference System" collapse="true"}

The `crsuggest` package has a useful function to determine an appropriate coordinate reference system. 

```{r}
suggest_crs(council_dists_bc) %>% head(2)
```

:::

Now I will bind them and transform the coordinate reference system to a local one in meters.

```{r}
council_dists <- rbind(council_dists_bc, los_ranchos) %>%
  st_transform(6528)
council_dists
```

I'll plot the district boundaries with `ggplot`, adding a base map with `ggspatial::annotation_map_tile`. I'll be doing the same map several times, so I'll make a function to do so.

```{r}
dist_plot <- function(df) {
  ggplot(df) +
  annotation_map_tile(
    type = "osm",
    cachedir = "~/.cache/maps/",
    zoomin = -1,
    alpha = 0.7
  ) +
  geom_sf() +
  theme_void() +
  geom_sf(aes(color = district), fill = NA, linewidth = 1) +
  geom_sf_text(aes(label = district)) +
  scale_color_viridis_d() +
  theme_void() +
  labs(title = "Albuquerque Council Districts")
}
```

```{r}
dist_plot(council_dists)
```

One thing to note from this map: there are numerous areas of unincorporated Bernalillo County which lie inside Albuquerque's borders but do not pertain to any city district. A closer look reveals some of these holes.

```{r}
dist_plot(council_dists %>% filter(district %in% c("District 2", "District 4", "District 5", "Los Ranchos")))
```

This will become relevant later as we map the census data onto the districts, as some of the data will lie outside city boundaries. On the other hand, I think it makes sense to fill in the holes in the middle of districts, to minimize the loss of data. I'll use the convenient `st_remove_holes()` function from the `nngeo` package.

```{r}
council_dists <- st_remove_holes(council_dists)
```

We can see that the holes are gone.

```{r}
dist_plot(council_dists)
```

```{r}
dist_plot(council_dists %>% filter(district %in% c("District 2", "District 4", "District 5", "Los Ranchos")))
```


#  Basic Demographics

I want to look at some basic demographic information, specifically age, sex, race and education, to see how they compare between council districts. I will use `tidycensus` to download data from `census.gov`, using tables from 2023's five-year American Community Survey. I want tract level data for my study area. I also want to compare to state-wide and country-wide data, so I'll grab those at a less detailed level.

The function to download the data is `get_acs()`, which takes a series of pretty self-explanatory arguments. Once downloaded, quite a bit of processing will be needed to prepare the data for analysis and visualization. I'll start with three functions to download the tables.


```{r}
get_tables_tract <- function(table, labels) {
  get_acs(
  geography = "tract",
  state = "NM",
  county = "Bernalillo",
  table = table,
  year = 2023,
  geometry = T,
  cache_table = T
  ) %>%
    clean_tables(labels)
}

get_tables_state <- function(table, labels) {
  get_acs(
    geography = "state",
    state = "NM",
    table = table,
    year = 2023,
    geometry = T,
    cache_table = T
  ) %>% 
    clean_tables(labels)
}

get_tables_us <- function(table, labels) {
  get_acs(
    geography = "us",
    table = table,
    year = 2023,
    geometry = T,
    cache_table = T
  ) %>% 
    clean_tables(labels)
}
```

A lot of the heavy lifting is down the the `clean_tables` function. The problem is that the tables contain variable names, which are un-descriptive alpha-numeric strings. For tables and plotting, they are useless. On the other hand, manually changing the names would be terrible. Instead, I'll use `tidycensus::load_variables`, which provides labels for each variable. The labels themselves need cleaning, so I'll start with a helper for just that.

```{r}
v23 <- load_variables(2023, "acs5")

get_labels <- function(table) {
  v23 %>% 
  filter(str_detect(name, table)) %>% 
  select(1:2) %>% 
  mutate(
    label = str_replace(label, "Estimate!!Total:!!", ""),
    label = str_replace(label, "Estimate!!", ""),
    label = str_replace(label, ":!!", "_"),
    label = str_replace(label, ":$", "")
  ) %>% 
  rename(variable = name)
}
```

Now I can write the clean tables function, which joins the labels to the data. I will also provide a local CRS, and calculate the area for each region. This will be needed when splitting the tract data between council districts.

```{r}
clean_tables <- function(df, labels) {
  df %>% 
    select(-2) %>% 
    clean_names() %>% 
    left_join(labels) %>% 
    st_transform(6528) %>% 
    mutate(area = st_area(.))
}
```

## Age and Sex

Now we can start looking at age and sex. For New Mexico and the US, I don't need the geometry, so I'll drop it. I also want to get rid of the rows which are totals and subtotals. The label is currently a combination of sex and age group, so I will split these into seperate columns. Finally, I'll just select the variables I want.

```{r}
#| message: false
age_sex_labels <- get_labels("B01001A")

age_sex_nm <- get_tables_state("B01001A", age_sex_labels) %>% 
  st_drop_geometry() %>% 
  filter(str_detect(label, "years")) %>% 
  separate(label, c("sex", "agegroup"), sep = "_") %>% 
  select(sex, agegroup, value = estimate) %>% 
  as_tibble()

age_sex_us <- get_tables_us("B01001A", age_sex_labels) %>% 
    st_drop_geometry() %>% 
  filter(str_detect(label, "years")) %>% 
  separate(label, c("sex", "agegroup"), sep = "_") %>% 
  select(sex, agegroup, value = estimate) %>% 
  as_tibble()
```


:::{.callout-tip title="Explanation" collapse="true"}

:::

For Bernalillo County, this more difficult, since I want to be able to split the data by district. To do so, I will first use `st_intersection`, which will create extra rows with new geographies for each sub-section of tracts which are divided among districts. The values are not automatically split among the subsections. Since the are all counts (extensive variables), I will calculate the areal proportion of each sub-section, and use the percentages to apportion the values among the sub-sections.


```{r}  
split_dists <- function(df, vars) {
  st_intersection(df, council_dists) %>% 
  mutate(
    area_split = st_area(.),
    area_pct = area_split / area,
    new_value = as.numeric(round(value * area_pct, 0))
  ) %>% 
  st_drop_geometry() %>% 
  select(all_of(vars)) %>% 
  arrange(district) %>% 
  as_tibble()
}
```

Now I can download the tract level data, and split the data.

```{r}
age_sex_bern <- get_tables_tract("B01001A", age_sex_labels) %>% 
  filter(str_detect(label, "years")) %>% 
  separate(label, c("sex", "agegroup"), sep = "_") %>% 
  select(geoid, sex, agegroup, value = estimate, area)

age_sex_dist <- age_sex_bern %>% 
  split_dists(c("geoid", "district", "sex", "agegroup", "value", "new_value", "area_pct"))
```

For now I'm keeping some variables I don't really need. This will help me evaluate the results of the splitting operation. I am expecting a certain amount of data loss, since parts of some of the tracts lie outside the council districts. I would like to do a visual sanity check to verify that the loss makes sense. I'll look at a small portion to see you much variance there is.

```{r}
age_sex_bern %>% 
  st_drop_geometry() %>% 
  group_by(sex, agegroup) %>% 
  summarise(value = sum(value)) %>% 
  filter(agegroup == "35 to 44 years")

age_sex_dist %>% 
  group_by(sex, agegroup) %>% 
  summarise(value = sum(new_value)) %>% 
  filter(agegroup == "35 to 44 years")
```

There is quite a difference. Let's see how this can occur by focusing in on one tract.

```{r}  
age_sex_dist %>% 
  filter(geoid == 35001003501 & agegroup == "35 to 44 years")
```

Here we are losing the majority of the population. Let's look at the tract.

```{r}
ggplot(age_sex_bern) +
  annotation_map_tile(
    type = "osm",
    cachedir = "~/.cache/maps/",
    zoomin = -1,
    alpha = 0.6
  ) +
  geom_sf(color = "black", fill = NA, linewidth = 0.5) +
  geom_sf(data = council_dists, aes(color = district), 
          fill = NA, linewidth = 2) +
  geom_sf(data = age_sex_bern %>% filter(geoid == 35001003501), 
          color = "red", fill = NA, linewidth = 2) +
  scale_fill_viridis_c() +
  scale_color_viridis_d() +
  theme_void() +
  guides(color = "none") +
  xlim(460000, 470000) +
  ylim(455000, 465000)
```

We can see that, in fact, only small portions of the tract are in each of the 2 council districts and Los Ranchos, and that most of the area is outside the city. I'm going to redo the split, dropping the unneeded variables.

```{r}
age_sex_dist <- age_sex_bern %>% 
  split_dists(c("district", "sex", "agegroup", value = "new_value"))
```

Now lets plot the results.

```{r}
mf_plot <- function(df, groups, pos) {
  df %>% 
  group_by_at(groups) %>% 
  summarise(value = sum(value)) %>% 
  mutate(value = ifelse(sex == "Male", -value, value)) %>% 
  ggplot(aes(x = value, y = .data[[groups[pos]]], fill = sex)) +
  geom_col()
}
```

```{r}
age_levels <- age_sex_dist %>% 
  select(agegroup) %>% 
  unique() %>% 
  pull()

age_sex_dist <- age_sex_dist %>% 
      mutate(agegroup = factor(agegroup, age_levels))
```

```{r}
#| message: false
p1 <- age_sex_us %>%
  mutate(agegroup = factor(agegroup, age_levels)) %>% 
  mf_plot(c("sex", "agegroup"), 2) +
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x = element_blank()
        ) +
  xlab("US")
p2 <- age_sex_nm %>%
  mutate(agegroup = factor(agegroup, age_levels)) %>% 
  mf_plot(c("sex", "agegroup"), 2) +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.text.x = element_blank()) +
  xlab("New Mexico")
p1 + p2
```

```{r}
#| message: false
age_sex_dist %>% 
  mutate(agegroup = factor(agegroup, age_levels)) %>% 
  filter(district != "Los Ranchos") %>% 
  mf_plot(c("district", "sex", "agegroup"), 3) +
  facet_wrap(~district) +
  xlab("population") +
  theme(axis.text.y = element_text(size = 6))
  
```

## Race

```{r}
#| collapse: false

race_labels <- get_labels("B03002")

race_clean <- function(df, area) {
  df %>% 
    filter(str_detect(label, "_")) %>% 
    separate(label, c("hisp", "race"), sep = "_") %>%
    {
      if (area) select(., geoid, hisp, race, value = estimate, area)
      else select(., hisp, race, value = estimate)
    } %>%
    mutate(hisp = if_else(hisp == "Not Hispanic or Latino", "Non-Hispanic", "Hispanic"),
           race = str_replace(race, " alone$", ""),
           race = if_else(str_detect(race, "Two"), "Two or more races", race),
           race = if_else(str_detect(race, "Black"), "Black", race),
           ) %>%
  group_by(hisp, race)
}
```

```{r}
race_us <- get_tables_us("B03002", race_labels) %>% 
  st_drop_geometry() %>% 
  race_clean(area = FALSE) %>% 
  mutate(region = "US")

race_nm <- get_tables_state("B03002", race_labels) %>% 
  st_drop_geometry() %>% 
  race_clean(area = FALSE) %>% 
  mutate(region = "New Mexico")
```

```{r}
race_bern <- get_tables_tract("B03002", race_labels) %>% 
  race_clean(area = T)
race_dist <- race_bern %>% 
  split_dists(c("district", "hisp", "race", "value"))
```

```{r}
library(zeallot)
race_bern_plt <- race_bern %>% st_drop_geometry() %>% select(-area)
race_bern_plt$region <- "Bernalillo"
race_compare <- rbind(race_bern_plt, race_us, race_nm)
c(race_compare_plt, race_dist_plt) %<-% 
  map(list(race_compare, race_dist), 
      \(x) (x %>% 
              filter(race != "Some other race" & 
                     race != "Native Hawaiian and Other Pacific Islander")))
```

```{r}
race_plot <- function(df, fill, dist = F) {
  ggplot(df, aes(x = factor(region, level = c("US", "New Mexico", "Bernalillo")), 
           y = value, 
           fill = fill)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d() +
  labs(x = "Region",
       y = "Race",
       title = "Racial Comparison",
       subtitle = "Bernalillo County, State and Country",
       caption = "Source: census.gov, 2023",
       fill = "Selected Races")
}

race_plot_labels <- function (plt, race = T) {
    labs(x = "Region",
       y = "Race",
       title = "Racial Comparison",
       subtitle = "Bernalillo County, State and Country",
       caption = "Source: census.gov, 2023",
       fill = "Selected Races")
}
```


```{r}
race_compare_plt %>% 
  ggplot(aes(x = factor(region, level = c("US", "New Mexico", "Bernalillo")), 
           y = value, 
           fill = race)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d() +
  labs(x = "Region",
       y = "Race",
       title = "Racial Comparison",
       subtitle = "Bernalillo County, State and Country",
       caption = "Source: census.gov, 2023",
       fill = "Selected Races") +
  labs()
```


```{r}
race_dist_plt %>% 
  ggplot(aes(x = district, 
             y = value, 
             fill = race)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Region",
       y = "Race",
       title = "Racial Comparison",
       subtitle = "Albuquerque Districts",
       caption = "Source: census.gov, 2023",
       fill = "Selected Races")
```

The proportions seem so similar in this plot. Let's look at actual values.

```{r}
race_dist_plt %>% 
  ggplot(aes(x = district, 
             y = value, 
             fill = race)) +
  geom_col() +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Region",
       y = "Race",
       title = "Racial Comparison",
       subtitle = "Bernalillo County, State and Country",
       caption = "Source: census.gov, 2023",
       fill = "Selected Races")
```

```{r}
ggplot(race_compare_plt, 
       aes(x = factor(region, level = c("US", "New Mexico", "Bernalillo")), 
           y = value, fill = hisp)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d() +
  labs(x = "Region",
       y = "Race",
       title = "Hispanic Comparison",
       subtitle = "Bernalillo County, State and Country",
       caption = "Source: census.gov")
```

```{r}
ggplot(race_dist_plt, 
       aes(x = district, 
           y = value, fill = hisp)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Region",
       y = "Race",
       title = "Hispanic Comparison",
       subtitle = "Bernalillo County, State and Country",
       caption = "Source: census.gov")
```

## Education

```{r}
edu_labels <- get_labels("B15002")

edu_us <- get_tables_us("B15002", edu_labels) %>% 
  st_drop_geometry() %>% 
  filter(str_detect(label, "_")) %>% 
  separate(label, c("sex", "education"), sep = "_") %>% 
  select(sex, education, value = estimate)

edu_nm <- get_tables_state("B15002", edu_labels) %>% 
  st_drop_geometry() %>% 
  filter(str_detect(label, "_")) %>% 
  separate(label, c("sex", "education"), sep = "_") %>% 
  select(geoid, sex, education, value = estimate)

edu_bern <- get_tables_tract("B15002", edu_labels) %>% 
  filter(str_detect(label, "_")) %>% 
  separate(label, c("sex", "education"), sep = "_") %>% 
  select(geoid, sex, education, value = estimate) %>% 
  st_transform(6528) %>% 
  mutate(area = st_area(.))

edu_dist <- edu_bern %>% 
  split_dists(c("district", "sex", "education", "value"))
```


```{r}
edu_plot <- function(df) {
  df %>% 
  group_by(sex, education) %>% 
  summarise(value = sum(value)) %>% 
  mutate(value = ifelse(sex == "Male", -value, value)) %>% 
  ggplot(aes(x = value, y = education, fill = sex)) +
  geom_col()
}
```

```{r}
edu_levels <- edu_us %>% 
  select(education) %>% 
  unique() %>% 
  pull()

edu_levels <- c("No HS Diploma", "High school graduate", "Some college", edu_levels[-(1:11)])

simplify_education <- function(df) {
  df %>% 
    mutate(
    education = case_when(
      education == "No schooling completed" |
        education == "Nursery to 4th grade" |
        education == "5th and 6th grade" |
        education == "7th and 8th grade" |
        education == "9th grade" |
        education == "10th grade" |
        education == "11th grade" |
        education == "12th grade, no diploma"
        ~ "No HS Diploma",
      education == "Some college, 1 or more years, no degree" |
        education == "Some college, less than 1 year"
        ~ "Some college",
      education == "High school graduate (includes equivalency)"
        ~ "High school graduate",
      TRUE ~ education
      )
    ) %>%
    {
      if ("district" %in% names(.)) group_by(., district, sex, education) 
      else group_by(., sex, education)
    } %>% 
    summarise(value = sum(value)) %>% 
   
  mutate(education = factor(education, edu_levels))
}

edu_us <- edu_us %>% simplify_education()
edu_nm <- edu_nm %>% simplify_education()
edu_dist <- edu_dist %>% simplify_education()
```

```{r}
p1 <- edu_us %>% edu_plot() +
  theme(axis.title.x = element_blank(),
        legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x = element_blank()
        )
p2 <- edu_nm %>% edu_plot() +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank())
p1 + p2
```

```{r}
edu_dist %>% 
  filter(district != "Los Ranchos") %>% 
  group_by(district,sex, education) %>% 
  summarise(value = sum(value)) %>% 
  mutate(value = ifelse(sex == "Male", -value, value)) %>% 
  ggplot(aes(x = value, y = education, fill = sex)) +
  geom_col() +
  facet_wrap(~district)
```


# Economic variables


```{r}
vars <- c(
  med_house_value = "B25077_001",
  income_hh = "B19001_001",
  gini = "B19083_001",
  in_poverty = "B17001A_002",
  population = "B01003_001"
)
```

```{r}
#| eval: false
econ_bern <- get_acs(
  geography = "tract",
  state = "NM",
  county = "Bernalillo",
  variables = vars,
  output = "wide",
  year = 2023,
  geometry = T,
  cache_table = T
)

# st_write(bern_data, "data/demographics.gpkg", layer = "county", append = F)

econ_nm <- get_acs(
  geography = "county",
  state = "NM",
  variables = vars,
  output = "wide",
  year = 2023,
  geometry = T,
  cache_table = T
)

# st_write(nm_data, "data/demographics.gpkg", layer = "state", append = F)

econ_us <- get_acs(
  geography = "state",
  variables = vars,
  output = "wide",
  year = 2023,
  geometry = T,
  cache_table = T
)


# st_write(us_data, "data/demographics.gpkg", layer = "country", append = F)
```

I would like to compare education levels as well. The data for this is broken down into 25 variables, far more than I need, so I will group them into four. I'm also going to drop the margin of error information. Notice that, in this case, I am requesting an entire table rather than specific variables, by using the `table` argument rather than the `variable` argument.


```{r}
bernalillo_demographics <- econ_bern
```

Now let's look at the data.

```{r}
summary(bernalillo_demographics)
```

There is clean-up work to do here. There are some `NA`s, some areas with 0 population, an unneeded NAME column, and the ugly variable names to deal with. I also need to make sure it is in the correct CRS. First I'll fix the names, using a small function to make the code cleaner, drop the uninteresting columns and transform the coordinates. Then I do a simple join with the educational table based on the GEOID, dropping the geometry from one, and then dropping the error values.

```{r}
fix_names <- function(df) {
  df %>%
    rename_with(~ sub("E$", "", .)) %>%
    rename_with(~ sub("M$", "_moe", .))
}

bernalillo_demographics <-
  bernalillo_demographics %>%
  select(-2) %>% 
  fix_names() %>%
  st_transform(6528)


# bernalillo_demographics <- bernalillo_demographics %>%
#   left_join(st_drop_geometry(bernalillo_edu), by = "GEOID") %>%
#   select(!ends_with("moe"))

summary(bernalillo_demographics)
```

That looks better. Now let's look at the 0 population areas.

```{r}
bernalillo_demographics %>% filter(population == 0)
```

```{r}
#| eval: false
base_plot(bernalillo_demographics %>% filter(population == 0)) +
  geom_sf_text(aes(label = GEOID))
```


One of these is way out of town. the other is the state fairgrounds. I will go ahead and drop them. For the few other `NA`s, I'll replace them with averages.

```{r}
bernalillo_demographics <- bernalillo_demographics %>%
  filter(population != 0) %>% 
  mutate(
    med_house_value = ifelse(is.na(med_house_value),
      median(med_house_value, na.rm = T),
      med_house_value
    ),
    gini = ifelse(is.na(gini),
      mean(gini, na.rm = T),
      gini
    )
  )

summary(bernalillo_demographics)
```

Most of the data seems normal, now. I am curious about the tract(s) with no household income and very small populations. Are they the same?

```{r}
bernalillo_demographics %>% 
  filter(income_hh < 100 | population < 150)
```

```{r}
#| eval: false
base_plot(bernalillo_demographics %>% filter(income_hh < 100 | population < 150)) +
  geom_sf_text(aes(label = GEOID))
```

One is again far out of town, and the other is the Petroglyph National Monument. I'll drop these as well.

```{r}
bernalillo_demographics <- 
  bernalillo_demographics %>% 
  filter(income_hh > 100 & population > 150)

summary(bernalillo_demographics)
```

Before combining this data with the council district shapes, I want to calculate the area for each tract, which I'll need later. I'll then calculate a population density for each area.

```{r}
bernalillo_demographics <- bernalillo_demographics %>% 
  mutate(
    area = st_area(.),
    pop_dens = population / area
  )
```

I would like to to calculate percentage values for the extensive variables. I'll use `purrr`'s `modify` with a lambda function to perform the calculations across all the relevant columns, then bind the new values back to the original data.

```{r}
demo_pcts <- 
  modify(bernalillo_demographics[7:11], 
       \(x) x / bernalillo_demographics$population) %>% 
  rename_with(~ sub("$", "_pct", .), .cols = everything()) %>% 
  st_drop_geometry()

bernalillo_demographics <- 
  bernalillo_demographics %>% 
  cbind(demo_pcts)

summary(bernalillo_demographics)
```


Now I want to divide the census data by council districts. Since many census tracts are split between multiple districts, areal interpolation will be necessary to assign appropriate values to each district. A number of options exist, such as `sf`'s own `st_interpolate_aw` and `aw_interpolate` from the `areal` package. Neither of these are what I really want, however, because I would like to keep the tract level data. I will instead intersect the datasets and do a manual areal interpolation.

First I'll use `st_intersection()` to split the tracts by council district. Since I'm not interested in areas outside the districts, I'll then crop the data.

```{r}
cd_demographics <- 
  st_intersection(bernalillo_demographics, council_dists) %>% 
  st_crop(st_bbox(council_dists))
cd_demographics %>% 
  arrange(GEOID) %>% 
  select(GEOID, population, no_hs_deg)
```

As you can see, we now have separate slices for tracts which span more than one district, but the data has been copied into each new row. For intensive variables like percentages or averages, this is okay, but for extensive variables like population, they must be spread across the new areas. I'll do this based on areal proportioning. First I'll calculate the areas of each new subdivision, and the percentage they represent of the former regions. This allows me to re-calculate the extensive variables, assigning the proper proportion to each.

```{r}
cd_demographics$area_split <- 
  st_area(cd_demographics) %>% 
  set_units(km^2)

cd_demographics <- cd_demographics %>% 
  group_by(GEOID) %>% 
  mutate(
    area_pct = area_split / area,
    population = round(population * area_pct, 0),
    in_poverty = round(in_poverty * area_pct, 0),
    no_hs_deg = round(no_hs_deg * area_pct, 0),
    hs_some_college = round(hs_some_college * area_pct, 0),
    bach_assoc_deg = round(bach_assoc_deg * area_pct, 0),
    grad_deg = round(grad_deg * area_pct, 0)
  ) %>% 
  arrange(GEOID)

st_write(cd_demographics, "data/cd-demographics.gpkg", append = F)

options(scipen = 100)

cd_demographics %>% 
  arrange(GEOID) %>% 
  select(GEOID, population, no_hs_deg)
```

## Visualization

Let's take a look, finally, at what we have. First, a few maps. I'll create a function to generate maps for each variable.

```{r}
base_plot_2 <- function(df, var) {
  var = df[[var]]
  ggplot(df) +
  annotation_map_tile(
    type = "osm",
    cachedir = "~/.cache/maps/",
    zoomin = -1,
    alpha = 0.6
  ) +
  geom_sf(aes(fill = as.numeric(var)), alpha = 0.4) +
  geom_sf(data = council_dists, aes(color = district), 
          fill = NA, linewidth = 0.5) +
  scale_fill_viridis_c() +
  scale_color_viridis_d() +
  guides(color = "none") +
  xlim(455000, 480000) +
  ylim(445000, 465000)
}
```


```{r}
poverty <- base_plot_2(cd_demographics, "in_poverty") +
  labs(
  title = "Poverty",
  subtitle = "By census tract",
  caption = "Source: census.gov ACS 2023",
  fill = "People\n in Poverty"
  )

household_value <- base_plot_2(cd_demographics, "med_house_value") +
  labs(
    title = "Median Household Value",
    subtitle = "By census tract",
    caption = "Source: census.gov ACS 2023",
    fill = "Med Household Value"
  )

gini <- base_plot_2(cd_demographics, "gini") +
  labs(
    title = "GINI",
    subtitle = "By census tract",
    caption = "Source: census.gov ACS 2023",
    fill = "GINI"
  )

income <- base_plot_2(cd_demographics, "income_hh") +
  labs(
    title = "Household Income",
    subtitle = "By census tract",
    caption = "Source: census.gov ACS 2023",
    fill = "Household Income"
  )

```

```{r}
(household_value + income) /
  (gini + poverty)
  
```



```{r}
cd_demographics %>% 
  st_drop_geometry() %>%
  group_by(district) %>% 
  summarise(
    population = sum(population), 
    med_house_value = median(med_house_value),
    income_hh = median(income_hh),
    med_age = median(med_age),
    in_poverty = sum(in_poverty),
    gini = mean(gini)
  ) %>% 
  select(district, population, med_house_value, income_hh, med_age, in_poverty, gini) %>%
  gt(rowname_col = "district") %>%
  cols_label(
    population = "Population",
    med_house_value = "Med House Value",
    income_hh = "HH Income",
    med_age = "Age",
    gini = "GINI"
  ) %>%
  tab_header(
    title = md("**Basic Demographics for Albuquerque**")
  ) %>%
  tab_source_note(
    source_note = md("*Source: census.gov, acs5, 2023*")
  ) %>%
  opt_stylize(style = 2) %>%
  fmt_currency(
    columns = c(med_house_value, income_hh),
    decimals = 0
  ) %>%
  fmt_number(columns = "med_age", decimals = 1) %>%
  fmt_number(columns = "gini", decimals = 3) %>%
  fmt_number(columns = "population", decimals = 0)
```

## Compare to State and Country
```{r}

```


