---
title: "Demographics"
format: 
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 2
exec:
  freeze: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Setup


```{r}
#| message: false
library(tidycensus)
library(tidyverse)
library(ggspatial)
library(sf)
library(units)
library(crsuggest)
library(gt)
library(ggtext)
library(patchwork)
library(scales)
```


```{r}
source("demographic-functions.R")
```

The function to download the data is `get_acs()`, which takes a series of pretty self-explanatory arguments. `tidycensus` allows you to supply a named vector for variables, so that the returned data will already have useful variable names. By specifying `geometry = TRUE` I will get an `sf` object back, and specifying `output=wide` gives me a table with a separate column for each variable.


```{r}
year <- 2023
crs <- 6528
caption <- str_glue("Source: census.gov, acs5, {year}")
```

I can use `get_acs` for variables, not just tables, as I did last time. The only difference is the `variables` argument in place of the `tables` argument. I will write this function a little differently than the `get_tables` function used before. Instead of building conditional logic into the function, I will write a simplified function.


```{r}
get_vars <-
  function(geography, vars, geometry,
           state = NULL, county = NULL) {
    get_acs(
      geography = geography,
      state = state,
      county = county,
      variables = vars,
      output = "wide",
      year = year,
      geometry = geometry,
      cache_table = T
    ) %>%
      select(-2) %>%
      fix_names()
  }
```

Now I will use `purrr::partial()` to create separate functions for each geography. I'll also write a function to fix the names.

```{r}
get_bern_tracts <- partial(get_vars,
  geography = "tract",
  state = "NM",
  county = "Bernalillo",
  geometry = T
)

get_nm <- partial(get_vars,
  geography = "state",
  state = "NM",
  geometry = F
)

get_us <- partial(get_vars,
  geography = "us",
  geometry = F
)

fix_names <- function(df) {
  df %>%
    rename_with(~ sub("E$", "", .)) %>%
    rename_with(~ sub("M$", "_moe", .)) %>%
    rename_with(~ sub("GEOID", "geoid", .))
}
```



```{r}
vars <- c(
  `Med House Value` = "B25077_001",
  `Owner Occupied` = "B25003_002",
  `Renter Occupied` = "B25003_003",
  Vacant = "B25002_003",
  Occupied = "B25002_002",
  `Med HH Income` = "B19013_001",
  `Med HH Income Hisp` = "B19013I_001",
  Gini = "B19083_001",
  `Med Rent` = "B25031_001"
)
```

```{r}
bern_data <- get_bern_tracts(vars) %>% 
  st_transform(crs)
```

Now let's look at the data.

```{r}
summary(bern_data)
```

There is clean-up work to do here. There are some `NA`s, some areas with minimums of 0.

```{r}
bern_data %>% 
  filter(if_any(everything(), is.na)) %>% 
  print(n = 14)
```

Four of these rows clearly have bad data

The `NA`s have to do with housing. I will replace them with average values, but I will wait until I divide the data by district and use district averages instead of city-wide ones.

I'm curious about the very low minimums for some variables.

```{r}
bern_data %>% 
  filter(is.na(`Med HH Income`))
```

These records seem responsible for a lot of the confusion. I'd like to replace these with `NA`, but I don't want to do it now, since the records would be filled with average values later. For now, I'll mark them with a dummy value.

```{r}
bern_data <- bern_data %>% 
  mutate(across(where(is.numeric), 
                \(x) if_else(is.na(`Med HH Income`), -999, x)))
```

I'll split the data between districts. Unlike in the last article, these values are all averages (intensive). After dividing the data between districts, these values will not need to be apportioned among the sub-sections. 

```{r}
council_dists <- st_read("data/abq_demographics.gpkg", layer = "council_dists") %>% 
  st_transform(crs)
dist_data <- st_intersection(bern_data, council_dists)
summary(dist_data)
```

I will fill in the missing data with averages, then turn the dummy -999 value to `NA`.

```{r}
vars_fin <- c("Med.House.Value", "Med.HH.Income", "Med.HH.Income.Hisp", "Med.Rent", "Gini")
vars_fin_all <- c(vars_fin, str_replace(vars_fin, "$", "_moe"))

var_avgs <- dist_data %>% 
  st_drop_geometry() %>% 
  group_by(district) %>% 
  summarise(across(all_of(vars_fin_all), list(dist = \(x) mean(x, na.rm = T)))) %>% 
  ungroup()

fill_avgs <- function(var, replace) {
  if_else(is.na(var), replace, var)
}

dist_data <- dist_data %>%
  left_join(var_avgs) %>% 
  mutate(
    Med.House.Value = fill_avgs(Med.House.Value, Med.House.Value_dist),
    Med.House.Value_moe = 
      fill_avgs(Med.House.Value_moe,Med.House.Value_moe_dist),
    Med.Rent = fill_avgs(Med.Rent, Med.Rent_dist),
    Med.Rent_moe = fill_avgs(Med.Rent_moe, Med.Rent_moe_dist),
    Med.HH.Income = fill_avgs(Med.HH.Income, Med.HH.Income_dist),
    Med.HH.Income_moe = fill_avgs(Med.HH.Income_moe, Med.HH.Income_moe_dist),
    Med.HH.Income.Hisp = fill_avgs(Med.HH.Income.Hisp, Med.HH.Income.Hisp_dist),
    Med.HH.Income.Hisp_moe = fill_avgs(Med.HH.Income.Hisp_moe, Med.HH.Income.Hisp_moe_dist),
    Gini = fill_avgs(Gini, Gini_dist),
    Gini_moe = fill_avgs(Gini_moe, Gini_moe_dist)
  ) %>% 
  mutate(across(where(is.numeric), 
                \(x) if_else(Med.HH.Income == -999, NA_real_, x)))

summary(dist_data)
```


## Visualization

Let's take a look, finally, at what we have. First, a few maps. I'll create a function to generate maps for each variable. This function will build on the `map_demo` function from last time.

```{r}
map_demo_value <- partial(map_demo, pct = F)

map_var <- function(df, var, label) {
  df %>%
    mutate(value = .data[[var]],
           label = scales::label_comma()(value)) %>% 
    map_demo_value() +
    labs(title = label)
}

dist_sum <- dist_data %>% 
  group_by(district) %>% 
  summarise(
    Med.House.Value = round(mean(Med.House.Value, na.rm = T)),
    Med.HH.Income = round(mean(Med.HH.Income, na.rm = T)),
    Med.HH.Income.Hisp = round(mean(Med.HH.Income.Hisp, na.rm = T)),
    Med.Rent = round(mean(Med.Rent, na.rm = T)),
    Gini = mean(Gini, na.rm = T)
  )

labels <- c("Median House Value", "Median Household Income", "Median Household Income Hispanic", "Median gross rent",
            "Gini index")

map2(vars_fin, labels,
      \(var, label) dist_sum %>%
        map_var(var, label))
```

```{r}
ggplot(dist_data, aes(x = log(Med.House.Value))) +
  geom_histogram(alpha = 0.5, fill = "navy", color = "navy",
                 bins = 40) +
  facet_wrap(~ district) +
  theme_minimal() +
  scale_x_continuous(labels = scales::label_number(accuracy = 0.1)) +
  labs(x = "Median home value")
```



```{r}
abq_data <- dist_data %>% 
  filter(district != "Los Ranchos")
library(classInt)
options(scipen = 999)
br_quantile <- classIntervals(
  abq_data$med_hh_income, 5,
  style = "quantile")
br_pretty <- classIntervals(
  abq_data$med_hh_income, 5,
  style = "pretty")
br_fisher <- classIntervals(
  abq_data$med_hh_income, 5,
  style = "fisher")
br_km <- classIntervals(
  abq_data$med_hh_income, 5,
  style = "kmeans")
```

```{r}
abq_data$med_hh_income_q <- cut(abq_data$med_hh_income, br_quantile$brks,
                        include.lowest = TRUE, dig.lab = 10)
pal <- colorRampPalette(c("darkred", "darkgreen", "navy"))(length(br_quantile$brks))
```

```{r}
map_tracts <- function(sf, sf_sum, pct = F, desc = T,
                       title = "Population", subtitle = "") {
  dist_plot(council_dists) +
    {
      if (pct) {
        geom_sf(data = sf, aes(fill = percent), alpha = 0.35)
      } else {
        geom_sf(data = sf, aes(fill = value), alpha = 0.35)
      }
    } +
    {
      if (pct) {
        scale_fill_viridis_c(labels = label_percent())
      } else if (desc) {
        scale_fill_viridis_d()
      } else {
        scale_fill_viridis_c(labels = label_comma())
      }
    } +
    geom_sf_label(
      data = council_dists,
      aes(label = district), fontface = "bold",
      nudge_y = 1000, nudge_x = -500, fill = "gray",
      label.padding = unit(0.1, "lines"), size = 3.5
    ) +
    geom_sf_label(
      data = sf_sum,
      aes(label = label),
      fill = "grey", nudge_y = -1000
    ) + guides(color = "none") +
    labs(
      title = title,
      subtitle = subtitle,
      caption = source,
      fill = NULL
    )
}

make_plot_data <- function(df, vars) {
  plot_data <- df %>% 
  select(district, value = .data[[vars[1]]], .data[[vars[2]]])
  plot_data_sum <- df %>% 
  group_by(district) %>% 
  summarise(avg = mean(.data[[vars[2]]], na.rm = T)) %>% 
      mutate(label = label_comma()(avg))
  list("plot_data" = plot_data, 
       "plot_data_sum" = plot_data_sum)
}

plot_data <- abq_data %>% 
    make_plot_data(c("med_hh_income_q", "med_hh_income"))
plot_data_sum <- plot_data$plot_data_sum
plot_data <- plot_data$plot_data

plot_data %>%
  map_tracts(plot_data_sum,
    title = "Median Household Income",
    subtitle = "Albuquerque Districts"
  )


```


```{r}
ggplot(abq_data) +
  annotation_map_tile(
    type = "osm",
    cachedir = "~/.cache/maps/",
    zoomin = -1,
    alpha = 0.5
  ) +
  geom_sf(aes(fill = med_hh_income_q), alpha = 0.4) +
  geom_sf(data = council_dists, aes(color = district), fill = NA, linewidth = 1) +
  geom_sf_text(data = council_dists, aes(label = district)) +
  scale_fill_manual(values = pal) +
  scale_color_viridis_d(guide = "none") +
  theme_void() +
  labs(
    title = "2023 Median Household Income in Albuquerque, NM",
    subtitle = "By census tract",
    caption = "Source: census.gov ACS",
    fill = md("Median<br>Household<br>Income")
  ) +
  theme(
    legend.title = element_markdown(),
  )
```

## Initial Visualization and Summary Data

```{r}
us_data <- get_us(vars) %>% 
  mutate(district = "US")
nm_data <- get_nm(vars) %>% 
  mutate(district = "New Mexico")

dist_data_simple <- dist_data %>% 
  st_drop_geometry() %>% 
  select(!ends_with("dist"))
nm_data$income_hh

all_data <- rbind(dist_data_simple, us_data, nm_data)
```


```{r}
all_data %>% 
  st_drop_geometry() %>%
  group_by(district) %>% 
  summarise(
    med_house_value = mean(med_house_value),
    income_hh = mean(income_hh),
    med_gross_rent = mean(med_gross_rent),
    gini = mean(gini)
  ) %>% 
  select(district, med_house_value, med_gross_rent, income_hh, gini) %>%
  gt(rowname_col = "district") %>%
  cols_label(
    med_house_value = "Med House Value",
    med_gross_rent = "Med Gross Rent",
    income_hh = "HH Income",
    gini = "GINI"
  ) %>%
  tab_header(
    title = md("**Basic Financial Indicators**"),
    subtitle = "Albuquerque by District"
  ) %>%
  tab_source_note(
    source_note = md("*Source: census.gov, acs5, 2023*")
  ) %>%
  opt_stylize(style = 2) %>%
  fmt_currency(
    columns = c(med_house_value, income_hh),
    decimals = 0
  ) %>%
  fmt_number(columns = vars_fin, decimals = 0) %>%
  fmt_number(columns = "gini", decimals = 3)
```

## Compare to State and Country


```{r}
ggplot(abq_data, aes(x = district, y = income_hh)) +
  geom_boxplot(fill = "lightgrey") +
  geom_jitter(width = .2) +
  theme_bw() +
  labs(
    y = "Household Income",
    title = "2023 Household Income in Albuquerque, NM",
    subtitle = "By district",
    caption = "Source: census.gov ACS"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.title = element_blank()
  )
```


## Adding demographics

```{r}
process <- function(df, var) {
  df %>%
    st_drop_geometry() %>%
    group_by_at(c("geoid", "district", var)) %>%
    summarise(value = sum(value)) %>%
    pivot_wider(names_from = var, values_from = value) %>% 
    ungroup()
}

variables <- c("agegroup", "education", "hisp")
layers <- c("age_sex_dist_sf", "edu_dist_sf", "race_dist_sf")

abq_demos <-
  map2(
    layers,
    variables,
    \(layer, variable) st_read("data/abq_demographics.gpkg", layer = layer) %>%
      process(variable)
  ) %>%
  reduce(full_join, by = c("geoid", "district")) %>% 
  ungroup()

dist_data <- left_join(dist_data, abq_demos,
  by = c("geoid", "district")
)
```