---
title: "Demographics"
format: 
  html:
    code-fold: false
    code-tools: true
    toc: true
    toc-depth: 2
exec:
  freeze: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

I would like to compare the demographics and some key economic data between the city council districts of Albuquerque, as well as comparing Albuquerque to the rest of the state and nation. This will involve obtaining data from `census.gov` as well as local government websites. Since census tracts do not line up with council districts, the census data will need to be interpolated into the council districts.

Throughout, I will emphasize good coding techniques, especially the creation of helper functions to avoid the significant amount of repetition involved in obtaining multiple data sets for three geographical levels.

I will use the following libraries:

```{r}
#| message: false
library(tidycensus)
library(tidyverse)
library(ggspatial)
library(sf)
library(units)
library(crsuggest)
library(gt)
library(ggtext)
library(patchwork)
library(janitor)
library(nngeo)
options(tigris_use_cache = TRUE)
```

# City Council Districts

I will start by obtaining the boundaries for the city council districts. The Albuquerque website provides a wealth of data, although it is not necessarily well-structured. In many of the data sets, the relevant data is contained in tables embedded in columns of `html` tables. I'll get to that when I incorporate crime statistics later. In this case, it is useless, as the website only provides councilors' names as a variable, not district numbers, and the names provided are for councilors from several years ago!

Fortunately, the Bernalillo County website provides what's needed, as well as providing the shape for The Village of Los Ranchos de Albuquerque. Los Ranchos is sort of the Beverly Hills of Albuquerque, in the sense that it is a separate municipal enclave (almost) entirely surrounded by Albuquerque, and where many the largest properties and mansions in the area are located.

The data portal for Bernalillo County is not very friendly for programmatic downloading, though. My first effort to download and unzip the file containing the city council boundaries failed to produce a usable file. Adding the argument `method = "curl"` allowed me to download the files, however.

```{r}
#| eval: false
download.file("https://www.berncoclerk.gov/wp-content/uploads/2025/07/BERNCO_CLERKJuly2025.gdb_.zip", "data/bernalillo.zip", method = "curl")
unzip("data/bernalillo.zip", overwrite = T)
```

I will read the file with `sf::st_read`, which will result in an `sf` object. The only variable I'm interested in is the district number, so I'll just grab that, fix the variable name, and make the content more descriptive for easy mapping. The `geometry` is sticky, so I don't need to specify it.

```{r}
council_dists_bc <- st_read("data/BC_CityCouncil/ABQ_CityCouncils.shp") %>%
  select(district = DISTRICTNU) %>%
  mutate(district = paste("District", district))
council_dists_bc
```

I'll do the same for Los Ranchos.

```{r}
#| eval: false
download.file("https://www.berncoclerk.gov/wp-content/uploads/2023/02/LosRanchos.zip", "data/test.zip", method = "curl")
unzip("data/LosRanchos.zip")
```

```{r}
los_ranchos <- st_read("data/LosRanchos/LosRanchos.shp") %>%
  select(district = Name)
```


:::{.callout-tip title="Finding a Coordinte Reference System" collapse="true"}

The `crsuggest` package has a useful function to determine an appropriate coordinate reference system. 

```{r}
suggest_crs(council_dists_bc) %>% head(2)
```

:::

Now I will bind them and transform the coordinate reference system to a local one in meters.

```{r}
council_dists <- rbind(council_dists_bc, los_ranchos) %>%
  st_transform(6528)
council_dists
```

I'll plot the district boundaries with `ggplot`, adding a base map with `ggspatial::annotation_map_tile`. I'll be doing the same map several times, so I'll make a function to do so.

```{r}
dist_plot <- function(df) {
  ggplot(df) +
  annotation_map_tile(
    type = "osm",
    cachedir = "~/.cache/maps/",
    zoomin = -1,
    alpha = 0.7
  ) +
  geom_sf() +
  theme_void() +
  geom_sf(aes(color = district), fill = NA, linewidth = 1) +
  geom_sf_text(aes(label = district), fontface = "bold") +
  scale_color_viridis_d() +
  theme_void() +
  labs(title = "Albuquerque Council Districts")
}
```

```{r}
dist_plot(council_dists)
```

One thing to note from this map: there are numerous areas of unincorporated Bernalillo County which lie inside Albuquerque's borders but do not pertain to any city district. A closer look reveals some of these holes.

```{r}
dist_plot(council_dists %>% filter(district %in% c("District 2", "District 4", "District 5", "Los Ranchos")))
```

This will become relevant later as we map the census data onto the districts, as some of the data will lie outside city boundaries. On the other hand, I think it makes sense to fill in the holes in the middle of districts, to minimize the loss of data. I'll use the convenient `st_remove_holes()` function from the `nngeo` package.

```{r}
council_dists <- st_remove_holes(council_dists)
```

We can see that the holes are gone.

```{r}
dist_plot(council_dists)
```

```{r}
dist_plot(council_dists %>% filter(district %in% c("District 2", "District 4", "District 5", "Los Ranchos")))
```


#  Basic Demographics

I want to look at some basic demographic information, specifically age, sex, race and education, to see how they compare between council districts. I will use `tidycensus` to download data from `census.gov`, using tables from 2023's five-year American Community Survey. I want tract level data for my study area. I also want to compare to state-wide and country-wide data, so I'll grab those at a less detailed level.

The function to download the data is `get_acs()`, which takes a series of pretty self-explanatory arguments. Once downloaded, quite a bit of processing will be needed to prepare the data for analysis and visualization. I'll start with three functions to download the tables.


```{r}
get_tables <- 
  function(geography, table, labels,
           state = NULL, county = NULL,
           geometry = F) {
    if (geography != "us") state <- "NM"
    if (geography == "tract") county <- "Bernalillo"
    get_acs(
      geography = geography,
      state = state,
      county = county,
      table = table,
      year = 2023,
      geometry = geometry,
      cache_table = T
    ) %>%
      clean_tables(labels)
  }
```

A lot of the heavy lifting will be down the the `clean_tables` function, which hasn't been written yet. The problem is that the tables contain variable names, which are un-descriptive alpha-numeric strings. For tables and plotting, they are useless. On the other hand, manually changing the names would be terrible. Instead, I'll use `tidycensus::load_variables`, which, in addition to being useful to discover variables, also provides labels for each variable which can be used programmatically. The labels themselves need cleaning, so I'll start with a helper for all this.

```{r}
v23 <- load_variables(2023, "acs5")

get_labels <- function(table) {
  v23 %>% 
  filter(str_detect(name, table)) %>% 
  select(1:2) %>% 
  mutate(
    label = str_replace(label, "Estimate!!Total:!!", ""),
    label = str_replace(label, "Estimate!!", ""),
    label = str_replace(label, ":!!", "_"),
    label = str_replace(label, ":$", "")
  ) %>% 
  rename(variable = name)
}
```

Now I can write the clean tables function, which joins the labels to the data. I will also provide a local CRS, and calculate the area for each region. This will be needed when splitting the tract data between council districts.

```{r}
clean_tables <- function(df, labels) {
  df %>% 
    select(-2) %>% 
    clean_names() %>% 
    left_join(labels)
}
```

And, finally, the splitting function. For Bernalillo County, I want to be able to split the data by district. To do so, I will first use `st_intersection`, which will create extra rows with new geographies for each subsection of tracts which are divided among districts. The values are not automatically split among the subsections, I must do that manually. Since they are all counts (extensive variables), I will calculate the areal proportion of each subsection, and use these percentages to apportion the values among the subsections.

```{r}  
split_dists <- function(df, vars) {
  st_intersection(df, council_dists) %>% 
  mutate(
    area_split = st_area(.),
    area_pct = area_split / area,
    new_value = as.numeric(round(value * area_pct, 0))
  ) %>% 
  st_drop_geometry() %>% 
  select(all_of(vars)) %>% 
  arrange(district) %>% 
  as_tibble()
}
```


## Age and Sex

Now we can start looking at age and sex. For New Mexico and the US, I don't need the geometry, so I'll drop it. I also want to get rid of the rows which are totals and subtotals. The label is currently a combination of sex and age group, so I will split these into seperate columns. Finally, I'll just select the variables I want.

```{r}
#| message: false
age_sex_labels <- get_labels("B01001A")

age_sex_nm <- get_tables("state", "B01001A", age_sex_labels) %>% 
  filter(str_detect(label, "years")) %>% 
  separate(label, c("sex", "agegroup"), sep = "_") %>% 
  select(sex, agegroup, value = estimate) %>% 
  as_tibble()

age_levels <- age_sex_nm %>% 
  select(agegroup) %>% 
  unique() %>% 
  pull()

age_sex_nm <- age_sex_nm %>% 
  mutate(agegroup = factor(agegroup, age_levels))


age_sex_us <- get_tables("us", "B01001A", age_sex_labels) %>% 
  filter(str_detect(label, "years")) %>% 
  separate(label, c("sex", "agegroup"), sep = "_") %>% 
  select(sex, agegroup, value = estimate) %>% 
  as_tibble() %>% 
  mutate(agegroup = factor(agegroup, age_levels))

age_sex_bern <- get_tables("tract", "B01001A", age_sex_labels, geometry = T) %>% 
  st_transform(6528) %>% 
  mutate(area = st_area(.)) %>% 
  filter(str_detect(label, "years")) %>% 
  separate(label, c("sex", "agegroup"), sep = "_") %>% 
  select(geoid, sex, agegroup, value = estimate, area)

age_sex_dist <- age_sex_bern %>% 
  split_dists(c("geoid", "district", "sex", "agegroup", "value", "new_value", "area_pct")) %>% 
  mutate(agegroup = factor(agegroup, age_levels))
```

For now I'm keeping some variables I don't really need. This will help me evaluate the results of the splitting operation. I am expecting a certain amount of data loss, since parts of some of the tracts lie outside the council districts. I would like to do a visual sanity check to verify that the loss makes sense. I'll look at a small portion to see you much variance there is.

```{r}
age_sex_bern %>% 
  st_drop_geometry() %>% 
  group_by(sex, agegroup) %>% 
  summarise(value = sum(value)) %>% 
  filter(agegroup == "35 to 44 years")

age_sex_dist %>% 
  group_by(sex, agegroup) %>% 
  summarise(value = sum(new_value)) %>% 
  filter(agegroup == "35 to 44 years")
```

There is quite a difference. Let's see how this can occur by focusing in on one tract.

```{r}  
age_sex_dist %>% 
  filter(geoid == 35001003501 & agegroup == "35 to 44 years")
```

Here we are losing the majority of the population. Let's look at the tract.

```{r}
ggplot(age_sex_bern) +
  annotation_map_tile(
    type = "osm",
    cachedir = "~/.cache/maps/",
    zoomin = -1,
    alpha = 0.6
  ) +
  geom_sf(color = "black", fill = NA, linewidth = 0.5) +
  geom_sf(data = council_dists, aes(color = district), 
          fill = NA, linewidth = 2) +
  geom_sf(data = age_sex_bern %>% filter(geoid == 35001003501), 
          color = "red", fill = NA, linewidth = 2) +
  scale_fill_viridis_c() +
  scale_color_viridis_d() +
  theme_void() +
  guides(color = "none") +
  xlim(460000, 470000) +
  ylim(455000, 465000)
```

We can see that, in fact, only small portions of the tract are in each of the 2 council districts and Los Ranchos, and that most of the area is outside the city. I'm going to redo the split, dropping the unneeded variables.

```{r}
age_sex_dist <- age_sex_bern %>% 
  split_dists(c("district", "sex", "agegroup", value = "new_value"))
```

Now lets plot the results.

```{r}
mf_plot <- function(df, groups, pos) {
  df %>% 
  group_by_at(groups) %>% 
  summarise(value = sum(value)) %>% 
  mutate(value = ifelse(sex == "Male", -value, value)) %>% 
  ggplot(aes(x = value, y = .data[[groups[pos]]], fill = sex)) +
  geom_col()
}
```

```{r}
#| message: false
p1 <- age_sex_us %>%
  mf_plot(c("sex", "agegroup"), pos = 2) +
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x = element_blank()
        ) +
  xlab("US")
p2 <- age_sex_nm %>%
  mf_plot(c("sex", "agegroup"), pos = 2) +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.text.x = element_blank()) +
  xlab("New Mexico")
p1 + p2
```

```{r}
#| message: false
age_sex_dist %>% 
  filter(district != "Los Ranchos") %>% 
  mf_plot(c("district", "sex", "agegroup"), pos = 3) +
  facet_wrap(~district) +
  xlab("population") +
  theme(axis.text.y = element_text(size = 6))
  
```

TODO: add a table

## Race

I'll start with a function to clean the race tables. Unlike the former data set, the labels need some cleanup for plotting purposes. Note the conditional block where the `.` is required in the `select` statement.

```{r}
#| collapse: false

race_labels <- get_labels("B03002")

race_clean <- function(df, area) {
  df %>% 
    filter(str_detect(label, "_")) %>% 
    separate(label, c("hisp", "race"), sep = "_") %>%
    {
      if (area) select(., geoid, hisp, race, value = estimate, area)
      else select(., hisp, race, value = estimate)
    } %>%
    mutate(hisp = if_else(hisp == "Not Hispanic or Latino", "Non-Hispanic", "Hispanic"),
           race = str_replace(race, " alone$", ""),
           race = if_else(str_detect(race, "Two"), "Two or more races", race),
           race = if_else(str_detect(race, "Black"), "Black", race),
           race = if_else(str_detect(race, "Indian"), "American Indian", race)
           ) %>%
  group_by(hisp, race)
}
```

Now I can get and process the data as before.

```{r}
race_us <- get_tables("us", "B03002", race_labels) %>% 
  race_clean(area = FALSE) %>% 
  mutate(region = "US")

race_nm <- get_tables("state", "B03002", race_labels) %>% 
  race_clean(area = FALSE) %>% 
  mutate(region = "New Mexico")

race_bern <- get_tables("tract", "B03002", race_labels, geometry = TRUE) %>% 
  st_transform(6528) %>% 
  mutate(area = st_area(.)) %>% 
  race_clean(area = T) %>% 
  mutate(region = "Bernalillo")

race_dist <- race_bern %>% 
  split_dists(c("district", "hisp", "race", "value")) %>% 
  mutate(region = "Albuquerque")
```

I'll prepare the datasets for plotting. I'm going to remove two race categories for plotting purposes, both with minimal presence in New Mexico. I will use `purrr:map` and the `zeallot`'s `%<-%` operator to filter both datasets with one command. The `%<-%` operator is used to unpack the data frames from the list returned by the `map` function.

```{r}
library(zeallot)
# race_bern <- race_bern %>% st_drop_geometry() %>% select(-area)
race_compare <- rbind(race_dist %>% select(-1), race_us, race_nm)

c(race_compare_plt, race_dist_plt) %<-% 
  map(list(race_compare, race_dist), 
      \(x) (x %>% 
              filter(race != "Some other race" & 
                     race != "Native Hawaiian and Other Pacific Islander")))
```

In this case, I will forgo writing a plotting function, even though it seems repetitive. There are so many minor differences between the plots that such a function would be tedious and over complicated.


```{r}
race_compare_plt <- race_compare_plt %>% 
  mutate(region = factor(region, level = c("US", "New Mexico", "Albuquerque")))

compare_plot <- function(df, fill, x_var, position = "fill") {
  df %>% 
    ggplot(aes(x = .data[[x_var]], 
           y = value, 
           fill = .data[[fill]])) +
    geom_col(position = position) +
    scale_fill_viridis_d() +
    theme(axis.title = element_blank()) +
    labs(caption = "Source: census.gov, 2023")
}

race_compare_plt %>% 
  compare_plot("race", "region") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Racial Comparison",
       subtitle = "Albuquerque, State and Country",
       fill = "Selected Races")
```


```{r}
p1 <- race_dist_plt %>% 
  compare_plot("race", "district") +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.position = "none") +
  labs(title = "Racial Comparison by District",
       subtitle = "Percentage")
  
p2 <- race_dist_plt %>% 
  compare_plot("race", "district", position = "stack") +
  scale_y_continuous(labels = scales::comma) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(subtitle = "Population",
       fill = "Selected Races")  

p1 + p2
```


```{r}
race_compare_plt %>% 
  compare_plot("hisp", "region") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Hispanic Comparison",
       subtitle = "Albuquerque, State and Country")
```

```{r}
p1 <- race_dist_plt %>% 
  compare_plot("hisp", "district") +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.position = "none") +
  labs(title = "Hispanic Comparison by District",
       subtitle = "Percentage")

p2 <- race_dist_plt %>% 
  compare_plot("hisp", "district", "stack") +
  scale_y_continuous(labels = scales::comma) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.title = element_blank()) +
  labs(subtitle = "Population")

p1 + p2
```

There is a high degree of variability across Albuquerque, with some districts very heavily hispanic, such as the third district, and some not, such as the eighth. This is worth plotting on a map to get an idea where these districts are.


```{r}
district_population <- race_dist_plt %>% 
  group_by(district) %>% 
  summarise(pop = sum(value))

race_dist_sf <- race_dist_plt %>% 
  filter(hisp == "Hispanic") %>% 
  group_by(district) %>% 
  summarise(value = sum(value)) %>% 
  left_join(council_dists) %>% 
  st_as_sf() %>% 
  left_join(district_population) %>% 
  mutate(hisp_pct = 100 * value / pop)



dist_plot(council_dists) +
  geom_sf(data = race_dist_sf, aes(fill = value), alpha = 0.4) +
  scale_fill_viridis_c("Hispanic\nPopulation") +
  guides(color = "none")
```

```{r}
dist_plot(council_dists) +
  geom_sf(data = race_dist_sf, aes(fill = hisp_pct), alpha = 0.4) +
  scale_fill_viridis_c("Hispanic\nPercent") +
  guides(color = "none")
```



## Education


```{r}
edu_labels <- get_labels("B15002")

edu_us <- get_tables("us", "B15002", edu_labels) %>% 
  filter(str_detect(label, "_")) %>% 
  separate(label, c("sex", "education"), sep = "_") %>% 
  select(sex, education, value = estimate)

edu_levels <- edu_us %>% 
  select(education) %>% 
  unique() %>% 
  pull()

edu_levels <- c("No HS Diploma", "High school graduate", "Some college", edu_levels[-(1:11)])

simplify_education <- function(df) {
  df %>% 
    mutate(
    education = case_when(
      education == "No schooling completed" |
        education == "Nursery to 4th grade" |
        education == "5th and 6th grade" |
        education == "7th and 8th grade" |
        education == "9th grade" |
        education == "10th grade" |
        education == "11th grade" |
        education == "12th grade, no diploma"
        ~ "No HS Diploma",
      education == "Some college, 1 or more years, no degree" |
        education == "Some college, less than 1 year"
        ~ "Some college",
      education == "High school graduate (includes equivalency)"
        ~ "High school graduate",
      TRUE ~ education
      )
    ) %>%
    {
      if ("district" %in% names(.)) group_by(., district, sex, education) 
      else group_by(., sex, education)
    } %>% 
    summarise(value = sum(value)) %>% 
   
  mutate(education = factor(education, edu_levels))
}


edu_us <- edu_us %>% simplify_education()
```

```{r}

# edu_us <- get_tables_us("B15002", edu_labels) %>% 
#   st_drop_geometry() %>% 
#   filter(str_detect(label, "_")) %>% 
#   separate(label, c("sex", "education"), sep = "_") %>% 
#   select(sex, education, value = estimate) %>% 
#   simplify_education() %>% 
#   mutate(education = factor(education, edu_levels))

edu_nm <- get_tables("state", "B15002", edu_labels) %>% 
  filter(str_detect(label, "_")) %>% 
  separate(label, c("sex", "education"), sep = "_") %>% 
  select(geoid, sex, education, value = estimate) %>% 
  simplify_education()

edu_bern <- get_tables("tract", "B15002", edu_labels, geometry = T) %>% 
  filter(str_detect(label, "_")) %>% 
  separate(label, c("sex", "education"), sep = "_") %>% 
  select(geoid, sex, education, value = estimate) %>% 
  st_transform(6528) %>% 
  mutate(area = st_area(.))

edu_dist <- edu_bern %>% 
  split_dists(c("district", "sex", "education", "value")) %>% 
  simplify_education()
```

I can use the `mf_plot` function I created earlier, and add the appropriate labels.

```{r}
p1 <- edu_us %>%
  mf_plot(c("sex", "education"), 2) +
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x = element_blank()
        ) +
  xlab("US")
p2 <- edu_nm %>%
  mutate(education = factor(education, edu_levels)) %>% 
  mf_plot(c("sex", "education"), 2) +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.text.x = element_blank()) +
  xlab("New Mexico")
p1 + p2
```

```{r}
edu_dist <- edu_dist %>% 
  mutate(value = value / 1000) %>% 
  filter(district != "Los Ranchos")

edu_dist %>% 
  mf_plot(c("district", "sex", "education"), 3) +
  labs(x = "Population (in thousands)",
       title = "Educational attainment in Albuquerque",
       subtitle = "By district") +
  theme(axis.text.y = element_text(size = 6))
```

```{r}
edu_dist %>% 
  mf_plot(c("district", "sex", "education"), 3) +
  facet_wrap(~district) +
  labs(x = "Population (in thousands)",
       title = "Educational attainment in Albuquerque",
       subtitle = "By district") +
  theme(axis.text.y = element_text(size = 6))
```

```{r}
edu_us_graduate <- edu_us %>% 
  mutate(degree = if_else(as.integer(education) <= 3, "No degree", "Degree"),
         region = "US")
edu_nm_graduate <- edu_nm %>% 
  mutate(degree = if_else(as.integer(education) <= 3, "No degree", "Degree"),
         region = "New Mexico")
edu_dist_graduate <- edu_dist %>% 
  group_by(sex, education) %>% 
  summarise(value = sum(value)) %>% 
  mutate(degree = if_else(as.integer(education) <= 3, "No degree", "Degree"),
         region = "Albuquerque")
edu_graduate_compare <- rbind(edu_dist_graduate, edu_nm_graduate, edu_us_graduate)
```

```{r}
edu_graduate_compare %>% 
  mutate(
    region = factor(region, 
                    level = c("US", "New Mexico", "Albuquerque"))
  ) %>% 
  compare_plot("degree", "region") +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.title = element_blank()) +
  labs(title = "Percentage with college degrees",
       subtitle = "Albuquerque, State and Country") 
```


cd_demographics %>% 
  st_drop_geometry() %>%
  group_by(district) %>% 
  summarise(
    population = sum(population), 
    med_house_value = median(med_house_value),
    income_hh = median(income_hh),
    med_age = median(med_age),
    in_poverty = sum(in_poverty),
    gini = mean(gini)
  ) %>% 
  select(district, population, med_house_value, income_hh, med_age, in_poverty, gini) %>%
  gt(rowname_col = "district") %>%
  cols_label(
    population = "Population",
    med_house_value = "Med House Value",
    income_hh = "HH Income",
    med_age = "Age",
    gini = "GINI"
  ) %>%
  tab_header(
    title = md("**Basic Demographics for Albuquerque**")
  ) %>%
  tab_source_note(
    source_note = md("*Source: census.gov, acs5, 2023*")
  ) %>%
  opt_stylize(style = 2) %>%
  fmt_currency(
    columns = c(med_house_value, income_hh),
    decimals = 0
  ) %>%
  fmt_number(columns = "med_age", decimals = 1) %>%
  fmt_number(columns = "gini", decimals = 3) %>%
  fmt_number(columns = "population", decimals = 0)
