---
title: "District Variance"
format: 
  gfm:
    toc: true
---

# Introduction

This article follows on from a [recent article](https://biscotty.net/posts/data-science/r-census/abq-demographics), in which I used data from the US Census Bureau to see how age, race, and education vary between city council districts in Albuquerque, New Mexico. In some cases, the variations were striking. Now I would like to apply some formal testing to see if this impression of variance is backed up statistically. Before being able to run the tests, and eventual models, a significant amount of data wrangling will be necessary in order to get the data in an appropriate form, and engineer the desired variables from the raw data. As in the prior article, I will use data from the US Census American Community Survey. In addition to the demographic variables of education and race, I will add a number of economic indicators such as income levels, housing values, and inequality measurements. 

Having obtained the data, I will need to "wrangle" it by cleaning out bad data, filling in missing values, engineering new variables of interest, and dividing the spatial dataset by council districts. I will then explore correlation and _spatial autocorrelation_ between the chosen variables. Next I will perform an analysis of variance, but since almost none of the variables will turn out to be normally distributed, some work will be necessary. I will also need to address issues of homogeneity of variance among the districts. Finally, I will do some regression modeling. Since my primary target variable is the county districts, this will require multinomial logistic regression models.

The current article will primarily be concerned with preparing the data and looking at correlation and spatial autocorrelation in the variables. But I do have a secondary purpose: to show how data analysis and transformation can be significantly sped up by using the `collapse` package instead of the ubiquitous `dplyr` library. `dplyr` is without doubt a wonderfully expressive library and is a key part of the popular `tidyverse`. It sets an excellent syntactic standard, but it is not, unfortunately, particularly fast. `collapse` is, and it is even faster than `data.table`. Written in C/C++, it has "fast" versions of many commonly-used `dplyr` functions, simply with a prepended "f", eg. `fmutate` rather than `mutate`. It provides many new fast statistical functions, transformations, and convenience functions. Throughout, I will show speed comparisons between `dplyr` and `collapse` using the `microbenchmark` package.

# Loading libraries

I will begin by loading the required libraries and setting some constants. This project will require functions from about 30 libraries, but in most cases only one or two functions from a given library are needed. Rather than loading all of the packages, I will only load the primary libraries, but only the necessary functions from the rest of them.

```{r}
#| message: false
options(paged.print = FALSE,
        tigris_use_cache = TRUE)
libraries <- list(
  "sf", "collapse", "ggplot2", "magrittr", "dplyr",
  "zeallot", "purrr", "patchwork", "spdep", "rgeoda"
)
invisible(lapply(libraries, library, character.only = TRUE))

microbenchmark <- microbenchmark::microbenchmark
glue <- glue::glue
annotation_map_tile <- ggspatial::annotation_map_tile
ggboxplot <- ggpubr::ggboxplot
ggqqplot <- ggpubr::ggqqplot
set_units <- units::set_units
st_remove_holes <- nngeo::st_remove_holes
ggpairs <- GGally::ggpairs
get_acs <- tidycensus::get_acs
rownames_to_columns <- tibble::rownames_to_column
classIntervals <- classInt::classIntervals
str_replace_all <- stringr::str_replace_all
stat_density_ridges <- ggridges::stat_density_ridges
melt <- reshape2::melt
tidy <- broom::tidy

year <- 2024
crs <- 6528
caption <- glue("Source: census.gov, acs5, {year}")
```

# Preparing the data

## Council District Geometries

I'll start by preparing the council district spatial dataset. I did the same in the last article, but this time I'll use `collapse` functions. 

```{r}
council_dists <- st_read("data/BC_CityCouncil/ABQ_CityCouncils.shp") %>% 
  fselect(district = DISTRICTNU) %>% 
  ftransform(district = paste("Dist", district)) %>%
  roworder(district) %>% 
  st_remove_holes %>% 
  st_cast("MULTIPOLYGON")
los_ranchos <- st_read("data/LosRanchos/LosRanchos.shp") %>%
  fselect(district = Name)
council_dists <-
  rowbind(council_dists, los_ranchos) %>%
  st_transform(crs)
```

`fselect` works just like `select`. `collapse` also has an `fmutate` function which is the fast version of `mutate`, and which I could have used here. When no grouping is required, `ftransform` is even faster than `fmutate` because it evaluates all arguments simultaneously. `roworder` replaces `arrange`, and `rowbind` replaces `rbind`. Finally, `qF` is used instead of `factor`. `collapse` offers a number of functions for fast conversion between data types, such as `qTBL` to convert to `tibble`s, `qM` to convert to matrices, and `qDF` to convert to data frames. _In general, the more primitive the type, the faster the processing_. When spatial information is unnecessary, I will use data frames for processing, converting back to spatial data frames for mapping.

Note that, after removing the "holes", the geometries of `council_dists` will be a mix of POLYGONSs and MULTIPOLYGONs. This will generate errors in various future operations, so I will use `st_cast` to ensure consistency.

My study area is a rectangular region that is a bit smaller than the actual city limits. This will cut out some outlying areas. Also, as mentioned before, there are many patches of unincorporated county in the geographic area. Above I used `st_remove_holes` to include enclaves of unincorporated county in the respective districts. For the rest, I will explicitly label these areas as "Unincorporated", obtaining the geometries for the region by taking the `st_difference` between the rectangular bounding box of the `council_dists` and the union of the council district geometries. 

```{r}
#| warning: false
dist_box = c(xmin = 452000, xmax = 479833, 
             ymin = 443500, ymax = 467858)

council_dists %<>% 
  st_crop(dist_box) %>% 
  st_cast("MULTIPOLYGON")

difference_sfc <-
  st_difference(
    st_as_sfc(st_bbox(council_dists)),
    st_union(council_dists$geometry)
  )
difference_sf <- 
  st_sf(data.frame(district = "Unincorporated"),
  geometry = difference_sfc
)
council_dists <- rowbind(difference_sf, council_dists)
```

`rowbind` is a fast function from `collapse`. I have used `%<>%`, a lesser-known function from `magrittr`, which replaces the tedious yet ubiquitous idiom `data <- data %>%` with the simpler and shorter `data %<>%`. I have also, again, needed to recast the geometries after the cropping operation. Now I will create a base plot for the rest of the study which shows the district boundaries.

```{r}
#| warning: false
dist_plot <- function(labels = TRUE) {
  p <- ggplot(council_dists) +
    annotation_map_tile(
      type = "osm", zoomin = -1, cachedir = "~/.cache/maps/"
    ) +
    geom_sf(aes(color = district), linewidth = 1) +
    labs(title = "Albuquerque, NM",caption = caption) +
    theme_void()
  if (labels) {
    p <- p + geom_sf_text(aes(label = district)) + guides(color = "none")
  }
  p
}

dist_plot(labels = FALSE)
```

## Census Data

### Fetching

The `get_acs()` function from `tidycensus` obtains information from the American Community Survey tables published by the US Census Department. By supplying a named vector to the `variables` argument, the columns of the returned data table will be automatically renamed. Setting `geometry = TRUE` returns an `sf` object, and `output = "wide"` gives a column for each variable. After obtaining the data, I will remove the margin of error columns, remove the trailing "E" from the remaining column names, and transform the geometry to an appropriate coordinate reference system. `get_vars` is a more flexible version of `fselect`. As you can see, it can select by either index or name, and can also use regular expressions for column selection.

```{r}
vars_census <- c(
  House.Value.Med = "B25077_001",
  Housing.Units = "B25001_001",
  Renter.Occupied = "B25003_003",
  Vacant = "B25002_003",
  Income.Med = "DP03_0062",
  Gini = "B19083_001",
  Rent.Med = "B25031_001",
  Foreign.Born.Pct = "DP02_0094P",
  Hisp.Pct = "DP05_0090P",
  In.Poverty = "B17001A_002",
  Age.Med = "B01002_001",
  College.Pct = "DP02_0068P",
  Population = "B01003_001"
)

bern_data <- get_acs(
  geography = "tract",
  state = "NM",
  county = "Bernalillo",
  variables = vars_census,
  output = "wide",
  year = year,
  geometry = TRUE,
  cache_table = TRUE
) %>%
  get_vars(c("E$", "GEOID"), regex = TRUE) %>%
  get_vars(-2) %>%
  frename(\(x) sub("E$", "", x)) %>%
  st_transform(crs)
```

### Cleaning

Let's take a look at the data before splitting it between districts. I'll use `qsu`, `collapse`'s fast version of `summary`.

```{r}
qsu(bern_data)
```

There is some clean-up to do. There are some missing values to explore, and oddities like areas where the population total is 0. I'll deal with the zero-population areas first. `fsubset` here is like `dplyr`'s `filter`. `collapse` even provides a fast versions of the logical operators `==`, `!=`, and `%in%` (`%==%`, `%!=%`, `%iin%`).

```{r}
bern_data %>% 
  fsubset(Population %==% 0)
```

It is worth comparing this with the `dyplr` equivalent.

```{r}
microbenchmark(
  dp = bern_data %>% filter(Population == 0),
  co = bern_data %>% fsubset(Population == 0)
)
```

The `collapse` version of this very common operation is almost 100 times faster than `dplyr`'s! Now, let's see where the tract is.

```{r}
dist_plot() +
  geom_sf(data = bern_data, color = "red") +
  geom_sf(data = fsubset(bern_data, Population %==% 0),
          fill = "navy") +
  lims(x = dist_box[1:2], y = dist_box[3:4])
```

This tract is the state fairgrounds. The other must be outside the study area. I'll remove both.

```{r}
bern_data %<>% fsubset(Population %!=% 0)
qsu(bern_data)
```

The population of 16 is somewhat suspicious. 

```{r}
dist_plot() +
  geom_sf(data = bern_data, color = "red") +
  geom_sf(
    data = fsubset(bern_data, Population < 20),
    fill = "navy"
  ) +
  lims(x = dist_box[1:2], y = dist_box[3:4])
```

This is the Petroglyph National Monument, and the 16 people must be park rangers. I'll remove this as well.

```{r}
bern_data %<>% fsubset(Population > 20)
```

This only leaves the missing values. Let's take a closer look. 

```{r}
bern_data %>%
    fsubset(missing_cases(bern_data))
```

As you might guess `missing_cases` is from `collapse`, and is the fast, though opposite, of `complete.cases`. `complete.cases`, however, cannot be used on an `sf` object without dropping the geometry, while `missing_cases` can, and it is nearly 100 times faster than the `dplyr` idiom. (One important aspect of `collapse` is that it preserves attributes, and is generally agnositc as to the class of a table, although this is not entirely seamless for `sf` objects.)

```{r}
microbenchmark(
  dp = bern_data %>%
    filter(if_any(everything(), is.na)),
  co = bern_data %>%
    fsubset(missing_cases(bern_data))
)
```

Again, nearly 100 times faster.

These observations do seem worth preserving. I will fill in the missing values based on the median values for each variable, mean in the case of Gini, but I would like these values to be based on the averages for the specific district, rather than an overall averages, so I will go ahead and split the data into districts at this point.

I want to remember the rows with missing values so I can check that they get filled in. I can't just use row indices, since these will change after splitting into districts. Instead I'll grab the `GEOID`s. I am also using shorthand verbs provided by `collapse`: `gv`, short for `get_vars`, and `ss`, short for `fsubset`.

```{r}
missing <- (bern_data %>%
  st_drop_geometry() %>%
  ss(missing_cases(bern_data)))[["GEOID"]]
```

I will preserve the county data for future use.

```{r}
st_write(bern_data, "data/district_variance.gpkg",
    layer = "bern_data", append = F)
```

## The Albuquerque District Dataset

Splitting the data involves intersecting the county census data with the geometries from the district dataset. Any given tract's boundaries may cross two or even three districts. Since some of the variables are extensive, representing counts, these values will need to be apportioned based on the percentage area represented by each subsection. I'll calculate the initial area of each tract, then the areas of each newly-created sub-section, which will then allow me to divide each tract's data by the percentage area lying in the newly created areas. I'll also make `district` a factor. `qF` is a faster form of `factor`.

```{r}
bern_data <- st_read("data/district_variance.gpkg", layer = "bern_data")
```


```{r}
#| warning: false
abq_data <- bern_data %>%
  ftransform(area = st_area(.)) %>%
  st_intersection(council_dists) %>%
  ftransform(area_pct = as.numeric(st_area(.) / area)) %>% 
  fsubset(district %!iin% c("Los Ranchos", "Unincorporated")) %>% 
  ftransform(district = qF(district))
```

### Variable Engineering

I want to create some new variables from the raw data. First, I will turn the raw poverty numbers into a percentage of each tract's population. I'll also create housing and population density variables based on tract area, and vacancy and renter occupied variables that are a percentage of available housing units in each tract.

`ftransformv` takes a vector of column identifiers, in this case column names, and applies a function to them, in this case an anonymous function, using the `FUN` argument. `ftransform` is similar to `fmutate`, although it has some different rules, such as the variables being non-recursive. I'm also taking advantage of some quick column arithmetic. `%c/%` is shorthand for column division. `collapse` provides a range of such functions for row and column math, eg. `%c*%`. Note also that `ftransform` allows you to remove colums easily with the syntax `variable = NULL`. 

Finally, I will use `st_cast` to ensure consistent geometries. After intersection operation, there are likely to be a mix of POLYGONs and MULTIPOLYGONs, which will cause trouble down the line for the spatial analyses.

```{r}
ext_vars <- c("Population", "Housing.Units", "Renter.Occupied", "Vacant", "In.Poverty")

abq_data %<>%
  ftransformv(
    ext_vars,
    FUN = function(x) round(x * abq_data$area_pct, 0)
  ) %>%
  ftransform(
    In.Poverty.Pct = In.Poverty %c/% Population * 100,
    Population.Density = Population %c/% area %>%
      set_units("1/km^2") %>% as.numeric(),
    Vacant.Pct = Vacant %c/% Housing.Units * 100,
    Renter.Occupied.Pct = Renter.Occupied %c/% Housing.Units * 100,
    Housing.Density = Housing.Units %c/% area %>%
      set_units("1/km^2") %>% as.numeric(),
    In.Poverty = NULL, Vacant = NULL, Housing.Units = NULL,
    Renter.Occupied = NULL, area = NULL
  ) %>%
  colorder(district, pos = "front") %>% 
  st_cast("MULTIPOLYGON")
```

### More Cleaning

Now I want to fill in the missing `NA`s with grouped averages. I'll start by creating a grouping object. I could simply provide a column, but it is somewhat faster to do it this way if you plan to re-use the same grouping multiple times. This is similar, to the `ftransform` above, but the syntax is a little more complicated since I need to supply the `g` (group) and `TRA` (transformation) arguments. There are ten different possible values for `TRA`, including "fill" for replacing all variables, "-" for subtracting (centering) and "/" for dividing (scaling). In this case I am simply replacing the `NA` values.

I am modifying the data by reference here by using `settransform` rather than `transform`. A number of `collapse`'s functions provide this alternative.


```{r}
dist_grp <- GRP(abq_data$district)

settransform(
  abq_data,
  fmean(list(
    In.Poverty.Pct = In.Poverty.Pct, Vacant.Pct = Vacant.Pct,
    Renter.Occupied.Pct = Renter.Occupied.Pct),
    g = dist_grp, TRA = "replace_NA"))

settransform(
  abq_data,
  fmedian(list(
      House.Value.Med = House.Value.Med, 
      Rent.Med = Rent.Med, Income.Med = Income.Med),
    g = dist_grp, TRA = "replace_NA"))

abq_data %>%
  fsubset(GEOID %iin% missing)
```

All missing values have been filled in. However, there is a new problem:

```{r}
qsu(abq_data)
```

The split data now contains new areas with a population of 0. As a result of the splitting, there are numerous small fragments of tracts containing little if any information. To address this, I will remove all of the new rows which represent less than 1\% of the original area. I'll also remove unnecessary columns.

```{r}
abq_data %<>% 
  fsubset(area_pct > .01) %>% 
  ftransform(GEOID = NULL, Population = NULL, area_pct = NULL)
qsu(abq_data)
```

As an aside, it is worth noting that the `geometry` column does not show up here, as it does if you use `summary`. One of the hallmarks of the `collapse` package is that it leaves alone attributes, basically ignoring them but passing them on. Also, `qsu` can be used to produce weighted and grouped summaries. For example

```{r}
qsu(abq_data, Income.Med ~ district) %>% 
  qDF %>% rownames_to_columns(var = "district") %>% 
  ggplot(aes(x = district, y = Mean)) +
  geom_col() +
  geom_errorbar(aes(ymin = Mean - SD, ymax = Mean + SD), 
                color = "black", linewidth = 1) +
  labs(title = "Median Income by District in Albuquerque", caption = caption)
```

Before going further, I want to split the data from the geometries, so I can work with a simpler object for data processing. I'll preserve the geometries so I can recreate the `sf` object when I want to do mapping or spatial autocorrelation. `st_drop_geometry` returns a `tibble` which, while convenient for many things, carries extra overhead. I will use the fast `qDF` to convert to a simple data frame. Since I will be converting between `sf` and `data.frame` multiple times, I'll make a couple helpers to keep the code clean and concise. Taking advantage of `%<>%`, conversion can be made with a simple `abq_data %<>% to_df`.

```{r}
to_sf <- function(df, geometry = abq_data_geom) {
  st_sf(df, geometry = geometry)
}
to_df <- function(df) {
  st_drop_geometry(df) %>% qDF
}

abq_data_geom <- st_geometry(abq_data)
abq_data %<>% to_df
```

# Exploring the data

## Visualization

I'll start by visualizing the data. First, I'd like to visualize the overall distribution of each variable, to see how they are distributed. I will construct a series of histograms and qqplots for this purpose. Note the `.data[[var]]` syntax in the plotting function, which is necessary to pass variables to `aes()` when using `ggplot` _within_ a function. Then I will use `map()` to save each plot with `zeallot`'s assignment operator `%<-%`, after which I can review all or select plots.

I use `num_vars` here, which is a useful function that efficiently selects the numeric variables from a data frame, or any similar object. The `return` argument makes it very flexible. In this case, rather than grabbing the actual data, I'm just grabbing the variable names.

```{r}
num_var_names <- num_vars(abq_data, return = "names")

plot_hist <- function(df, var) {
  df %>%
    ggplot(aes(.data[[var]])) +
    geom_histogram(aes(y = after_stat(density)),
      fill = "lightgray", col = "black", bins = 20
    ) +
    geom_line(stat = "density", adjust = 2, color = "blue") +
    labs(caption = caption)
}

hists <- map(num_var_names, \(x) abq_data %>% plot_hist(x))

qqs <- map(num_var_names, \(x) ggqqplot(abq_data[[x]]))

c(
  pHouse.Value.Med, pGini,
  pRent.Med, pAge.Med, pIncome.Med, pForeign.Born.Pct,
  pHisp.Pct, pCollege.Pct, pPopulation.Density,
  pVacant.Pct, pRenter.Occupied.Pct, pHousing.Density
) %<-%
  map(
    1:length(hists),
    \(x) hists[[x]] + qqs[[x]]
  )
pHouse.Value.Med; pGini; pIncome.Med; pHisp.Pct; pCollege.Pct
```

The data is not, for the most part, normally distributed. This will pose challenges to analysis of variance, and will certainly require appropriate transformations. It is interesting to get a sense of the geographic distribution with regard to these variables. I want to split the data into quantiles rather than using a continuous scale. This is where `tmap` really shines, and one of the few areas where `ggplot` makes life a bit difficult, since the cuts must be manually generated prior to plotting. Never-the-less, I will use `ggplot`.  

The `classInt` package will generate the break points, which I will use to `cut` the data. I find the bracket/parenthesis syntax for ranges visually tedious, hence the `str_replace` funcitons. Also note the use of `%!in%`, another fast function from `collapse`.

```{r}
#| message: false
abq_data %<>% to_sf
make_cuts <- function(var) {
  quantiles <- classIntervals(var, 5, style = "quantile")
  cut(
    var, quantiles$brks,
    include.lowest = TRUE, dig.lab = 6
  ) %>%
    str_replace_all(",", " - ") %>%
    str_replace_all("[\\(\\]\\[]", "")
}

cuts <- map(num_var_names,
    \(x) make_cuts(to_df(abq_data)[[x]]))

quantile_map <- function(cut, lab) {
  dist_plot() +
      geom_sf(
        data = fselect(abq_data, num_var_names),
        aes(fill = cut), alpha = 0.5
      ) +
      geom_sf_text(
        data = council_dists %>%
          fsubset(district %!iin% c("Los Ranchos", "Unincorporated")),
        aes(label = district)
      ) +
      scale_fill_viridis_d() +
      labs(fill = lab) +
      theme_void()
}

c(
  pHouse.Value.Med, pGini,
  pRent.Med, pAge.Med, pIncome.Med, pForeign.Born.Pct,
  pHisp.Pct, pCollege.Pct, pPopulation.Density,
  pVacant.Pct, pRenter.Occupied.Pct, pHousing.Density
) %<-%
  map2(
    cuts, num_var_names,
    \(cut, lab) quantile_map(cut, lab)
  )

pHouse.Value.Med; pIncome.Med; pGini; pHisp.Pct; pCollege.Pct
```

Housing values and income levels seem to be geographically distributed similarly, as one might expect, with higher levels in the more outlying areas and lower levels in centrally located areas. The patches of high income and high percent of college graduatges in the central areas are those aroung the University of New Mexico. This "abberation" is evident in all the maps. Inequality levels seem to follow an opposite tendency, with higher levels in the interior of the city. 

The percentage of Hispanics is highest in the southwest of the city, decreasing northward and, much more significantly, eastward. It is in some ways the mirror image of the percentage of college graduates. When looking at variable correlation, I expect to see a strong negative correlation between these variables.

Now, let's look at the variation of variables across districts. For that, I will use boxplots and ridge plots.

```{r}
plot_dist <- function(df, var) {
  df %>%
    ggboxplot("district", var, fill = "district") +
    geom_point(position = position_jitter(width = 0.1)) +
    scale_fill_viridis_d() +
    theme(
      axis.text.x = element_text(
        angle = 45, vjust = .5, hjust = .5)) +
    guides(fill = "none") +
    labs(caption = caption)
}

c(
  pHouse.Value.Med, pGini,
  pRent.Med, pAge.Med, pIncome.Med, pForeign.Born.Pct,
  pHisp.Pct, pCollege.Pct, pPopulation.Density,
  pVacant.Pct, pRenter.Occupied.Pct, pHousing.Density
) %<-%
  map(num_var_names, \(x) abq_data %>% plot_dist(x))
pHouse.Value.Med; pGini; pIncome.Med; pHisp.Pct; pCollege.Pct
```


```{r}
#| message: false
#| warning: false
plot_dist_ridge <- function(df, var) {
  ggplot(df, aes(
    x = .data[[var]], y = district,
    fill = factor(stat(quantile))
  )) +
    stat_density_ridges(
      geom = "density_ridges_gradient", calc_ecdf = TRUE,
      quantiles = 4, quantile_lines = TRUE, alpha = 0.7,
      jittered_points = TRUE, position = "points_sina"
    ) +
    scale_fill_viridis_d(name = "Quartiles") +
    theme(axis.title.x = element_blank())
}

c(
  pHouse.Value.Med, pGini,
  pRent.Med, pAge.Med, pIncome.Med, pForeign.Born.Pct,
  pHisp.Pct, pCollege.Pct, pPopulation.Density,
  pVacant.Pct, pRenter.Occupied.Pct, pHousing.Density
) %<-%
  map(
    num_var_names,
    \(x) abq_data %>%
      plot_dist_ridge(x) +
      labs(title = x)
  )
pHouse.Value.Med; pGini; pIncome.Med; pHisp.Pct; pCollege.Pct
```

These plots confirm some of the observations based on the maps. They also reveal how different the distributions are shaped. In most variables, district 3 has a rather compact distribution, which is closer to normal than any other district. On the other hand, the distributions of the fourth district are generally very wide and not at all normal. The sixth and eighth district show the same tendancy for some of the variables.

The apparent non-homogeneity of variance across the districts will be another issue to address when performing the analysis of variance.

## Correlation

Next, let's look at the correlation between variables.

```{r}
#| message: false
abq_data %<>% to_df
ggpairs(num_vars(abq_data))
```

```{r}
correlation <- cor(num_vars(abq_data)) %>% melt
ggplot(correlation) +
  geom_tile(aes(x = Var1, y = Var2, fill = value)) +
  labs(title = "Correlation Heatmap",
       x = "Variable 1",
       y = "Variable 2") +
  scale_fill_distiller(palette = "RdBu", direction = 1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = NULL, fill = "Corr", caption = caption)
```

As anticipated, a strong negative correlation between areas with high percentage of Hispanics and those with high percentage of college grads jumps out in the heatmap. It is not surprising that many variables are correlated, some quite highly. for example housing density and population density, or housing values with income, and even more strongly with percentage of college graduates. On the other hand, the negative correspondance between house prices and Hispanics is strong.

Economic inequality is most evident in areas with a high percentage of rentals, vacancies and lower rent prices. Areas with higher incomes and lower poverty levels have less economic inequality.

In data in which so many variables appear correlated, it is interesting to note those which are not. For example, inequality does not seem to be related to any of the demographic variables, only the economic indicators. Poverty seems not to care about age, race, or where one is from.

The percentage of rentals in a given area have a strong negative correlation with the income levels in the area. Some of the negative correlations are striking. Hispanic percentage is strongly negatively correlated with both housing values and percentage of people with a college education.

## Spatial autocorrelation

We are often influenced by our neighbors. Birds of a feather flock together, as they say. Or, more canonically, “everything is related to everything else, but near things are more related than distant things.” (Waldo Tobler’s First Law of Geography)  Spatial autocorrelation is based on this idea. Neighbor relationships can be generally defined in two ways: if only considering regions which share a border to be neighbors, this is described as a "rook" relationship. Were one to also include areas which merely touch, which are in a sense "diagonal" to each other, would be a "queen" relationship. Since Albuquerque is laid out almost entirely as a strict rectangular grid, with little opportunity for diagonal movement, I will use the "rook" relationship.

### Global

Moran's $I$ is commonly used to test whether or not variables are randomly distributed through an area. Significant p-values indicate spatial autocorrelation, in other words, that variable values in one area are correlated to their neighbors. Performing the test requires creating a neighbor list and, from that, a weights list, which is then passed to the `moran.test` function.


```{r}
abq_data %<>% to_sf(abq_data_geom)
nb <- poly2nb(abq_data, queen = F, snap = 25)
nbw <- nb2listw(nb, style = "W", zero.policy = T)

vars_list <- to_df(abq_data) %>% num_vars

map(
  vars_list,
  \(x) moran.test(x, nbw, alternative = "two.sided") %>% tidy
) %>%
  unlist2d %>% 
  get_vars(c(1, 5:6))
```

The positive statistic values indicate positive spatial autocorrelation, or clustering of similar values. Negative values would indicate the opposite, ie. clustering of dissimilar values. The low p-values show that all of these are statistically significant. Higher values indicate stronger autocorrelation. In this case, all of the variables show such correlation.

One way to visualize this is with `moran.plot`, which produces spatially lagged plots. I will only select five of the variables for this exercise, and will subset the data in the interest of efficiency.

```{r}
#| message: false
#| output: false
abq_data_ss <- abq_data %>% 
  fselect(House.Value.Med, Income.Med, Gini, Hisp.Pct, College.Pct)
vars_list_ss <-  num_vars(to_df(abq_data_ss))
num_var_names_ss <- num_vars(abq_data_ss, return = "names")
mplots <-
  map2(
    vars_list_ss,
    num_var_names_ss,
    \(x, y) moran.plot(x, nbw, xlab = y)
  )
```

The spatial lag plots show the spatial autocorrelation of the variables, as they are generally gathered near the mean line. The lower left quadrant contains tracts with low values clustered with other low values (low-low), while the upper right quadrant contains observations with spatially autocorrelated high values (high-high). Both the Hispanic and college percentages have high statistic values (15.80 and 13.46), and the plots show most observations tightly grouped along the diagonal. Gini, on the other hand, has the relatively low value of 5.32, and the graph displays how much wider spread the laggged observations are.

### Local Clusters

To get a sense of where these variables are, and where similar values are clustered, we need to use the local versions of the Moran test, which compares each tract to it's neighbors. The `localmoran` function does the calculation.

```{r}
lmorans <- map(num_var_names_ss,
               \(x) localmoran(abq_data_ss[[x]], nbw, alternative = "two.sided"))
```

I will use this to produce maps showing 

```{r}
#| message: false
moran_map <- function(cut) {
  dist_plot() +
      geom_sf(
        data = abq_data_ss,
        aes(fill = cut), alpha = 0.5
      ) +
      geom_sf_text(
        data = council_dists %>%
          fsubset(district %!iin% c("Los Ranchos", "Unincorporated")),
        aes(label = district)
      ) +
      theme_void()
}

mmaps <- map(
  seq(1, length(num_var_names_ss)),
  \(idx) {
    p1 <- quantile_map(
      make_cuts(abq_data_ss[[num_var_names_ss[idx]]]),
      glue("{num_var_names_ss[idx]}")
    )
    p2 <- moran_map(
      cut(lmorans[[idx]][, "Z.Ii"], c(-Inf, -1.96, 1.96, Inf))
    ) +
      scale_fill_viridis_d(
        glue("{num_var_names_ss[idx]}\nLocal Moran's I"),
        labels = c("Negative SAC", "No SAC", "Positive SAC")
      )
    p3 <- moran_map(
      cut(lmorans[[idx]][, "Pr(z != E(Ii))"], c(-Inf, 0.05, Inf))
    ) +
      scale_fill_viridis_d(
        glue("{num_var_names_ss[idx]}\np-value"),
        labels = c("< 0.05", ">= 0.05")
      )
    list(p1, p2, p3)
  }
)

(lmaps <- c(
  lmpHouse.Value.Med, lmpIncome.Med, 
  lmpGini, mpHisp.Pct, lmpCollege.Pct
) %<-% map(mmaps,
    \(x) {
      c(p1, p2, p3) %<-% x
      p1 + plot_spacer() + p2 + p3
    }))
```

The Moran plots we saw earlier were divided into quadrants, the lower left being areas of "low-low" correlation, with the upper right being "high-high". As a final exercise, I will map the statistically significant areas of correlation. I will start by scaling the data with `collapse`'s `fscale`. 

```{r}
abq_data_scaled <- abq_data_ss %>% to_df %>% fscale %>% to_sf
mplots <-
  map2(
    num_vars(to_df(abq_data_scaled)),
    num_var_names_ss,
    \(x, y) moran.plot(x, nbw, xlab = y)
  )

plot_nbr <- function(df, var, fill = quadrant) {
  dist_plot() +
    geom_sf(data = df, aes(fill = quadrant), 
            alpha = 0.3, show.legend = T) +
    scale_fill_discrete(
      var,
      palette = c("red", "blue", "lightpink", "skyblue2", "white"),
      labels = c(
        "High-High", "Low-Low", "High-Low",
        "Low-High", "Non-significant"
      ),
      drop = FALSE
    )
}

map_clusters <- function(var) {
  data <- fselect(abq_data_scaled, var)
  lmp <- localmoran(data[[var]],
    nbw,
    alternative = "two.sided"
  )[, 5]

  mpc <- mplots[[var]]

  data %<>% fmutate(
    lmp = lmp,
    quadrant = factor(
      case_when(
        lmp > 0.05 ~ 5,
        mpc$x >= 0 & mpc$wx >= 0 ~ 1,
        mpc$x <= 0 & mpc$wx <= 0 ~ 2,
        mpc$x >= 0 & mpc$wx <= 0 ~ 3,
        mpc$x <= 0 & mpc$wx >= 0 ~ 4
      ),
      levels = 1:5
    )
  )

  plot_nbr(data, var)
}

map(num_var_names_ss, map_clusters)
```
Neighborhoods with high house values are clustered in the north east. The southern area of low-low house values is the area with the highest crime and homelessness, euphemistically dubbed the "International District". 

The income map is similar, although the high-high areas in the east stretch down along the foothills of Sandia Mountain, an . There is also a surprisingly large patch of high-high in the northwest in the fifth district. The low-low income areas are mainly located along the I-25 corridor. 

As for inequality, high areas are clustered in the center, while low areas of inequality are clustered in the west. The university areas show areas of low inequality surrounded by areas of high inequality.

College percentage and Hispanic percentage are virtual mirror images geographically. The number of significant areas is much higher, and both show a significant NW-SE axis. The patch of low Hispanic and high percentage of college degrees in the center is the are around the University of New Mexico.






