---
title: "District Variance"
format: gfm
---

# Introduction

The object of this article is two-fold. The primary purpose is to further explore the differences between Albuquerque's council districts. In the [last article](https://biscotty.net/posts/data-science/r-census/abq-demographics), I looked at variation in  age, race, and educational attainment between districts Albuquerque. In some cases, the variation was striking. In particular, the third district seemed quite different from the others in a number of ways. Now I would like to apply some statistical testing to see if this impression of variance is backed up statistically. As in the prior article, I will use data from the US Census's American Community Survey. In addition to the demographic variables of education and race, I will add a number of economic indicators such as income levels and housing values.

My second purpose is aimed at programmers. In the last article I used functional programming techniques from `purrr` to write clear and structurally efficient code. In this article I will show how the `collapse` library can replace the `dplyr` library to produce computationally efficient code. `dplyr` is a wonderfully expressive library and a key part of the popular `tidyverse`. It is not, unfortunately, particularly fast. `collapse` can be orders of magnitude faster than `dplyr`, and even `data.table`. It provides fast versions of many `dplyr` functions, simply with a prepended "f", eg. `fmutate()`, as well as many other fast statistical functions and transformations. Throughout, I will show comparisons of some functions.

# Data Preparation

I will begin by loading the required libraries and setting some constants. In many cases, I only actually need one function from a given library. In the interest of memory efficiency, I will alias the individual commands rather than loading the entire library. I will load `dplyr` so that I can run comparison tests with `microbenchmark`.

```{r}
#| message: false
options(paged.print = FALSE,
        tigris_use_cache = TRUE)
# libraries <- list(
#   "corrr", "ggspatial", "ggtext", "patchwork", "sf",
#   "spatialreg", "spdep", "tidycensus", "tidyverse", 
#   "zeallot", "DescTools", "GWmodel", "scales", "glue",
#   "ggpubr", "units", "rstatix", "rcompanion", "collapse",
#   "classInt", "nngeo", "microbenchmark", "ggridges"
# )
libraries <- list(
  "sf", "collapse", "ggplot2", "magrittr", "dplyr",
  "units", "zeallot", "purrr", "patchwork"
)
invisible(lapply(libraries, library, character.only = TRUE))

microbenchmark <- microbenchmark::microbenchmark
glue <- glue::glue
annotation_map_tile <- ggspatial::annotation_map_tile
ggboxplot <- ggpubr::ggboxplot
ggqqplot <- ggpubr::ggqqplot
shapiro_test <- rstatix::shapiro_test
blom <- rcompanion::blom

year <- 2024
crs <- 6528
caption <- glue("Source: census.gov, acs5, {year}")
```

## Council Districts

Now I'll begin prepare the council district spatial dataset. I did this in the last article, but this time I'll use `collapse` functions. `fselect` works just like `select`. `collapse` does have an `fmutate` function which is the fast version of `mutate`. When no grouping is required, `ftransform` is even faster because it evaluates all arguments simultaneously. `roworder` replaces `arrange`, and `rowbind` replaces `rbind`. Finally, `qF` is used instead of `factor`.

```{r}
council_dists <- st_read("data/BC_CityCouncil/ABQ_CityCouncils.shp") %>% 
  fselect(district = DISTRICTNU) %>% 
  ftransform(district = paste("District", district)) %>% 
  roworder(district) %>% 
  nngeo::st_remove_holes() %>% 
  st_cast("MULTIPOLYGON")
los_ranchos <- st_read("data/LosRanchos/LosRanchos.shp") %>%
  fselect(district = Name)
council_dists <-
  rowbind(council_dists, los_ranchos) %>%
  st_transform(crs) %>% 
  ftransform(district = qF(district))
```

My study area is a rectangular region that is a bit smaller than the actual city limits, which will cut out some outlying areas. Also, as mentioned before, there are many patches of unincorporated county land lying within the rectangular area of Albuquerque. I will explicitly label these areas as "Unincorporated", obtaining the geometries by taking the `st_difference` of the rectangular bounding box of the `council_dists` and the geometries already defined in the `sf` object.

```{r}
#| warning: false
dist_box = c(xmin = 452000, xmax = 479833, 
             ymin = 443500, ymax = 467858)
council_dists %<>% 
  st_crop(dist_box) %>% 
  st_cast("MULTIPOLYGON")
difference_sfc <-
  st_difference(
    st_as_sfc(st_bbox(council_dists)),
    st_union(council_dists$geometry)
  )
difference_sf <- 
  st_sf(data.frame(district = "Unincorporated"),
  geometry = difference_sfc
)
council_dists <- rowbind(difference_sf, council_dists)
```


```{r}
#| warning: false
ggplot(council_dists, aes(color = district)) +
  geom_sf() +
  theme_void()
```


## Census Data

### Fetching

The `get_acs()` function obtains information from the American Community Survey tables. By supplying a named vector, the columns of the returned data table will be automatically renamed. Setting `geometry = TRUE` returns an `sf` object, and `output = "wide"` gives a column for each variable. After obtaining the data, I will remove the margin of error columns, remove the trailing "E" from the remaining column names, and transform the geometry to an appropriate coordinate reference system. `get_vars` is a more flexible version of `fselect`.

```{r}
vars_census <- c(
  House.Value.Med = "B25077_001",
  Renter.Occupied = "B25003_003",
  Vacant = "B25002_003",
  Income.Med = "DP03_0062",
  Gini = "B19083_001",
  Rent.Med = "B25031_001",
  Foreign.Born.Pct = "DP02_0094P",
  Hisp.Pct = "DP05_0090P",
  In.Poverty = "B17001A_002",
  Age.Med = "B01002_001",
  College.Pct = "DP02_0068P",
  Population = "B01003_001"
)

bern_data <- tidycensus::get_acs(
  geography = "tract",
  state = "NM",
  county = "Bernalillo",
  variables = vars_census,
  output = "wide",
  year = year,
  geometry = TRUE,
  cache_table = TRUE
) %>%
  get_vars(c("E$", "GEOID"), regex = TRUE) %>%
  get_vars(-2) %>%
  frename(\(x) sub("E$", "", x)) %>%
  st_transform(crs)
```

### Cleaning

Let's take a look at the data before splitting it between districts. I'll start with `collapse`'s version of `summary`.

```{r}
qsu(bern_data)
```

There is some clean-up to do. There are some `na`s to explore, and oddities like areas where the population total is 0. I'll deal with the zero-population areas first. `fsubset` is like `filter`. `collapse` even provides a fast versions of the logical operators `==`, `!=`, and `%in%`.

```{r}
bern_data %>% 
  fsubset(Population %==% 0)
```

It is worth comparing this with the `dyplr` equivalent.

```{r}
microbenchmark(
  dp = bern_data %>%
    filter(Population == 0),
  co = bern_data %>%
    fsubset(Population %==% 0)
)
```

The `collapse` version takes two percent of the time `dplyr` takes!

```{r}
bern_data_s <- fsubset(bern_data, Population %==% 0)
ggplot(bern_data) +
  geom_sf(color = "red") +
  geom_sf(data = bern_data_s, fill = "navy") +
  annotation_map_tile(
    type = "osm", alpha = 0.7,
    zoomin = -1, cachedir = "~/.cache/maps/"
  ) +
  theme_void() +
  lims(
    x = dist_box[1:2],
    y = dist_box[3:4]
  )
```


This tract is the state fairgrounds. The other is outside the study area. I'll remove them.

```{r}
bern_data %<>% fsubset(Population %!=% 0)
qsu(bern_data)
```

The population of 16 is suspicious. 

```{r}
bern_data_s <- fsubset(bern_data, Population < 20)
ggplot(bern_data) +
  geom_sf(color = "red") +
  geom_sf(
    data = bern_data_s,
    fill = "navy"
  ) +
  annotation_map_tile(
    type = "osm", alpha = 0.7,
    zoomin = -1, cachedir = "~/.cache/maps/"
  ) +
  theme_void() +
  lims(
    x = dist_box[1:2],
    y = dist_box[3:4]
  )
```

This tract contains federal land, and no useful data. I'll remove this as well.

```{r}
bern_data %<>% fsubset(Population > 20)
```

This only leaves `na`s in two variables. Let's take a closer look. As you might guess `missing_cases` is from `dyplr`, and is the fast (opposite) of `complete.cases`. `complete.cases`, however, cannot be used on an `sf` object without dropping the geometry. `missing_cases` is over 80 times faster than the `dplyr` idiom.

```{r}
microbenchmark(
  dp = bern_data %>%
    filter(if_any(everything(), is.na)),
  co = bern_data %>%
    fsubset(missing_cases(bern_data))
)
bern_data %>%
    fsubset(missing_cases(bern_data))
```

This data seems worth preserving. I will fill in these values based on the median values for each variable, but I would like these values to be based on the median for the specific district, rather than an overall median, so I will split the data into districts at this point.

I want to remember the rows with missing values so I can make sure they get filled in. I can't just use row indices, since these will change after splitting into districts. Instead I'll grab the `GEOID`s. I want them as a vector, and `dplyr` has a wonderful `pull` function to do this. `collapse` has no such function. The `mctl` function does something similar, but it requires a `matrix` and returns a list. So I will use `purrr::compose` to make my own `pull` function, which turns out to be quite a bit faster than `dplyr`'s.

I am also using shorthand verbs provided by `collapse`: `gv`, short for `get_vars`, and `ss`, short for `fsubset`.

```{r}
pull <- compose(unlist, mctl, qM)
microbenchmark(
  dpull = bern_data %>%
    st_drop_geometry %>%
    ss(missing_cases(bern_data)) %>%
    dplyr::pull(GEOID),
  mypull = bern_data %>%
    st_drop_geometry %>%
    ss(missing_cases(bern_data)) %>%
    gv("GEOID") %>%
    pull
)
missing <- bern_data %>%
  st_drop_geometry() %>%
  ss(missing_cases(bern_data)) %>%
  gv("GEOID") %>%
  pull
```


## Create District Dataset

Splitting the data involves intersecting the county data with the districts. Since some of the variables are extensive, representing counts, these values will need to be apportioned based on the percentage area represented by each subsection. I'll calculate the initial area of each tract, then the areas of each newly-created sub-section, allowing me to divide each tract's data by the percentage area lying in the new tracts.


```{r}
#| warning: false
abq_data <- bern_data %>%
  ftransform(area = st_area(.)) %>%
  st_intersection(council_dists) %>%
  ftransform(area_pct = as.numeric(st_area(.) / area)) %>% 
  fsubset(district %!iin% c("Los Ranchos", "Unincorporated"))

ext_vars <- c("Population", "Renter.Occupied", "Vacant", "In.Poverty")

abq_data %<>%
  ftransformv(
    ext_vars,
    FUN = function(x) round(x * abq_data$area_pct, 0)
  ) %>% 
  gv(-14) %>% 
  colorder(district, pos = "front")
```

`ftransformv` takes a vector of column identifiers, in this case column names, and applies a function to them, in this case an anonymous function, using the `FUN` argument. `colorder` is self-explanatory.

Before going further, I want to split the data from the geometries, so I can work with a simpler object. I'll preserve the geometries so I can recreate the `sf` object when I want to do mapping or spatial autocorrelation. I'll also create a `GRP` object for `district`. This variable is already a factor, but `GRP` objects are faster, and can be created from multiple variables.



```{r}
abq_data_geom <- st_geometry(abq_data)
abq_data <- st_drop_geometry(abq_data)
abq_dist_g <- GRP(abq_data, ~ district)
```

Now I can fill in the per-district averages for the two variables with NAs. `TRA` offers various options, such as replacing all values, centering and scaling. Here `dplyr` is over 20 times slower.


```{r}
microbenchmark(
  co = abq_data %<>%
    fgroup_by(district) %>%
    fmutate(
      House.Value.Med = fmedian(House.Value.Med, TRA = "replace_NA"),
      Rent.Med = fmedian(Rent.Med, TRA = "replace_NA")
    ) %>%
    fungroup(),
  dp = abq_data %>%
    group_by(district) %>%
    mutate(
      House.Value.Med = if_else(is.na(House.Value.Med),
        median(House.Value.Med, na.rm = T),
        House.Value.Med,
      ),
      Rent.Med = if_else(is.na(Rent.Med),
        median(Rent.Med, na.rm = T),
        Rent.Med,
      )
    ) %>%
    ungroup()
)
```


```{r}
abq_data %<>%
  fgroup_by(district) %>%
  fmutate(
    House.Value.Med = fmedian(House.Value.Med, TRA = "replace_NA"),
    Rent.Med = fmedian(Rent.Med, TRA = "replace_NA")
  ) %>%
  fungroup
```


All missing values have been filled in. However, there is a new problem:

```{r}
qsu(abq_data)
```

The split data now contains more areas with a population of 0. As a result of the splitting, there are numerous small fragments containing little information. To address this, I will remove all of the new rows which represent less than 1\% of the original area.

```{r}
abq_data %<>% fsubset(area_pct > .01)
qsu(abq_data)
```


That looks good. I'll remove unnecessary columns and can proceed with exploring the data distributions.

```{r}
abq_data %<>% get_vars(-c(2,15))
```

# Data Normalization

To perform the standard analysis of variance, data is expected to be normally distributed. Further, there should be equality of variance across districts. 

## Visualize

I'll start by plotting the raw data. I want to construct nine histograms and nine qqplots, and display them side-by-side with `patchwork`. This is where I really appreciate the functional programming aspects of __R__. To do this, I'll first create a histogram function. Note the `.data[[var]]` syntax. It is necessary to pass variables to `aes()` when using the `aes()` function within a function. Then I will use `map()` four times to produce the plots.

```{r}
num_var_names <- num_vars(abq_data, return = "names")

plot_dist <- function(df, var) {
  df %>%
    ggboxplot("district", var, fill = "district") +
    geom_point(position = position_jitter(width = 0.1)) +
    theme(axis.text.x = element_text(angle = 45, vjust = .5)) +
    guides(fill = "none") +
    labs(caption = caption)
}

c(p1, p2, p3, p4, p5, p6, p7, p8, p9) %<-%
  map(num_var_names, \(x) abq_data %>% plot_dist(x))

(p1 + p2)
(p3 + p4)
(p5 + p6)
(p7 + p8)
p9
```

```{r}
plot_hist <- function(df, var) {
  df %>%
    ggplot(aes(.data[[var]])) +
    geom_histogram(aes(y = after_stat(density)),
      fill = "lightgray", col = "black", bins = 20
    ) +
    geom_line(stat = "density", adjust = 2, color = "blue") +
    labs(caption = caption)
}

hists <- map(num_var_names, \(x) abq_data %>% plot_hist(x))

vars_list <- get_vars(abq_data, num_var_names) %>% as.list()

qqs <- map(vars_list, ggqqplot)

map(
  1:length(hists),
  \(x) hists[[x]] + qqs[[x]]
)
```

With a couple of exceptions, none of these variables appear to be normally distributed.


## Shapiro tests

The Shapiro-Wilk test can be used to evaluate statistically whether a variable is normally distributed. The null hypothesis in this test is that the data is normally distributed. $p$-values less than 0.05 indicate that the null hypothesis should be rejected, implying that the data are not normally distributed. I'll start by running a `shapiro_test` on all of the variables, filtering for $p$-values greater than 0.05. These variables are normally distributed.

```{r}
map(vars_list, shapiro_test) %>% 
  unlist2d %>% 
  fmutate(
    variable = num_var_names,
    sig = ifelse(p.value <= .05, "Yes", "No")
  ) %>%
  fselect(-1) %>%
  fsubset(sig == "No")
```

## Transformations

I'd like to test a number of different transformations, so I will create a function to do so and filter the data as above. Happily, in R, functions are first-class data types, allowing functions to be passed as arguments to functions. In fact, this custom function highlights the richness of R's functional programming capabilities, using the built-in `lapply()` and `purrr`'s `map()`. The pipe itself is akin to functional composition.


```{r}
run_shapiro_abq <- function(f) {
  data <- gv(abq_data, num_var_names) %>% as.list() %>% lapply(f)
  
  map(data, shapiro_test) %>%
    unlist2d %>%
    fmutate(
      variable = num_var_names,
      sig = ifelse(p.value <= .05, "Yes", "No")
    ) %>%
    fselect(-1) %>%
    fsubset(sig == "No")
}
```


I'll start with a simple square-root transformation.

```{r}
run_shapiro_abq(sqrt)
```

That doesn't help. How about a cube root?

```{r}
cube_root <- function(x) {
  sign(x) * abs(x)^(1/3)
}

run_shapiro(abq_data, num_var_names, cube_root)
```

That helps a little, but not much. I can't use a straight-forward log transformation due to the many 0 values, so I will use a modified log transformation.

```{r}
ihs_transform <- function(x) {
  log(x + (x^2 + 1)^0.5)
}

run_shapiro(abq_data, num_var_names, ihs_transform)
```

```{r}
map(list(sqrt, cube_root, ihs_transform), run_shapiro_abq)
```

# Here

That's even worse. I'll try an inverse normal transformation. This transformation is based on ranking, and is able to normalize highly skewed distributions. The downside of this transformation is that the "distance" between each observation is lost, obscuring information about the variability within the variable. The `blom()` function which performs the transformation defaults to a "general" method. 

```{r}
xf_blom <- partial(blom, method = "blom")
run_shapiro_abq(xf_blom)
```

Remove non-standard variables.

```{r}
abq_data %<>% fselect(-Vacant)
num_var_names <- num_var_names[-3]
```


```{r}
abq_data_blom <- abq_data %>% 
  fmutate(across(num_var_names, function(x) xf_blom(x)))
```

```{r}
dmutate <- dplyr::mutate
dacross <- dplyr::across
dall_of <- dplyr::all_of
microbenchmark::microbenchmark(
  collapse = abq_data %>%
    fmutate(across(num_var_names, function(x) xf_blom(x))),
  dyplr = abq_data %>%
    dmutate(dacross(dall_of(num_var_names), ~ xf_blom(.x)))
)
```

## Distribution Plots

```{r}
c(p1, p2, p3, p4, p5, p6, p7, p8, p9) %<-%
  map(var_names, \(x) abq_data %>% plot_dist(x))

(p1 + p2)
(p3 + p4)
(p5 + p6)
(p7 + p8)
p9
```

```{r}
var_vals <- get_vars(abq_data_elf, var_names) %>% as.list()

hists <- map(var_names, \(x) abq_data_elf %>% plot_hist(x))
lines <- map(var_vals, ggpubr::ggqqplot)
map(1:10,
     \(x) hists[[x]] + lines[[x]])
```

## Equality of variance (Bartlett's test)


```{r}
# var_vals <- map(var_names, \(x) pull(abq_data_elf, x))
btests <- map2(var_names, 
     var_vals,
    \(x, y) list(x, bartlett.test(y ~ district, abq_data_elf)))
data.frame(
  variable = unlist(map(1:length(btests), \(x) btests[[x]][[1]])),
  statistic = unlist(map(1:length(btests), \(x) btests[[x]][[2]]$statistic)),
  p_value = unlist(map(1:length(btests), \(x) btests[[x]][[2]]$p.value %>% round(., 3)))
)
```

```{r}
var_names_aov <- var_names[2:7]
var_vals_aov <- get_vars(abq_data_elf, var_names_aov) %>% as.list()

btests <- map2(var_names_aov, 
     var_vals_aov,
    \(x, y) list(x, bartlett.test(y ~ district, abq_data_elf)))
data.frame(
  variable = unlist(map(1:length(btests), \(x) btests[[x]][[1]])),
  statistic = unlist(map(1:length(btests), \(x) btests[[x]][[2]]$statistic)),
  p_value = unlist(map(1:length(btests), \(x) btests[[x]][[2]]$p.value %>% round(., 3)))
)
```

# Variance Between Council Districts

## ANOVA

```{r}
aov_tests <- map(
  var_names_aov,
  \(x) list(x, aov(as.formula(glue("{x} ~ district")),
    data = abq_data_elf)
  )
)
map2(1:length(aov_tests), var_names_aov, 
     \(x, y) list(y, summary(aov_tests[[x]][[2]])))

```

## Welch's tests

```{r}
var_names_welch <- var_names[c(1, 8:10)]
map(
  var_names_welch,
  \(x) list(x, oneway.test(as.formula(glue("{x} ~ district")),
    data = abq_data_elf, var.equal = F)
  )
)
```



```{r}
map(
  var_names_welch,
  \(x) list(x, rstatix::welch_anova_test(as.formula(glue("{x} ~ district")),
    data = abq_data_elf %>% st_drop_geometry)
  )
)
```

## Ridge Plots

```{r}
#| message: false
#| warning: false
plot_dist_ridge <- function(df, var) {
  ggplot(df, aes(
    x = .data[[var]], y = district,
    fill = factor(stat(quantile))
  )) +
    ggridges::stat_density_ridges(
      geom = "density_ridges_gradient", calc_ecdf = TRUE,
      quantiles = 4, quantile_lines = TRUE, alpha = 0.7,
      jittered_points = TRUE, position = "points_sina"
    ) +
    scale_fill_viridis_d(name = "Quartiles") +
    theme(axis.title.x = element_blank())
}
```

```{r}
#| warning: false
map(var_names, 
    \(x) abq_data_elf %>% 
      plot_dist_ridge(x) +
      labs(title = x))
```

## Tukey's

```{r}
results_tky <- map(
  aov_tests,
  \(x) TukeyHSD(x[[2]])$district %>% 
    as.data.frame %>%
  fsubset(`p adj` <= .05) %>% 
  fmutate(variable = x[[1]], .before = 1) %>% 
    rownames_to_column("districts")) %>% 
  unlist2d(idcols = FALSE) %>% 
  separate(districts, c("District.1", "District.2"), sep = "-")

map(
  aov_tests,
  \(x) TukeyHSD(x[[2]])$district %>%
    as.data.frame() %>%
    fsubset(`p adj` <= .05) %>%
    fmutate(variable = x[[1]], .before = 1)
) %>%
  unlist2d(idcols = FALSE) %>%
  ftransform(tstrsplit(districts, split = "-", names = c("District.1", "District.2"))) %>% 
  get_vars(-1) %>% 
  


library(data.table)

  separate(districts, c("District.1", "District.2"), sep = "-")

ftransform(data.table::tstrsplit(districts, split = "-", names = c("District.1", "District.2")))

dist_names <- unique(c(results_tky$District.1, results_tky$District.2)) %>% 
  sort()
var_names_tky <- unique(results_tky$variable)

results_tky_dist <- map(
  dist_names,
  \(x) results_tky %>%
    filter(District.1 == x | District.2 == x) %>%
    mutate(
      district = x,
      compare = if_else(District.1 == x, District.2, District.1),
      .before = variable
    ) %>% 
    select(-(1:2)) %>% 
    arrange(district, compare)
)
```

```{r}
map(
  dist_names,
  \(x) results_tky %>%
    filter(District.1 == x | District.2 == x) %>%
    mutate(
      district = x,
      compare = if_else(District.1 == x, District.2, District.1),
      .before = variable
    ) %>% 
    select(-(1:2)) %>% 
    arrange(district, compare) %>% 
    as_tibble()
)
```

```{r}
map(unique(results_tky$variable),
    \(x) results_tky %>% 
      filter(variable == x) %>% 
  relocate(variable, .before = 1) %>% 
  as_tibble())
```

## Linear Model

```{r}

map(var_names,
    \(x) list(x, lm(as.formula(glue("{x} ~ district")),
      data = abq_data_elf
    ) %>% summary()))
```

```{r}
abq_data %>% 
  st_drop_geometry() %>% 
  group_by(district) %>% 
  summarise(Hisp.Pct = mean(Hisp.Pct))
```


```{r}
var_names_str <- glue_collapse(var_names, sep = " + ")
abq_data_elf <- abq_data_elf %>% 
  mutate(
    isDist1 = if_else(district == "District 1", 1, 0),
    isDist2 = if_else(district == "District 2", 1, 0),
    isDist3 = if_else(district == "District 3", 1, 0),
    isDist4 = if_else(district == "District 4", 1, 0),
    isDist5 = if_else(district == "District 5", 1, 0),
    isDist6 = if_else(district == "District 6", 1, 0),
    isDist7 = if_else(district == "District 7", 1, 0),
    isDist8 = if_else(district == "District 8", 1, 0),
    isDist9 = if_else(district == "District 9", 1, 0),
  )
dist_vars <- c("isDist1", "isDist2", "isDist3", 
               "isDist4", "isDist5", "isDist6", 
               "isDist7", "isDist8", "isDist9")
forms <- map(
  dist_vars,
  \(x) as.formula(glue("{x} ~ {var_names_str}"))
)
logit_models <- map(
  forms,
  \(x) list(x,
            glm(x, data = abq_data_elf, family = binomial(link = "logit")))
)
logit_models[[1]]
map(logit_models, \(x) list(x[[1]], summary(x[[2]])))
form = as.formula(glue("isDist1 ~ {var_names_str}"))
logit_model <- glm(form,
                   data = abq_data_elf, family = binomial(link = "logit"))
summary(logit_model)

abq_data_elf <- abq_data_elf %>% 
  mutate(dist_num = parse_number(district))
abq_data <- abq_data %>% 
  mutate(dist_num = parse_number(district))
form <- as.formula(glue("dist_num ~ {var_names_str}"))
lm(form,
   data = abq_data %>% st_drop_geometry()) %>% 
  summary()
```

## Spatial autocorrelation

```{r}
abq_data_nona <- abq_data %>% 
  mutate(vbl = Income.Med) %>% 
  filter(!is.na(vbl))
nb <- poly2nb(abq_data_nona, queen = F, snap = 25)
nbw <- nb2listw(nb, style = "W", zero.policy = T)

gmoran <- moran.test(abq_data_nona$vbl, nbw, alternative = "two.sided")
gmoran
```

```{r}
moran.plot(abq_data_nona$vbl, nbw)
```

```{r}
lmoran <- localmoran(abq_data_nona$vbl, nbw, alternative = "greater")
head(lmoran)
```

```{r}
abq_data_nona$lmI <- lmoran[, "Ii"] # local Moran's I
abq_data_nona$lmZ <- lmoran[, "Z.Ii"] # z-scores
abq_data_nona$lmp <- lmoran[, "Pr(z > E(Ii))"]
```

```{r}
make_quantiles <- function(var) {
  classIntervals(var, 5, style = "quantile")
}

quantile_breaks <- make_quantiles(abq_data_nona$Income.Med)
br_quantile <- classIntervals(
  abq_data_nona$Income.Med, 5,
  style = "quantile"
)
abq_data_nona$Income.Med_q <- cut(
  abq_data_nona$Income.Med, br_quantile$brks,
  include.lowest = TRUE, dig.lab = 10
)
pal <- colorRampPalette(c("darkred", "darkgreen", "navy"))(length(br_quantile$brks))
ggplot(abq_data_nona) +
  geom_sf(aes(fill = Income.Med_q)) +
  scale_fill_viridis_d() +
  labs(fill = "Median Income") +
  theme_void()
```


```{r}
library(tmap)

p1 <- tm_shape(abq_data_nona) +
  tm_polygons(col = "vbl", title = "vbl", style = "quantile") +
  tm_layout(legend.outside = TRUE)

p2 <- tm_shape(abq_data_nona) +
  tm_polygons(col = "lmI", title = "Local Moran's I",
              style = "quantile") +
  tm_layout(legend.outside = TRUE)

p3 <- tm_shape(abq_data_nona) +
  tm_polygons(col = "lmZ", title = "Z-score",
              breaks = c(-Inf, 1.65, Inf)) +
  tm_layout(legend.outside = TRUE)

p4 <- tm_shape(abq_data_nona) +
  tm_polygons(col = "lmp", title = "p-value",
              breaks = c(-Inf, 0.05, Inf)) +
  tm_layout(legend.outside = TRUE)

tmap_arrange(p1, p2, p3, p4)
```

```{r}
tmap5 <- tm_shape(abq_data_nona) + 
  tm_polygons(col = "lmZ",
              title = "Local Moran's I", 
              style = "fixed",
              breaks = c(-Inf, -1.96, 1.96, Inf),
              labels = c("Negative SAC", "No SAC", "Positive SAC"),
              palette =  c("blue", "white", "red")) +
  tm_layout(legend.outside = TRUE)
tmap5
```

## Clusters

```{r}
lmoran <- localmoran(abq_data_nona$vbl, nbw, alternative = "two.sided")
head(lmoran)
```

```{r}
abq_data_nona$lmp <- lmoran[, 5]
mp <- moran.plot(as.vector(scale(abq_data_nona$vbl)), nbw)
```

```{r}
abq_data_nona$quadrant <- NA
abq_data_nona[(mp$x >= 0 & mp$wx >= 0) & (abq_data_nona$lmp <= 0.05), "quadrant"] <- 1
abq_data_nona[(mp$x <= 0 & mp$wx <= 0) & (abq_data_nona$lmp <= 0.05), "quadrant"] <- 2
abq_data_nona[(mp$x >= 0 & mp$wx <= 0) & (abq_data_nona$lmp <= 0.05), "quadrant"] <- 3
abq_data_nona[(mp$x <= 0 & mp$wx >= 0) & (abq_data_nona$lmp <= 0.05), "quadrant"] <- 4
abq_data_nona[(abq_data_nona$lmp > 0.05), "quadrant"] <- 5
```

```{r}
tm_shape(abq_data_nona) +
  tm_fill(
    col = "quadrant", title = "",
    breaks = c(1, 2, 3, 4, 5, 6),
    palette = c("red", "blue", "lightpink", "skyblue2", "white"),
    labels = c(
      "High-High", "Low-Low", "High-Low",
      "Low-High", "Non-significant"
    )
  ) +
  tm_legend(text.size = 1) + tm_borders(alpha = 0.5) +
  tm_layout(frame = FALSE, title = "Clusters") +
  tm_layout(legend.outside = TRUE)
```


# Save data

```{r}
data_sets <- c(
  "dist_data", "bern_data", "abq_data", "abq_data_elf"
)

walk2(
  map(data_sets, \(x) get(x)),
  data_sets,
  \(df, level) st_write(df, "data/district_variance.gpkg",
    layer = level, append = F
  )
)
```






