---
title: "District Variance"
format: gfm
---

# Introduction

The object of this article is two-fold. The primary purpose is to further explore the differences between Albuquerque's council districts. In the [last article](https://biscotty.net/posts/data-science/r-census/abq-demographics), I looked at variation in age, race, and educational attainment between districts. In some cases, the variation was striking. In particular, the third district seemed quite different from the others in a number of ways. Now I would like to apply some statistical testing to see if this impression of variance is backed up statistically. As in the prior article, I will use data from the US Census's American Community Survey. In addition to the demographic variables of education and race, I will add a number of economic indicators such as income levels and housing values.

My secondary purpose is to show how the `collapse` package can be used in place of `dplyr` to significantly speed up data processing. `collapse`'s syntax has a familiar feel to the `tidyverse`'s, but is in many cases orders of maginitude faster at statistical operations and transformations. In the last article I used `purrr` to achieve syntactical efficiency and robustness. In this article I will show how the `collapse` library can computationally efficient code. `dplyr` is a wonderfully expressive library and a key part of the popular `tidyverse`. It is not, unfortunately, particularly fast. `collapse` can be orders of magnitude faster than `dplyr`, and even `data.table`. It even provides fast versions of many `dplyr` functions, simply with a prepended "f", eg. `fmutate()`, as well as many other fast statistical functions and transformations. Throughout, I will show comparisons of some functions using the `microbenchmark` package.

# Loading libraries

I will begin by loading the required libraries and setting some constants. In many cases, I only actually need one function from a given library. Some of these packages contain many functions. In the interest of memory efficiency, I will alias the individual commands rather than loading the entire library. I will load `dplyr` only so that I can run comparison tests with `microbenchmark`. It will not be used in the actual project code.

```{r}
#| message: false
options(paged.print = FALSE,
        tigris_use_cache = TRUE)
libraries <- list(
  "sf", "collapse", "ggplot2", "magrittr", "dplyr",
  "zeallot", "purrr", "patchwork", "spdep"
)
invisible(lapply(libraries, library, character.only = TRUE))

microbenchmark <- microbenchmark::microbenchmark
glue <- glue::glue
annotation_map_tile <- ggspatial::annotation_map_tile
ggboxplot <- ggpubr::ggboxplot
ggqqplot <- ggpubr::ggqqplot
shapiro_test <- rstatix::shapiro_test
blom <- rcompanion::blom
tidy <- broom::tidy
welch_anova_test <- rstatix::welch_anova_test
stat_density_ridges <- ggridges::stat_density_ridges
rownames_to_column <- tibble::rownames_to_column
separate <- tidyr::separate
games_howell_test <- rstatix::games_howell_test
rescale <- scales::rescale
classIntervals <- classInt::classIntervals
str_replace_all <- stringr::str_replace_all
set_units <- units::set_units
iif <- kit::iif
st_remove_holes <- nngeo::st_remove_holes

year <- 2024
crs <- 6528
caption <- glue("Source: census.gov, acs5, {year}")
```

# Preparing the data

## Council Districts

Now I'll prepare the council district spatial dataset. I did the same in the last article, but this time I'll use `collapse` functions. `fselect` works just like `select`. `collapse` does have an `fmutate` function which is the fast version of `mutate`. When no grouping is required, `ftransform` is even faster because it evaluates all arguments simultaneously. `roworder` replaces `arrange`, and `rowbind` replaces `rbind`. Finally, `qF` is used instead of `factor`.

```{r}
council_dists <- st_read("data/BC_CityCouncil/ABQ_CityCouncils.shp") %>% 
  fselect(district = DISTRICTNU) %>% 
  ftransform(district = paste("District", district)) %>%
  roworder(district) %>% 
  st_remove_holes %>% 
  st_cast("MULTIPOLYGON")
los_ranchos <- st_read("data/LosRanchos/LosRanchos.shp") %>%
  fselect(district = Name)
council_dists <-
  rowbind(council_dists, los_ranchos) %>%
  st_transform(crs) %>% 
  ftransform(district = qF(district))
```

My study area is a rectangular region that is a bit smaller than the actual city limits, which will cut out some outlying areas. Also, as mentioned before, there are many patches of unincorporated county land lying within the rectangular area of Albuquerque. Above I used `st_remove_holes` to include enclaves of unincorporated county in the respective districts. For the rest, I will explicitly label these areas as "Unincorporated", obtaining the geometries by taking the `st_difference` of the rectangular bounding box of the `council_dists` and the geometries already defined in the `sf` object.

```{r}
#| warning: false
dist_box = c(xmin = 452000, xmax = 479833, 
             ymin = 443500, ymax = 467858)
council_dists %<>% 
  st_crop(dist_box) %>% 
  st_cast("MULTIPOLYGON")

difference_sfc <-
  st_difference(
    st_as_sfc(st_bbox(council_dists)),
    st_union(council_dists$geometry)
  )
difference_sf <- 
  st_sf(data.frame(district = "Unincorporated"),
  geometry = difference_sfc
)
council_dists <- rowbind(difference_sf, council_dists)
```

I have used `%<>%`, a lesser-known function from `magrittr`. It replaces the tedious yet ubiquitous idiom `data <- data %>%` with the simpler and shorter `data %<>%`. Also, note that after many `sf` operations, it is necessary to ensure consistent geometries. Without recasting the cropped council districts, an error would ensue.

```{r}
#| warning: false
ggplot(council_dists) +
  annotation_map_tile(
    type = "osm", zoomin = -1, cachedir = "~/.cache/maps/"
  ) +
  geom_sf(aes(color = district), linewidth = 1) +
  theme_void()

dist_plot <- ggplot(council_dists) +
  annotation_map_tile(
    type = "osm", zoomin = -1, cachedir = "~/.cache/maps/"
  ) +
  geom_sf(aes(color = district), linewidth = 1, show.legend = F) +
  theme_void()
```

## Census Data

### Fetching

The `get_acs()` function obtains information from the American Community Survey tables published by the US Census Department. By supplying a named vector to the `variables` argument, the columns of the returned data table will be automatically renamed. Setting `geometry = TRUE` returns an `sf` object, and `output = "wide"` gives a column for each variable. After obtaining the data, I will remove the margin of error columns, remove the trailing "E" from the remaining column names, and transform the geometry to an appropriate coordinate reference system. `get_vars` is a more flexible version of `fselect`.

```{r}
vars_census <- c(
  House.Value.Med = "B25077_001",
  Housing.Units = "B25001_001",
  Renter.Occupied = "B25003_003",
  Vacant = "B25002_003",
  Income.Med = "DP03_0062",
  Gini = "B19083_001",
  Rent.Med = "B25031_001",
  Foreign.Born.Pct = "DP02_0094P",
  Hisp.Pct = "DP05_0090P",
  In.Poverty = "B17001A_002",
  Age.Med = "B01002_001",
  College.Pct = "DP02_0068P",
  Population = "B01003_001"
)

bern_data <- tidycensus::get_acs(
  geography = "tract",
  state = "NM",
  county = "Bernalillo",
  variables = vars_census,
  output = "wide",
  year = year,
  geometry = TRUE,
  cache_table = TRUE
) %>%
  get_vars(c("E$", "GEOID"), regex = TRUE) %>%
  get_vars(-2) %>%
  frename(\(x) sub("E$", "", x)) %>%
  st_transform(crs)
```

### Cleaning

Let's take a look at the data before splitting it between districts. I'll use `qsu`, `collapse`'s version of `summary`.

```{r}
qsu(bern_data)
```

There is some clean-up to do. There are some `na`s to explore, and oddities like areas where the population total is 0. I'll deal with the zero-population areas first. `fsubset` is like `filter`. `collapse` even provides a fast versions of the logical operators `==`, `!=`, and `%in%`.

```{r}
bern_data %>% 
  fsubset(Population %==% 0)
```

It is worth comparing this with the `dyplr` equivalent.

```{r}
microbenchmark(
  dp = bern_data %>% filter(Population == 0),
  co = bern_data %>% fsubset(Population %==% 0)
)
```

The `collapse` version is almost 100 times faster than `dplyr`'s!

```{r}
dist_plot +
  geom_sf(data = bern_data, color = "red") +
  geom_sf(data = fsubset(bern_data, Population %==% 0),
          fill = "navy") +
  lims(x = dist_box[1:2], y = dist_box[3:4])  
```

This tract is the state fairgrounds. The other is outside the study area. I'll remove both.

```{r}
bern_data %<>% fsubset(Population %!=% 0)
qsu(bern_data)
```

The population of 16 is suspicious. 

```{r}
dist_plot +
  geom_sf(data = bern_data, color = "red") +
  geom_sf(
    data = fsubset(bern_data, Population < 20),
    fill = "navy"
  ) +
  lims(x = dist_box[1:2], y = dist_box[3:4])
```

This tract contains federal land, and no useful data. I'll remove this as well.

```{r}
bern_data %<>% fsubset(Population > 20)
```

This only leaves `na`s. Let's take a closer look. As you might guess `missing_cases` is from `collapse`, and is the fast (opposite) of `complete.cases`. `complete.cases`, however, cannot be used on an `sf` object without dropping the geometry, while `missing_cases` can, and it is nearly 100 times faster than the `dplyr` idiom. (One important aspect of `collapse` is that it preserves attributes, although this is not entirely seamless for `sf` objects.)

```{r}
microbenchmark(
  dp = bern_data %>%
    filter(if_any(everything(), is.na)),
  co = bern_data %>%
    fsubset(missing_cases(bern_data))
)
```

```{r}
bern_data %>%
    fsubset(missing_cases(bern_data))
```

This data seems worth preserving. I will fill in these values based on the median values for each variable, mean in the case of Gini, but I would like these values to be based on the averages for the specific district, rather than an overall averages, so I will go ahead and split the data into districts at this point.

I want to remember the rows with missing values so I can check that they get filled in. I can't just use row indices, since these will change after splitting into districts. Instead I'll grab the `GEOID`s. I want them as a vector, and `dplyr` has a wonderful `pull` function to do this. `collapse` has no such function. The `mctl` function does something similar, but it requires a `matrix` and returns a list. So I will use `purrr::compose` to make my own `pull` function.

I am also using shorthand verbs provided by `collapse`: `gv`, short for `get_vars`, and `ss`, short for `fsubset`.

```{r}
# pull <- compose(unlist, mctl, qM)
missing <- (bern_data %>%
  st_drop_geometry() %>%
  ss(missing_cases(bern_data)))[["GEOID"]]
```


## Create the Albuquerque District Dataset

Splitting the data involves intersecting the county data with the districts. Since some of the variables are extensive, representing counts, these values will need to be apportioned based on the percentage area represented by each subsection. I'll calculate the initial area of each tract, then the areas of each newly-created sub-section, allowing me to divide each tract's data by the percentage area lying in the new tracts.


```{r}
#| warning: false
abq_data <- bern_data %>%
  ftransform(area = st_area(.)) %>%
  st_intersection(council_dists) %>%
  ftransform(area_pct = as.numeric(st_area(.) / area)) %>% 
  fsubset(district %!iin% c("Los Ranchos", "Unincorporated"))

ext_vars <- c("Population", "Housing.Units", "Renter.Occupied", "Vacant", "In.Poverty")

abq_data %<>%
  ftransformv(
    ext_vars,
    FUN = function(x) round(x * abq_data$area_pct, 0)
  ) %>%
  ftransform(
    In.Poverty.Pct = In.Poverty %c/% Population * 100,
    Population.Density = Population %c/% area %>%
      set_units("1/km^2") %>% as.numeric(),
    Vacant.Pct = Vacant %c/% Housing.Units * 100,
    Renter.Occupied.Pct = Renter.Occupied %c/% Housing.Units * 100,
    Housing.Density = Housing.Units %c/% area %>%
      set_units("1/km^2") %>% as.numeric(),
    Gini = Gini * 100,
    In.Poverty = NULL, Vacant = NULL, 
    Renter.Occupied = NULL, area = NULL
  ) %>%
  colorder(district, pos = "front") %>% 
  st_cast("MULTIPOLYGON")
```


```{r}
qsu(abq_data)
```


`ftransformv` takes a vector of column identifiers, in this case column names, and applies a function to them, in this case an anonymous function, using the `FUN` argument. `%c/%` is shorthand for column division. `collapse` provides a range of such functions for rows and columns, eg. `%c*%`, `%r+%`, etc.

Now I can fill in the per-district averages for the variables with NAs. `TRA` offers various options, such as replacing all values, centering and scaling. Here `collapse` is, again, nearly 100 times faster.

Now I want to fill in the missing `NA`s with grouped averages. I'll start by creating a grouping object. I could simply provide a column, but it is somewhat faster to do it this way. This is similar, to the `ftransformv` above, but the syntax is a little more complicated due to grouping. Many of the `collapse` functions have variants which modify the data by reference. These are preceded by `set`.


```{r}
dist_grp <- GRP(abq_data$district)

settransform(
  abq_data,
  fmean(list(
    In.Poverty.Pct = In.Poverty.Pct, Vacant.Pct = Vacant.Pct,
    Renter.Occupied.Pct = Renter.Occupied.Pct),
    g = dist_grp, TRA = "replace_NA"))

settransform(
  abq_data,
  fmedian(list(
      House.Value.Med = House.Value.Med, 
      Rent.Med = Rent.Med, Income.Med = Income.Med),
    g = dist_grp, TRA = "replace_NA"))

abq_data %>%
  fsubset(GEOID %iin% missing)
```

All missing values have been filled in. However, there is a new problem:

```{r}
qsu(abq_data)
```

The split data now contains new areas with a population of 0. As a result of the splitting, there are numerous small fragments containing little information. To address this, I will remove all of the new rows which represent less than 1\% of the original area. I'll also remove unnecessary columns, and make sure that the geometries are consistent.

```{r}
abq_data %<>% 
  fsubset(area_pct > .01) %>% 
  ftransform(GEOID = NULL, Population = NULL, area_pct = NULL)
qsu(abq_data)
```

I'm curious where there are such low-density housing.

```{r}
dist_plot +
  geom_sf(data = abq_data, color = "red") +
  geom_sf(data = abq_data %>% fsubset(Housing.Density < 50),
          fill = "navy") 
```



Before going further, I want to split the data from the geometries, so I can work with a simpler object. I'll preserve the geometries so I can recreate the `sf` object when I want to do mapping or spatial autocorrelation. `st_drop_geometry` returns a `tibble`. `tibble`s, while convenient for many things, carry extra overhead. I will use the fast `qDF` to convert to a simple data frame.

```{r}
abq_data_geom <- st_geometry(abq_data)
abq_data <- st_drop_geometry(abq_data) %>% qDF
```

## Visualizing the data

I'll start by plotting the raw data. I want to construct boxplots, histograms and qq-plots for all the variables, and display them side-by-side with `patchwork`. This is where I really appreciate the functional programming aspects of __R__. I'll first create a histogram function. Note the `.data[[var]]` syntax, which is necessary to pass variables to `aes()` when using `aes()` _within_ a function. Then I will use `map()` multiple times to produce the plots with `zeallot`'s assignment operator `%<-%`.

`num_vars` is a useful function which efficiently selects the numeric variables from a data frame, or any similar object. The `return` argument makes it very flexible. In this case, rather than the data, I'm just grabbing the variable names.


```{r}
num_var_names <- num_vars(abq_data, return = "names")

plot_dist <- function(df, var) {
  df %>%
    ggboxplot("district", var, fill = "district") +
    geom_point(position = position_jitter(width = 0.1)) +
    theme(axis.text.x = element_text(angle = 45, vjust = .5)) +
    guides(fill = "none") +
    labs(caption = caption)
}

c(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, p13) %<-%
  map(num_var_names, \(x) abq_data %>% plot_dist(x))

(p1 + p2)
(p3 + p4)
(p5 + p6)
(p7 + p8)
(p9 + p10)
(p11 + p12)
p13
```

```{r}
plot_hist <- function(df, var) {
  df %>%
    ggplot(aes(.data[[var]])) +
    geom_histogram(aes(y = after_stat(density)),
      fill = "lightgray", col = "black", bins = 20
    ) +
    geom_line(stat = "density", adjust = 2, color = "blue") +
    labs(caption = caption)
}

hists <- map(num_var_names, \(x) abq_data %>% plot_hist(x))

vars_list <- get_vars(abq_data, num_var_names) %>% as.list

qqs <- map(vars_list, ggqqplot)

map(
  1:length(hists),
  \(x) hists[[x]] + qqs[[x]]
)
```

# Analysis of Variance

To perform the standard analysis of variance, data is expected to be normally distributed. Further, there should be equality of variance across districts. As the above plots show, only a couple of the variables appear to be normally distributed.

## Data Normalization

With a couple of exceptions, visualization showed that none of these variables appear to be normally distributed. Formally, we can use the Shapiro-Wilk test for normal distribution to be sure.

The Shapiro-Wilk test can be used to evaluate statistically whether a variable is normally distributed. The null hypothesis in this test is that the data _is_ normally distributed. $p$-values less than 0.05 indicate that the null hypothesis should be rejected, implying that the data are not normally distributed. I'll start by running a `shapiro_test` on all of the variables, filtering for $p$-values greater than 0.05, which are the normally distributed variables.

The `unlist2d` function collapses a list of data frames to a single data frame.

```{r}
map(vars_list, shapiro_test) %>% 
  unlist2d %>% 
  fmutate(
    variable = num_var_names,
    sig = iif(p.value <= .05, "Yes", "No")
  ) %>%
  fsubset(sig == "No") %>%
  fselect(-1)
```

As expected, only two of the variables are normally distributed.

### Transformations

I will test some different transformations to see if I can coerce the data to normalcy. I will create a function to do so and filter the data as above. Happily, in R, functions are first-class data types, allowing functions to be passed as arguments to functions. In fact, this custom function highlights the richness of R's functional programming capabilities, using the built-in `lapply()` and `purrr`'s `map()`. I will write a generic function, and partially apply the first two arguments with `partial` to create a new function specific to my data.


```{r}
run_shapiro <- function(df, vars, f) {
  data <- gv(df, vars) %>% as.list() %>% lapply(f)
  map(data, shapiro_test) %>%
    unlist2d %>%
    fmutate(
      variable = vars,
      sig = iif(p.value <= .05, "Yes", "No")
    ) %>%
    fselect(-1) %>%
    fsubset(sig == "No")
}

run_shapiro_abq <- partial(run_shapiro, abq_data, num_var_names)
```


I'll try a few common transformations such as square root, cube root, and log transformations. I'll use a modified log transformation, since there are many 0 values.

```{r}
cube_root <- function(x) {
  sign(x) * abs(x)^(1/3)
}
ihs_transform <- function(x) {
  log(x + (x^2 + 1)^0.5)
}

map(list(sqrt, cube_root, ihs_transform), run_shapiro_abq)
```

Only a few variables can be normalized with these techniques. Now I'll try an inverse normal transformation. This transformation is based on ranking, and is able to normalize highly skewed distributions. The downside of this transformation is that the "distance" between each observation is lost, obscuring information about the variability within the variable. The `blom()` function which performs the transformation defaults to a "general" method. 

```{r}
xf_blom <- partial(blom, method = "blom")
run_shapiro_abq(xf_blom)
```

All of the variables are normalizable with this transformation, with the exception of vacancy percentage, so I'll remove it, and create a frame with the transformed data. Again, this is 78 times faster than with `dplyr`.

```{r}
abq_data %<>% fselect(-Vacant.Pct)
num_var_names <- num_vars(abq_data, return = "names")
abq_data_blom <- fmutate(abq_data, across(num_var_names, xf_blom))
```

```{r}
dmutate <- dplyr::mutate
dacross <- dplyr::across
dall_of <- dplyr::all_of
microbenchmark(
  collapse = abq_data %>%
    fmutate(across(num_var_names, xf_blom)),
  dyplr = abq_data %>%
    dmutate(dacross(dall_of(num_var_names), ~ xf_blom(.x)))
)
```

## Equality of variance (Bartlett's test)

Now that I have normally distributed variables, I need to address the second assumption of ANOVA tests, which is equality of variance among the districts. [Bartlett's test](https://en.wikipedia.org/wiki/Bartlett's_test) can be used for that purpose, but before running the tests, we can visualize the distributions with ridge plots.

```{r}
#| message: false
#| warning: false
plot_dist_ridge <- function(df, var) {
  ggplot(df, aes(
    x = .data[[var]], y = district,
    fill = factor(stat(quantile))
  )) +
    stat_density_ridges(
      geom = "density_ridges_gradient", calc_ecdf = TRUE,
      quantiles = 4, quantile_lines = TRUE, alpha = 0.7,
      jittered_points = TRUE, position = "points_sina"
    ) +
    scale_fill_viridis_d(name = "Quartiles") +
    theme(axis.title.x = element_blank())
}

map(num_var_names, 
    \(x) abq_data_blom %>% 
      plot_dist_ridge(x) +
      labs(title = x))
```

The null hypothesis in Bartlett's test is that all population variances are equal among the districts, while the alternative hypothesis is that at least two are different.

```{r}
# var_vals <- map(var_names, \(x) pull(abq_data_elf, x))
vars_list <- get_vars(abq_data_blom, num_var_names) %>% as.list
btests <- map2(
  num_var_names,
  vars_list,
  \(x, y) list(x, bartlett.test(y ~ district, abq_data_blom))
)
data.frame(
  variable = unlist(map(1:length(btests), \(x) btests[[x]][[1]])),
  statistic = unlist(map(1:length(btests), \(x) btests[[x]][[2]]$statistic)),
  p_value = unlist(map(1:length(btests), \(x) btests[[x]][[2]]$p.value %>% round(., 3)))
) %>% fsubset(p_value > .05)
```

Median house value, median income, percent Hispanic, and percentage of college graduates do not show homogeneity of variance, and so I will remove them for purposes of the classic ANOVA test.

```{r}
var_names_aov <- num_var_names[c(2:4, 9:11)]
var_vals_aov <- get_vars(abq_data_blom, var_names_aov) %>% as.list
```

## ANOVA

```{r}
aov_tests <- map(
  var_names_aov,
  \(x) list(x, aov(as.formula(glue("{x} ~ district")),
    data = abq_data_blom)
  )
)

unlist2d_nid <- partial(unlist2d, idcols = FALSE)

map2(
  1:length(aov_tests), var_names_aov,
  \(x, y) data.frame(
    variable = y,
    statistic = tidy(aov_tests[[x]][[2]])$statistic[1],
    p.value = tidy(aov_tests[[x]][[2]])$p.value[1]
  )
) %>% 
  unlist2d_nid
```

## Tukey's

```{r}
results_tky <- map(
  aov_tests,
  \(x) TukeyHSD(x[[2]])$district %>%
    as.data.frame %>%
    fsubset(`p adj` <= .05) %>%
    fmutate(variable = x[[1]]) %>%
    tibble::rownames_to_column("districts")
) %>%
  unlist2d(idcols = FALSE) %>%
  tidyr::separate(districts, c("District.1", "District.2"), sep = "-")

dist_names <- 
  sort(funique(c(results_tky$District.1, results_tky$District.2)))
var_names_tky <- funique(results_tky$variable)

# results_tky_dist <- map(
#   dist_names,
#   \(x) results_tky %>%
#     fsubset(District.1 == x | District.2 == x) %>%
#     fmutate(
#       district = x,
#       compare = if_else(District.1 == x, District.2, District.1),
#       .before = variable
#     ) %>% 
#     gv(-c(1:2)) %>% 
#     roworder(district, compare)
# )
```

```{r}
map(
  dist_names,
  \(x) results_tky %>%
    fsubset(District.1 == x | District.2 == x) %>%
    fmutate(
      district = x,
      compare = iif(District.1 == x, District.2, District.1)
    ) %>% 
    fselect(district, variable, compare, diff, `p adj`) %>%
    roworder(district, variable) %>% as_tibble
)
```


## Welch's tests


```{r}
map(
  num_var_names,
  \(x) welch_anova_test(as.formula(glue("{x} ~ district")), data = abq_data_blom)
  ) %>% 
  unlist2d_nid %>% gv(-7) %>% as_tibble
```

```{r}
results_gh <- map(
  num_var_names,
  \(x) games_howell_test(
    abq_data_blom,
    as.formula(glue("{x} ~ district"))
  ) %>%
    fsubset(p.adj.signif %!=% "ns") %>%
    fselect(group1, .y., group2, estimate, p.adj, p.adj.signif) %>%
    frename(.y. = "variable", group1 = "district", group2 = "compare")
) %>% 
  unlist2d_nid() %>% 
  roworder(district, variable)
```

```{r}
map(
  dist_names,
  \(x) results_gh %>%
    fsubset(district == x | compare == x) %>%
    fmutate(
      district = x,
      compare = iif(district == x, compare, district)
    ) %>% 
    as_tibble
)
```

# Spatial autocorrelation

```{r}
abq_data <- st_sf(abq_data, geometry = abq_data_geom)
nb <- poly2nb(abq_data, queen = F, snap = 25)
nbw <- nb2listw(nb, style = "W", zero.policy = T)

map(vars_list,
    \(x) moran.test(x, nbw, alternative = "two.sided") %>% tidy) %>% 
  unlist2d
```

```{r}
map(vars_list, \(x) moran.plot(x, nbw))
```



# Here

```{r}
make_cuts <- function(var) {
  quantiles <- classIntervals(var, 5, style = "quantile")
  cut(
    var, quantiles$brks,
    include.lowest = TRUE, dig.lab = 6
  ) %>%
    str_replace_all(",", " - ") %>%
    str_replace_all("[\\(\\]\\[]", "")
}

abq_data$quantiles <- make_cuts(abq_data$Income.Med)
tmp <- st_drop_geometry(abq_data) %>% fselect("Income.Med") %>% pull
make_cuts(tmp)

cuts <- map(num_var_names,
    \(x) make_cuts(st_drop_geometry(abq_data)[[x]]))

map2(
  cuts, num_var_names,
  \(cut, lab) abq_data %>%
    fselect(num_var_names) %>%
    ggplot() +
    annotation_map_tile(
      type = "osm",
      zoomin = -1, cachedir = "~/.cache/maps/"
    ) +
    geom_sf(aes(fill = cut), alpha = 0.7) +
    geom_sf(data = council_dists, aes(color = district), linewidth = 1, show.legend = F) +
    scale_fill_viridis_d() +
    scale_color_viridis_d() +
    labs(fill = lab) +
    theme_void()
)
```


```{r}
get_var <- function(df, var) {
  st_drop_geometry(df)[var] %>% pull 
    
}

localmoran(abq_data[["Income.Med"]], nbw, alternative = "greater")
lmoran <- localmoran(get_var(abq_data, "Income.Med"), nbw, alternative = "greater")
lmorans <- map(num_var_names,
               \(x) localmoran(get_var(abq_data, x), nbw, alternative = "greater"))
```

```{r}
abq_data$lmI <- lmoran[, "Ii"] # local Moran's I
abq_data$lmZ <- lmoran[, "Z.Ii"] # z-scores
abq_data$lmp <- lmoran[, "Pr(z > E(Ii))"]
```


```{r}
library(tmap)

p1 <- tm_shape(abq_data_nona) +
  tm_polygons(col = "vbl", title = "vbl", style = "quantile") +
  tm_layout(legend.outside = TRUE)

p2 <- tm_shape(abq_data_nona) +
  tm_polygons(col = "lmI", title = "Local Moran's I",
              style = "quantile") +
  tm_layout(legend.outside = TRUE)

p3 <- tm_shape(abq_data_nona) +
  tm_polygons(col = "lmZ", title = "Z-score",
              breaks = c(-Inf, 1.65, Inf)) +
  tm_layout(legend.outside = TRUE)

p4 <- tm_shape(abq_data_nona) +
  tm_polygons(col = "lmp", title = "p-value",
              breaks = c(-Inf, 0.05, Inf)) +
  tm_layout(legend.outside = TRUE)

tmap_arrange(p1, p2, p3, p4)
```

```{r}
tmap5 <- tm_shape(abq_data_nona) + 
  tm_polygons(col = "lmZ",
              title = "Local Moran's I", 
              style = "fixed",
              breaks = c(-Inf, -1.96, 1.96, Inf),
              labels = c("Negative SAC", "No SAC", "Positive SAC"),
              palette =  c("blue", "white", "red")) +
  tm_layout(legend.outside = TRUE)
tmap5
```

## Clusters

```{r}
lmoran <- localmoran(abq_data_nona$vbl, nbw, alternative = "two.sided")
head(lmoran)
```

```{r}
abq_data_nona$lmp <- lmoran[, 5]
mp <- moran.plot(as.vector(scale(abq_data_nona$vbl)), nbw)
```

```{r}
abq_data_nona$quadrant <- NA
abq_data_nona[(mp$x >= 0 & mp$wx >= 0) & (abq_data_nona$lmp <= 0.05), "quadrant"] <- 1
abq_data_nona[(mp$x <= 0 & mp$wx <= 0) & (abq_data_nona$lmp <= 0.05), "quadrant"] <- 2
abq_data_nona[(mp$x >= 0 & mp$wx <= 0) & (abq_data_nona$lmp <= 0.05), "quadrant"] <- 3
abq_data_nona[(mp$x <= 0 & mp$wx >= 0) & (abq_data_nona$lmp <= 0.05), "quadrant"] <- 4
abq_data_nona[(abq_data_nona$lmp > 0.05), "quadrant"] <- 5
```

```{r}
tm_shape(abq_data_nona) +
  tm_fill(
    col = "quadrant", title = "",
    breaks = c(1, 2, 3, 4, 5, 6),
    palette = c("red", "blue", "lightpink", "skyblue2", "white"),
    labels = c(
      "High-High", "Low-Low", "High-Low",
      "Low-High", "Non-significant"
    )
  ) +
  tm_legend(text.size = 1) + tm_borders(alpha = 0.5) +
  tm_layout(frame = FALSE, title = "Clusters") +
  tm_layout(legend.outside = TRUE)
```




# Save data

```{r}
data_sets <- c(
  "bern_data", "abq_data", "abq_data_blom"
)

walk2(
  map(data_sets, \(x) get(x)),
  data_sets,
  \(df, level) st_write(df, "data/district_variance.gpkg",
    layer = level, append = F
  )
)
abq_data_blom %<>%
  st_cast("MULTIPOLYGON")
```






