---
title: "District Variance"
format: gfm
---

# Introduction

In the [last article](https://biscotty.net/posts/data-science/r-census/abq-demographics) I compared age, race, and educational attainment between council districts in Albuquerque. In some cases, the variation was striking. In particular, the third district seemed quite different from the others in a number of ways. Now I would like to apply some statistical testing to see if this impression is backed up statistically.

To make the comparison, I will use some economic and demographic information obtained from `census.gov` using the `tidycensus` package.

# Setup

I will begin by loading the required libraries.

```{r}
#| message: false
options(paged.print = FALSE,
        tigris_use_cache = TRUE)
libraries <- list(
  "corrr", "ggspatial", "ggtext", "patchwork", "sf",
  "spatialreg", "spdep", "tidycensus", "tidyverse", 
  "zeallot", "DescTools", "GWmodel", "scales", "glue",
  "ggpubr", "units", "rstatix", "rcompanion", "collapse"
)
invisible(lapply(libraries, library, character.only = TRUE))
```

I'd like to check variance across a number of economic and demographic variables.

```{r}
year <- 2024
crs <- 6528
caption <- str_glue("Source: census.gov, acs5, {year}")
```

## Load Data

```{r}
vars_census <- c(
  House.Value.Med = "B25077_001",
  Renter.Occupied = "B25003_003",
  Vacant = "B25002_003",
  Income.Med = "DP03_0062",
  Gini = "B19083_001",
  Rent.Med = "B25031_001",
  Foreign.Born.Pct = "DP02_0094P",
  Hisp.Pct = "DP05_0090P",
  In.Poverty = "B17001A_002",
  Age.Med = "B01002_001",
  College.Pct = "DP02_0068P",
  Population.Total = "B01003_001"
)
```

```{r}
#| eval: false
tidycensus::load_variables(year, "acs5/profile") %>% 
  View()
```
 

```{r}
bern_data <- get_acs(
  geography = "tract",
  state = "NM",
  county = "Bernalillo",
  variables = vars_census,
  output = "wide",
  year = year,
  geometry = TRUE,
  cache_table = TRUE
) %>%
  select(-2 & !ends_with("M")) %>%
  rename_with(~ sub("E$", "", .)) %>% 
  st_transform(crs)
```

## Clean Data

Bernalillo County contains some regions which are far from the urban area, these regions being largely unpopulated and distantly connected to the city. I want to look at the areas of the county proximate to the city, so I will define bounds which encompass this region.



One of the peculiarities of Albuquerque is the fact that there are numerous areas in the geographic area of Albuquerque which are not incorporated into the city. This includes "islands" in the middle of districts, as well as areas on the fringe of the city limits. In order to capture this data, I will create an "Unincorporated" district, and choose a geographic extent for the study area. In order to specify the "Unincororated" area, I will take the _difference_ between the full county and the areas attributed to one of the districts or Los Ranchos. After giving this data their label, I bind the identified region back to the `council_dists` data frame.

### Council Districts

```{r}
council_dists <-
  st_read("data/BC_CityCouncil/ABQ_CityCouncils.shp") %>%
  select(district = DISTRICTNU) %>%
  arrange(district) %>% 
  mutate(district = paste("District", district))
los_ranchos <-
  st_read("data/LosRanchos/LosRanchos.shp") %>%
  select(district = Name)
council_dists <-
  rbind(council_dists, los_ranchos) %>%
  st_transform(crs)

difference <- 
   st_difference(
  st_as_sfc(st_bbox(council_dists)),
  st_union(council_dists$geometry)
) %>% 
  st_as_sf() %>%
  rename(geometry = x) %>% 
  mutate(district = "Unincorporated")

council_dists <- bind_rows(difference, council_dists)
```

```{r}
dist_box = c(xmin = 452000, xmax = 479833, 
             ymin = 443500, ymax = 467858)
council_dists <- council_dists %>% 
  st_crop(dist_box)
```



```{r}
ggplot(council_dists, aes(color = district)) +
  geom_sf() +
  theme_void()
```

### County Census Data

Let's take a look at the data before splitting it.

```{r}
summary(bern_data)
```

There is some clean-up to do. There are some `na`s to explore, and oddities like areas where the population total is 0. I'll start there.

```{r}
bern_data %>% 
  filter(Population.Total == 0)
```

```{r}
ggplot(bern_data) +
  geom_sf(color = "red") +
  geom_sf(data = bern_data %>% filter(Population.Total == 0), fill = "navy") +
  annotation_map_tile(
    type = "osm", alpha = 0.7,
    zoomin = -1, cachedir = "~/.cache/maps/"
  ) +
  theme_void() +
  lims(
    x = c(435000, 475000),
    y = c(444000, 455000)
  )
```

This tract is the state fairgrounds. I'll just remove it.

```{r}
bern_data <- bern_data %>% 
  filter(Population.Total > 0)
summary(bern_data)
```

The population of 16 is still suspicious. 

```{r}
bern_data %>% 
  filter(Population.Total < 20)
```

```{r}
ggplot(bern_data) +
  geom_sf(color = "red") +
  geom_sf(data = bern_data %>% filter(Population.Total < 20), fill = "navy") +
  annotation_map_tile(
    type = "osm", alpha = 0.7,
    zoomin = -1, cachedir = "~/.cache/maps/"
  ) +
  theme_void()
```

This tract contains federal land, with no useful data. I'll remove this as well.

```{r}
bern_data <- bern_data %>% 
  filter(Population.Total > 20)
```

This only leaves `na`s in two variables. Let's take a closer look.

```{r}
bern_data %>% 
  filter(if_any(everything(), is.na))
```

This data seems worth preserving. I will interpolate these values based on the median values, but I would like the values to be based on the median for the specific district. I will need to split the districts at this point in order to calculate the values.

Splitting the data involves intersecting the county data with the districts. Since some of the variables are extensive, representing counts, these values will need to be apportioned based on the percentage area represented by each subsection.

## Create District Dataset

```{r}
dist_data <- bern_data %>% 
  mutate(area = st_area(.)) %>% 
  st_intersection(council_dists) %>% 
  mutate(
    area_split = st_area(.),
    area_pct = as.numeric(area_split / area),
    Population = round(Population.Total * area_pct, 0),
    Population.Density = Population / area_split,
    Renter.Occupied = round(Renter.Occupied * area_pct, 0),
    Vacant = round(Vacant * area_pct, 0),
    In.Poverty = round(In.Poverty * area_pct, 0)
    )
```

```{r}
summary(dist_data)
```


Now I can calculate the per-district averages for the two variables.


```{r}
vars_na <- c("House.Value.Med", "Rent.Med")

var_avgs <- dist_data %>%
  st_drop_geometry() %>%
  group_by(district) %>%
  summarise(across(
    all_of(vars_na),
    list(dist = \(x) median(x, na.rm = T))
  )) %>%
  ungroup()
```

I'll make a small helper function to fill in the missing values.

```{r}
fill_avgs <- function(var, replace) {
  if_else(is.na(var), replace, var)
}

dist_data <- dist_data %>%
  left_join(var_avgs) %>% 
  mutate(
    House.Value.Med = fill_avgs(House.Value.Med, House.Value.Med_dist),
    Rent.Med = fill_avgs(Rent.Med, Rent.Med_dist),
  ) %>% 
  select(-c(Population.Total, area, area_split, ends_with("dist")))

summary(dist_data)
```

The split data now contains more areas with a population is 0. As a result of the splitting, there are numerous small fragments containing little information. To address this, I will remove all of the new rows which represent less than 1\% of the original area.

```{r}
dist_data <- dist_data %>% 
  filter(area_pct > .01)
summary(dist_data)
```

That looks good. I'll remove the percentage column, and can proceed with exploring the data.

```{r}
dist_data <- dist_data %>% 
  select(-c(area_pct, Population, Population.Density))
```




# Check Distributions

## Visualize

```{r}
var_names <- names(dist_data)[-c(1, 13:14)]
```

```{r}
plot_dist <- function(df, var) {
  df %>% 
   ggboxplot("district", var, fill = "district") +
  geom_point(position = position_jitter(width = 0.1)) +
  theme(axis.text.x = element_text(angle = 45, vjust = .5)) +
  guides(fill = "none")
}

map(var_names, \(x) dist_data %>% plot_dist(x))
```

```{r}
plot_hist <- function(df, var) {
  df %>% 
    ggplot(aes(.data[[var]])) +
    geom_histogram(aes(y = after_stat(density)),
    fill = "lightgray", col = "black", bins = 20
  ) +
  geom_line(stat = "density", adjust = 2, color = "blue")
}

hists <- map(var_names, \(x) dist_data %>% plot_hist(x))
vars_list <- map(var_names, \(x) pull(dist_data, x))
lines <- map(vars_list, ggqqplot)
map(1:length(hists),
     \(x) hists[[x]] + lines[[x]])
```

```{r}
abq_data <- dist_data %>% 
  filter(!district %in% c("Unincorporated", "Los Ranchos"))
```

## Shapiro tests

The shapiro test

### Raw

```{r}
map(
  map(
    var_names,
    \(x) pull(abq_data, x)
  ),
  shapiro_test
) %>%
  collapse::unlist2d() %>%
  mutate(
    variable = var_names,
    sig = if_else(p.value < .05, "Yes", "No")
  ) %>%
  select(-1) %>%
  filter(sig == "No")
```

### Square-root transform

```{r}
run_shapiro <- function(df, vars, fctn) {
  data <- map(
    vars,
    \(x) pull(df, x)
  ) %>%
    lapply(fctn)
  map(data, shapiro_test) %>%
    collapse::unlist2d() %>%
    mutate(
      variable = vars,
      sig = if_else(p.value < .05, "Yes", "No")
    ) %>%
    select(-1) %>%
    filter(sig == "No")
}

run_shapiro(abq_data, var_names, sqrt)
```

### Cube root

```{r}
cube_root <- function(x) {
  sign(x) * abs(x)^(1/3)
}

run_shapiro(abq_data, var_names, cube_root)
```

### Log transform (ihs)

```{r}
ihs_transform <- function(x) {
  log(x + (x^2 + 1)^0.5)
}

run_shapiro(abq_data, var_names, ihs_transform)
```

### Elf transform

```{r}
run_shapiro(abq_data, var_names, blom)
```

Remove non-standard variables.

```{r}
abq_data <- abq_data %>% 
  select(-c("GEOID", "Vacant"))
var_names <- var_names[-3]
```


```{r}
abq_data_elf <- abq_data %>% 
  mutate(across(all_of(var_names), ~ blom(.x)))
```

## Distribution Plots

```{r}
map(var_names, \(x) abq_data_elf %>% plot_dist(x))
```

```{r}
hists <- map(var_names, \(x) abq_data_elf %>% plot_hist(x))
var_vals <- map(var_names, \(x) pull(abq_data_elf, x))
lines <- map(var_vals, ggqqplot)
map(1:10,
     \(x) hists[[x]] + lines[[x]])
```

## Equality of variance (Bartlett's test)


```{r}
# var_vals <- map(var_names, \(x) pull(abq_data_elf, x))
btests <- map2(var_names, 
     map(var_names, \(x) pull(abq_data_elf, x)),
    \(x, y) list(x, bartlett.test(y ~ district, abq_data_elf)))
tibble(
  variable = unlist(map(1:length(btests), \(x) btests[[x]][[1]])),
  statistic = unlist(map(1:length(btests), \(x) btests[[x]][[2]]$statistic)),
  p_value = unlist(map(1:length(btests), \(x) btests[[x]][[2]]$p.value %>% round(., 3)))
)
```

```{r}
var_names_aov <- var_names[2:7]
btests <- map2(var_names_aov, 
     map(var_names_aov, \(x) pull(abq_data_elf, x)),
    \(x, y) list(x, bartlett.test(y ~ district, abq_data_elf)))
tibble(
  variable = unlist(map(1:length(btests), \(x) btests[[x]][[1]])),
  statistic = unlist(map(1:length(btests), \(x) btests[[x]][[2]]$statistic)),
  p_value = unlist(map(1:length(btests), \(x) btests[[x]][[2]]$p.value %>% round(., 3)))
)
```

# AOV

## ANOVA

```{r}
aov_tests <- map(
  var_names_aov,
  \(x) list(x, aov(as.formula(glue::glue("{x} ~ district")),
    data = abq_data_elf)
  )
)
map2(1:length(aov_tests), var_names_aov, 
     \(x, y) list(y, summary(aov_tests[[x]][[2]])))

```

## Welch's tests

```{r}
var_names_welch <- var_names[c(1, 8:10)]
map(
  var_names_welch,
  \(x) list(x, oneway.test(as.formula(glue::glue("{x} ~ district")),
    data = abq_data_elf, var.equal = F)
  )
)
```



```{r}
map(
  var_names_welch,
  \(x) list(x, welch_anova_test(as.formula(glue::glue("{x} ~ district")),
    data = abq_data_elf %>% st_drop_geometry)
  )
)
```

## Ridge Plots

```{r}
#| message: false
#| warning: false
library(ggridges)
plot_dist_ridge <- function(df, var) {
  ggplot(df, aes(
    x = .data[[var]], y = district,
    fill = factor(stat(quantile))
  )) +
    stat_density_ridges(
      geom = "density_ridges_gradient", calc_ecdf = TRUE,
      quantiles = 4, quantile_lines = TRUE, alpha = 0.7,
      jittered_points = TRUE, position = "points_sina"
    ) +
    scale_fill_viridis_d(name = "Quartiles") +
    theme(axis.title.x = element_blank())
}
```

```{r}
#| warning: false
map(var_names, 
    \(x) abq_data_elf %>% 
      plot_dist_ridge(x) +
      labs(title = x))
```

## Tukey's

```{r}
results_tky <- map(
  aov_tests,
  \(x) TukeyHSD(x[[2]])$district %>% 
    as.data.frame %>%
  filter(`p adj` <= .05) %>% 
  mutate(variable = x[[1]], .before = 1) %>% 
    rownames_to_column("districts")) %>% 
  unlist2d(idcols = FALSE) %>% 
  separate(districts, c("District.1", "District.2"), sep = "-")

dist_names <- unique(c(results_tky$District.1, results_tky$District.2)) %>% 
  sort()
var_names_tky <- unique(results_tky$variable)

results_tky_dist <- map(
  dist_names,
  \(x) results_tky %>%
    filter(District.1 == x | District.2 == x) %>%
    mutate(
      district = x,
      compare = if_else(District.1 == x, District.2, District.1),
      .before = variable
    ) %>% 
    select(-(1:2)) %>% 
    arrange(district, compare)
)
```

```{r}
map(
  dist_names,
  \(x) results_tky %>%
    filter(District.1 == x | District.2 == x) %>%
    mutate(
      district = x,
      compare = if_else(District.1 == x, District.2, District.1),
      .before = variable
    ) %>% 
    select(-(1:2)) %>% 
    arrange(district, compare) %>% 
    as_tibble()
)
```

```{r}
map(unique(results_tky$variable),
    \(x) results_tky %>% 
      filter(variable == x) %>% 
  relocate(variable, .before = 1) %>% 
  as_tibble())
```

## Linear Model

```{r}

map(var_names,
    \(x) list(x, lm(as.formula(glue("{x} ~ district")),
      data = abq_data_elf
    ) %>% summary()))
```

```{r}
var_names_str <- glue_collapse(var_names, sep = " + ")
lm(as.formula(glue("district ~ {var_names_str}")),
   data = abq_data_elf %>% st_drop_geometry())
abq_data_elf <- abq_data_elf %>% 
  mutate(dist_num = parse_number(district))
abq_data <- abq_data %>% 
  mutate(dist_num = parse_number(district))
form <- as.formula(glue("dist_num ~ {var_names_str}"))
lm(form,
   data = abq_data %>% st_drop_geometry()) %>% 
  summary()
summary(abq_data_elf)
```

```{r}
data_sets <- c(
  "dist_data", "bern_data", "abq_data", "abq_data_elf"
)

walk2(
  map(data_sets, \(x) get(x)),
  data_sets,
  \(df, level) st_write(df, "data/district_variance.gpkg",
    layer = level, append = F
  )
)
```






