---
title: "Collapse Data Object Conversions"
format: gfm
---

[Source](https://fastverse.org/collapse/articles/collapse_intro.html)

```{r}
library(collapse)
library(magrittr)
library(microbenchmark)
```


`collapse` provides a set of functions to speed these conversions

- `qDF` - convert to data.frame
- `qDT` - convert to data.table
- `qTBL` - convert to tibble
- `qM` - converts vectors, higher-dim arrays, data frames, and suitable lists to matrix
- `qF` - convert vector to factor
- `mrtl` - row-wise convert a matrix into list, data.frame or data.table
- `mctl` - column-wise convert a matrix into list, data.frame or data.table

```{r}
str(EuStockMarkets)
```

```{r}
microbenchmark(
  qDT(wlddev),
  qDT(EuStockMarkets),
  data.table::as.data.table(wlddev),
  as.data.frame(EuStockMarkets)
)
```

By default these functions drop all unnecessary attributes from matrices or lists / data frames in the conversion, but this can be changed using the keep.attr = TRUE argument.

`row.names.col` saves names/row names in a column.

```{r}
head(qDT(mtcars, "car"))
```

```{r}
(N_distinct <- fndistinct(GGDC10S))
```

Convert to data frame, preserving names

```{r}
head(qDF(N_distinct, "variable"))
```

This converts the matrix to a list of 1860 row-vectors of length 4.

```{r}
microbenchmark(mrtl(EuStockMarkets))
```

```{r}
mrtl(EuStockMarkets) %>% head()
```

```{r}
microbenchmark(rowSums(qM(mtcars)), 
               rowSums(mtcars),
               kit::psum(mtcars))
```

Factor From Character

```{r}
str(wlddev$country)
fndistinct(wlddev$country)
microbenchmark(
  qF(wlddev$country),
  as.factor(wlddev$country)
)
```

Factor from numeric

```{r}
fndistinct(wlddev$PCGDP)
microbenchmark(
  qF(wlddev$PCGDP),
  as.factor(wlddev$PCGDP)
)
```









