---
title: "collapse"
format: gfm
---

[Source](https://fastverse.org/collapse/articles/collapse_intro.html)

```{r}
library(collapse)
library(magrittr)
```


# Data and summary tools

## World Bank Development Data

```{r}
library(collapse)
head(wlddev)
```

Name attributes, set with `vlabels()`.

```{r}
namlab(wlddev, class = TRUE)
```

```{r}
descr(wlddev)
```

Convert output to data frame

```{r}
descr(wlddev) %>% 
  as.data.frame() %>% 
  head()
```

Check which variables are time-varying.

```{r}
varying(wlddev, wlddev$iso3c)
```

Check variation within each country

```{r}
varying(wlddev, wlddev$iso3c, any_group = F) %>% 
  head()
```

In general data is varying if it has two or more distinct non-missing values. We could also take a closer look at observation counts and distinct values using:

```{r}
fnobs(wlddev, wlddev$iso3c) %>% head()
fndistinct(wlddev, wlddev$iso3c) %>% head()
```

Summarize. `higher = TRUE` provides skewness.

```{r}
qsu(wlddev, cols = 9:12, higher = TRUE)
```

Same by region

```{r}
qsu(wlddev, by = ~region, cols = 9:12, vlabels = TRUE, higher = TRUE)
```

```{r}
qsu(wlddev, pid = ~ iso3c, cols = c(1,4,9:12), vlabels = TRUE, higher = TRUE)
```

The above output reports 3 sets of summary statistics for each variable: Statistics computed on the Overall (raw) data, and on the Between-country (i.e. country averaged) and Within-country (i.e. country-demeaned) data1. This is a powerful way to summarize panel data because aggregating the data by country gives us a cross-section of countries with no variation over time, whereas subtracting country specific means from the data eliminates all cross-sectional variation. 

The N/T columns shows that for `PCGDP` we have 9470 total observations, that we observe GDP data for 206 countries and that we have on average 45.97 observations (time-periods) per country. In contrast the GINI Index is only available for 167 countries with 10.4 observations on average. The Overall and Within mean of the data are identical by definition, and the Between mean would also be the same in a balanced panel with no missing observations. In practice we have unequal amounts of observations for different countries, thus countries have different weights in the Overall mean and the difference between _Overall_ and _Between_-country mean reflects this discrepancy. The most interesting statistic in this summary arguably is the standard deviation, and in particular the comparison of the _Between_-SD reflecting the variation between countries and the _Within_-SD reflecting average variation over time. This comparison shows that `PCGDP`, `LIFEEX` and `GINI` vary more between countries, but ODA received varies more within countries over time. The 0 _Between_-SD for the year variable and the fact that the _Overall_ and _Within_-SD are equal shows that year is individual invariant. Thus `qsu` also provides the same information as `varying`, but with additional details on the relative magnitudes of cross-sectional and time series variation. It is also a common pattern that the _kurtosis_ increases in within-transformed data, while the _skewness_ decreases in most cases.

Perform the same across regions

```{r}
(m <- qsu(wlddev, by = ~ region, pid = ~ iso3c, cols = 9:12, 
    vlabels = TRUE, higher = TRUE))
```

```{r}
class(m); str(m)
```

Notice that the output here is a 4D array of summary statistics, which we could also subset ([) or permute (aperm) to view these statistics in any convenient way. 

Output as nested list.

```{r}
l <- qsu(wlddev, by = ~ region, pid = ~ iso3c, cols = 9:12,
          vlabels = T, higher = T, array = F)
str(l, give.attr = F)
```

Convert to data frame

```{r}
unlist2d(l, idcols = c("Variable", "Trans"), row.names = "Region") %>% 
  head()
```

Weighted pairwise correlations

```{r}
pwcor(wlddev[9:12], N = TRUE, P = TRUE)
```

Pairwise correlations computed on averaged and within-transformed data.

```{r}
pwcor(fmean(wlddev[9:12], wlddev$iso3c), N = TRUE, P = TRUE) %>% 
  print(show = "lower.tri")
```

```{r}
pwcor(fwithin(wlddev[9:12], wlddev$iso3c), P = TRUE) %>% 
  print(show = "lower.tri")
```

```{r}
pwnobs(wlddev)
```

Note that both pwcor/pwcov and pwnobs are faster on matrices.

## GGDC 10-sector database

The Groningen Growth and Development Centre 10-Sector Database provides long-run data on sectoral productivity performance in Africa, Asia, and Latin America. Variables covered in the data set are annual series of value added (VA, in local currency), and persons employed (EMP) for 10 broad sectors.

```{r}
head(GGDC10S); class(GGDC10S); namlab(GGDC10S, class = T)
```

```{r}
unique(GGDC10S$Variable)
```

Problems: `VA` is in local currency, and `VA` and `EMP` share the same column. To address the first problem, convert to percentages based on the sector sum in column 16. The second problem is solved using group statistics.

```{r}
pGGDC10S <- ftransformv(GGDC10S, 6:15, `*`, 100 / SUM)
su <- qsu(pGGDC10S, by = ~ Variable, pid = ~ Variable + Country,
          cols = 6:16, higher = T)
str(su)
```

Permute the array to a more readable format

```{r}
aperm(su, c(4L, 2L, 3L, 1L))
```

The statistics show that the dataset is very consistent: Employment data cover 42 countries and 53 time-periods in almost all sectors. Agriculture is the largest sector in terms of employment, amounting to a 35% share of employment across countries and time, with a standard deviation (SD) of around 27%. The between-country SD in agricultural employment share is 24% and the within SD is 12%, indicating that processes of structural change are very gradual and most of the variation in structure is between countries. The next largest sectors after agriculture are manufacturing, wholesale and retail trade and government, each claiming an approx. 15% share of the economy. In these sectors the between-country SD is also about twice as large as the within-country SD.

In terms of value added, the data covers 43 countries in 50 time-periods. Agriculture, manufacturing, wholesale and retail trade and government are also the largest sectors in terms of VA, but with a diminished agricultural share (around 17%) and a greater share for manufacturing (around 20%). The variation between countries is again greater than the variation within countries, but it seems that at least in terms of agricultural VA share there is also a considerable within-country SD of 8%. This is also true for the finance and real estate sector with a within SD of 9%, suggesting (using a bit of common sense) that a diminishing VA share in agriculture and increased VA share in finance and real estate was a pattern characterizing most of the countries in this sample.

## Plot

```{r}
library(ggplot2)
plotGGDC <- function(ctry) {
  fsubset(GGDC10S, Country == ctry, Variable, Year, AGR:SUM) %>% 
    ftransform(fselect(., AGR:OTH) %>% 
                 lapply(`*`, 1 / SUM) %>% 
                 replace_outliers(0, NA, "min")) %>% 
    ftransform(Variable = recode_char(Variable,
                                      VA = "Value Added Share",
                                      EMP = "Employment Share"),
               SUM = NULL) %>% 
    qDT() %>% 
    data.table::melt(1:2, variable.name = "Sector", na.rm = T) %>% 
    ggplot(aes(x = Year, y = value, fill = Sector)) +
    geom_area(position = "fill", alpha = 0.9) +
    labs(x = NULL, y = NULL) +
    theme_linedraw(base_size = 14L) +
    facet_wrap(~ Variable) +
    scale_fill_manual(values = sub("#00FF66", "#00CC66", rainbow(10L))) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 7L), expand = c(0, 0)) +
    scale_y_continuous(scales::pretty_breaks(n = 10L), expand = c(0,0),
                       labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 315, hjust = 0, margin = ggplot2::margin(t = 0)),
          strip.background = element_rect(colour = "grey20", fill = "grey20"),
          strip.text = element_text(face = "bold"))    
}

plotGGDC("BWA")
```

# Fast Data Manipulation

## Selecting and Replacing Columng

```{r}
wlddev %>% 
  fselect(country, year, PCGDP:ODA) %>% 
  head(2)
```

```{r}
wlddev %>% 
  fselect(-country, -year, -(PCGDP:ODA)) %>% 
  head(2)
```

```{r}
wlddev %>% 
  fselect(-c(country, year, PCGDP:ODA)) %>% 
  head(2)
```

```{r}
library(microbenchmark)
microbenchmark(fselect = collapse::fselect(wlddev, country, year, PCGDP:ODA),
               select = dplyr::select(wlddev, country, year, PCGDP:ODA))
```

`fselect` can also replace

```{r}
fselect(wlddev, PCGDP:POP) <- lapply(fselect(wlddev, PCGDP:POP), log)
head(wlddev, 2)
```

Efficient deletion

```{r}
fselect(wlddev, country, year, PCGDP:POP) <- NULL
head(wlddev, 2)
```

```{r}
rm(wlddev)
```

Other information returned by `fselect`.

```{r}
wlddev %>% 
  fselect(PCGDP:POP, return = "names")
wlddev %>% 
  fselect(PCGDP:POP, return = "indices")
wlddev %>% 
  fselect(PCGDP:POP, return = "named_indices")
wlddev %>% 
  fselect(PCGDP:POP, return = "logical")
wlddev %>% 
  fselect(PCGDP:POP, return = "named_logical")
```

There are no special methods for grouped tibbles.

Select variables with `get_vars`

```{r}
wlddev %>% 
  get_vars(9:13) %>% 
  head(1)
wlddev %>% 
  get_vars( c("PCGDP","LIFEEX","GINI","ODA","POP")) %>% 
  head(1)
wlddev %>% 
  get_vars("[[:upper:]]", regex = T) %>% 
  head(1)
wlddev %>% 
  get_vars("PC|LI|GI|OD|PO", regex = T) %>% 
  head(1)
wlddev %>% 
  get_vars(c("PC","LI","GI","OD","PO"), regex = TRUE) %>% 
  head(1)
wlddev %>% 
  get_vars(is.numeric) %>% 
  head(1)
wlddev %>% 
  get_vars(is.numeric, return = "names") %>% 
  head(1)
```

Can also replace with `get_vars`

```{r}
get_vars(wlddev, 9:13) <- lapply(
  get_vars(wlddev, 9:13), log
)
get_vars(wlddev, 9:13) <- NULL
head(wlddev)
rm(wlddev)
```

collapse offers a set of functions to efficiently select and replace data by data type: `num_vars`, `cat_vars` (for categorical = non-numeric columns), `char_vars`, `fact_vars`, `logi_vars` and `date_vars` (for date and date-time columns).


```{r}
num_vars(wlddev) %>% head(2)
cat_vars(wlddev) %>% head(2)
fact_vars(wlddev) %>% head(2)
```

Replacing

```{r}
fact_vars(wlddev) <- fact_vars(wlddev)
```

## Subsetting

`fsubset` allows multiple comma-separated select arguments after the subset argument, and it also preserves all attributes of subsetted columns:


```{r}
fsubset(GGDC10S, Variable == "VA" & Year > 1990,
        Country, Year, AGR:GOV) %>% head(2)
```

```{r}
fsubset(GGDC10S, Variable == "VA" & Year > 1990,
        -(Regioncode:Variable), -(OTH:SUM)) %>% head(2)
```

```{r}
ss(GGDC10S, 1:2, 6:16)
ss(GGDC10S, -(1:2), c("AGR","MIN")) %>% head(2)
```


```{r}
microbenchmark(
  base = subset(GGDC10S, Variable == "VA" & Year > 1990, AGR:SUM),
  collapse = fsubset(GGDC10S, Variable == "VA" & Year > 1990, AGR:SUM)
)
microbenchmark(GGDC10S[1:10, 1:10], ss(GGDC10S, 1:10, 1:10))
```

## Reordering Rows and Columns

Replaces `arrange`. Negative variable names indicate descending sort.

```{r}
roworder(GGDC10S, -Variable, Country) %>% 
  ss(1:2, 1:8)
```


```{r}
microbenchmark(collapse = collapse::roworder(GGDC10S, -Variable, Country),
               dplyr = dplyr::arrange(GGDC10S, desc(Variable), Country))
```

The function roworderv is a standard evaluation analogue to roworder:

```{r}
roworderv(GGDC10S, c("Variable", "Country"), 
          decreasing = c(TRUE, FALSE)) %>% 
  ss(1:2, 1:8)
```

Move or exchange rows with `roworderv`

```{r}
GGDC10S %>% roworderv(
  neworder = which(GGDC10S$Country == "GHA")
) %>% 
  ss(1:2, 1:8)
GGDC10S %>% roworderv(
  neworder = which(GGDC10S$Country == "BWA"), pos = "end"
) %>% 
  ss(1:2, 1:8)
```

`pos = "exchange"` arranges selected rows in the order they are passed, without affecting other rows

```{r}
GGDC10S %>% roworderv(
  neworder = with(GGDC10S, c(which(Country == "GHA"),
                             which(Country == "BWA"))),
  pos = "exchange"
) %>% 
  ss(1:2, 1:8)
```

## Transforming and Computing New Columns

`ftransform` always returns the entire data frame.

```{r}
GGDC10S %>% 
  ftransform(AGR_perc = AGR / SUM * 100,
             Year = as.integer(Year),
             AGR = NULL) %>% 
  tail(2)
```

```{r}
GGDC10S %>% 
  ftransform(MIN_mean = fmean(MIN),
             Intercept = 1) %>% 
  tail(2)
```

The modification `ftransformv` exists to transform specific columns using a function:


```{r}
GGDC10S %>% 
  ftransformv(6:16, log) %>% 
  tail(2)
GGDC10S %>% 
  ftransformv(6:16, `*`, 100 / SUM) %>% 
  tail(2)
GGDC10S %>% 
  ftransformv(is.numeric, log) %>% 
  tail(2)
```

```{r}
GGDC10S %>%
  ftransform(num_vars(.) %>%
    lapply(log) %>%
    replace_Inf()) %>%
  tail(2)
```

The function settransform(v) can be used to change the input data frame by reference: 


```{r}
settransform(GGDC10S, FIRE_MAN = FIRE / MAN,
                      Regioncode = NULL, Region = NULL)
tail(GGDC10S, 2)
```

```{r}
rm(GGDC10S)
settransformv(GGDC10S, 6:16, `*`, 100 / SUM)
tail(GGDC10S)
```

`fcompute` can be used to compute new columns, returning the computed columns in a new data frame.

```{r}
fcompute(GGDC10S, AGR_perc = AGR / SUM * 100, FIRE_MAN = FIRE / MAN) %>% 
  tail(2)
```

## Adding and Binding Columns

`add_vars`

```{r}
rm(wlddev)
add_vars(wlddev) <- 
  get_vars(wlddev, 9:13) %>% 
  lapply(log10) %>% 
  add_stub("log10.")
head(wlddev, 2)
```

Replace at other positions

```{r}
rm(wlddev)
add_vars(wlddev, "front") <- 
  get_vars(wlddev, 9:13) %>% 
  lapply(log10) %>% 
  add_stub("log10.")
head(wlddev, 2)
```

```{r}
rm(wlddev)

add_vars(wlddev, c(10L, 12L, 14L, 16L, 18L)) <- 
  get_vars(wlddev, 9:13) %>%
  lapply(log10) %>%
  add_stub("log10.")
head(wlddev, 2)
```

Using `add_vars` without replacement.

```{r}
rm(wlddev)
add_vars(
  wlddev, 
  get_vars(wlddev, 9:13) %>% lapply(log) %>% add_stub("log."),
  get_vars(wlddev, 9:13) %>% lapply(log10) %>% add_stub("log10.")
) %>% head(2)
```

```{r}
add_vars(wlddev,
  get_vars(wlddev, 9:13) %>% lapply(log) %>% add_stub("log."),
  get_vars(wlddev, 9:13) %>% lapply(log10) %>% add_stub("log10."),
  pos = c(10L, 13L, 16L, 19L, 22L, 11L, 14L, 17L, 20L, 23L)
) %>% head(2)
```


```{r}
identical(cbind(wlddev, wlddev), add_vars(wlddev, wlddev))

microbenchmark(cbind(wlddev, wlddev), add_vars(wlddev, wlddev))
```

## Renaming Columns

```{r}
GGDC10S %>% 
  frename(AGR = Agriculture, MIN = Mining) %>% 
  head(2)
```

```{r}
frename(GGDC10S, tolower) %>% head(2)

frename(GGDC10S, tolower, cols = .c(AGR, MIN)) %>% head(2)
```

Rename by reference

```{r}
setrename(GGDC10S, AGR = Agriculture, MIN = Mining)
head(GGDC10S, 2)
rm(GGDC10S)
```

## Shortcuts

Only recommended for personal scripts

- `fselect` -> `slt`
- `fsubset` -> `sbt`
- `ftransform(v)` -> `tfm(v)`
- `settransform(v)` -> `settfm(v)`
- `get_vars -> gv`
- `num_vars -> nv`
- `add_vars` -> `av`. 

## Missing Values/Rows

```{r}
microbenchmark(
  na_omit(wlddev, na.attr = TRUE), 
  na.omit(wlddev))
```

Act on certain columns only

```{r}
na_omit(wlddev, cols = .c(PCGDP, LIFEEX)) %>% 
  head(2)
na_omit(wlddev, cols = is.numeric) %>% head(2)
```


For atomic vectors the function `na_rm` also exists which is 2x faster than `x[!is.na(x)]`. Both `na_omit` and `na_rm` return their argument if no missing cases were found.

The existence of missing cases can be checked using `missing_cases`, which is also considerably faster than `complete.cases` for data frames.

There is also a function `na_insert` to randomly insert missing values into vectors, matrices and data frames. The default is `na_insert(X, prop = 0.1)` so that 10% of values are randomly set to missing.

Finally, a function `allNA` provides the much needed opposite of anyNA for atomic vectors.

## Unique Values / Rows

```{r}
funique(GGDC10S$Variable, sort = T)
```

If all values/rows are unique, the original data is returned

```{r}
identical(funique(GGDC10S), GGDC10S)
```

Remove duplicate rows

```{r}
funique(GGDC10S, cols = .c(Country, Variable)) %>% 
  ss(1:2, 1:8)
```

```{r}
funique(GGDC10S, cols = c("Country", "Variable")) %>% 
  ss(1:2, 1:8)
```

```{r}
funique(GGDC10S, cols = .c(Country, Variable), sort = T) %>% 
  ss(1:2, 1:8)
```

## Recoding and Replacing

- `recode_num`
- `recode_char`
- `replace_NA`
- `replace_Inf`
- `replace_outliers`

```{r}
microbenchmark(replace_NA(GGDC10S, 0))
```

```{r}
add_vars(GGDC10S, 6:16*2-5) <- 
  fselect(GGDC10S, AGR:SUM) %>% 
  lapply(log) %>% 
  replace_Inf() %>% 
  add_stub("log.")
head(GGDC10S, 2)
rm(GGDC10S)
```

```{r}
recode_char(month.name,
            ber = "C", "^J" = "A", default = "B",
            regex = TRUE)
```

`replace_outliers` replaces values falling outside a 1- or 2-sided numeric threshold or outside a certain number of column- standard deviations with a value (default is NA).

Replace all values below 2 and above 100 with NA

```{r}
replace_outliers(mtcars, c(2, 100)) %>% 
  head(3)
```

Replace only values smaller than 2

```{r}
replace_outliers(mtcars, 2, single.limit = "min") %>% 
  head(3)
```

Replace only values larger than 100

```{r}
replace_outliers(mtcars, 100, single.limit = "max") %>% 
  head(3)
```

Replace values above or below 3 column-standard-deviations with NA

```{r}
replace_outliers(mtcars, 3) %>% tail(3)
```

# Data Object Conversions

`collapse` provides a set of functions to speed these conversions

- `qDF` - convert to data.frame
- `qDT` - convert to data.table
- `qTBL` - convert to tibble
- `qM` - converts vectors, higher-dim arrays, data frames, and suitable lists to matrix
- `qF` - convert vector to factor
- `mrtl` - row-wise convert a matrix into list, data.frame or data.table
- `mctl` - column-wise convert a matrix into list, data.frame or data.table

```{r}
str(EuStockMarkets)
```

```{r}
microbenchmark(
  qDT(wlddev),
  qDT(EuStockMarkets),
  data.table::as.data.table(wlddev),
  as.data.frame(EuStockMarkets)
)
```

By default these functions drop all unnecessary attributes from matrices or lists / data frames in the conversion, but this can be changed using the keep.attr = TRUE argument.

`row.names.col` saves names/row names in a column.

```{r}
head(qDT(mtcars, "car"))
```

```{r}
(N_distinct <- fndistinct(GGDC10S))
```

Convert to data frame, preserving names

```{r}
head(qDF(N_distinct, "variable"))
```

This converts the matrix to a list of 1860 row-vectors of length 4.

```{r}
microbenchmark(mrtl(EuStockMarkets))
```

```{r}
mrtl(EuStockMarkets) %>% head()
```

```{r}
microbenchmark(rowSums(qM(mtcars)), 
               rowSums(mtcars),
               kit::psum(mtcars))
```

Factor From Character

```{r}
str(wlddev$country)
fndistinct(wlddev$country)
microbenchmark(
  qF(wlddev$country),
  as.factor(wlddev$country)
)
```

Factor from numeric

```{r}
fndistinct(wlddev$PCGDP)
microbenchmark(
  qF(wlddev$PCGDP),
  as.factor(wlddev$PCGDP)
)
```

# Advanced Statistical Programming

## Fast statistical functions

`fsum`, `fprod`, `fmean`, `fmedian`, `fmode`, `fvar`, `fsd`, `fmin`, `fmax`, `fnth`, `ffirst`, `flast`, `fnobs`, `fndistinct`

For column-wise means

```{r}
fmean(mtcars$mpg)
(f1 <- fmean(mtcars))
```

```{r}
(f2 <- fmean(mtcars, drop = F))
```

```{r}
class(f1); class(f2)
```

```{r}
m <- qM(mtcars)
fmean(m)
fmean(m, drop = F)
```

Weighted statistics

```{r}
weights <- abs(rnorm(fnrow(mtcars)))
fmean(mtcars)
fmean(mtcars, w = weights)
```

Frequency-weighted standard deviation

```{r}
fsd(mtcars, w = weights)
```

Grouped statistics

```{r}
fmean(mtcars, mtcars$cyl)
```

```{r}
fmean(mtcars, fselect(mtcars, cyl, vs, am))
```

Getting column indices

```{r}
(ind <- fselect(mtcars, cyl, vs, am, return = "indices"))
fmean(get_vars(mtcars, -ind), get_vars(mtcars, ind))
```

## Factors, Grouping Objects and Grouped Data Frames

Passing factors or grouping objects to the `g` argument is faster than passing vectors. `na.exclude = FALSE` attaches a class `na.included`

Create a factor. 
```{r}
f <- qF(mtcars$cyl, na.exclude = F)
attributes(f)
```

Grouped standard deviation

```{r}
dat <- get_vars(mtcars, -ind)
fsd(dat, f)
```

Without option na.exclude = FALSE, anyNA needs to be called on the factor (noticeable on larger data).

```{r}
f2 <- qF(mtcars$cyl)
microbenchmark(fsd(dat, f), fsd(dat, f2))
```

Group (GRP) objects are prefereable over factors.

Create a `GRP` object

```{r}
(g <- GRP(mtcars, ~ cyl + vs + am))
str(g)
```

```{r}
plot(g)
```

```{r}
microbenchmark(fsd(dat, g), 
               fsd(dat, get_vars(mtcars, ind)))
```

Or, create grouped data frame.

```{r}
gmtcars <- fgroup_by(mtcars, cyl, vs, am)
fmedian(gmtcars)
head(fgroup_vars(gmtcars))
```

```{r}
fmedian(gmtcars, keep.group_vars = FALSE)
```

Now suppose we wanted to create a new dataset which contains the mean, sd, min and max of the variables mpg and disp grouped by cyl, vs and am:








