---
title: "Fast Data Manipulation"
format: gfm
---

[Source](https://fastverse.org/collapse/articles/collapse_intro.html)

```{r}
library(collapse)
library(magrittr)
```


## Selecting and Replacing Columng

```{r}
wlddev %>% 
  fselect(country, year, PCGDP:ODA) %>% 
  head(2)
```

```{r}
wlddev %>% 
  fselect(-country, -year, -(PCGDP:ODA)) %>% 
  head(2)
```

```{r}
wlddev %>% 
  fselect(-c(country, year, PCGDP:ODA)) %>% 
  head(2)
```

```{r}
library(microbenchmark)
microbenchmark(fselect = collapse::fselect(wlddev, country, year, PCGDP:ODA),
               select = dplyr::select(wlddev, country, year, PCGDP:ODA))
```

`fselect` can also replace

```{r}
fselect(wlddev, PCGDP:POP) <- lapply(fselect(wlddev, PCGDP:POP), log)
head(wlddev, 2)
```

Efficient deletion

```{r}
fselect(wlddev, country, year, PCGDP:POP) <- NULL
head(wlddev, 2)
```

```{r}
rm(wlddev)
```

Other information returned by `fselect`.

```{r}
wlddev %>% 
  fselect(PCGDP:POP, return = "names")
wlddev %>% 
  fselect(PCGDP:POP, return = "indices")
wlddev %>% 
  fselect(PCGDP:POP, return = "named_indices")
wlddev %>% 
  fselect(PCGDP:POP, return = "logical")
wlddev %>% 
  fselect(PCGDP:POP, return = "named_logical")
```

There are no special methods for grouped tibbles.

Select variables with `get_vars`

```{r}
wlddev %>% 
  get_vars(9:13) %>% 
  head(1)
wlddev %>% 
  get_vars( c("PCGDP","LIFEEX","GINI","ODA","POP")) %>% 
  head(1)
wlddev %>% 
  get_vars("[[:upper:]]", regex = T) %>% 
  head(1)
wlddev %>% 
  get_vars("PC|LI|GI|OD|PO", regex = T) %>% 
  head(1)
wlddev %>% 
  get_vars(c("PC","LI","GI","OD","PO"), regex = TRUE) %>% 
  head(1)
wlddev %>% 
  get_vars(is.numeric) %>% 
  head(1)
wlddev %>% 
  get_vars(is.numeric, return = "names") %>% 
  head(1)
```

Can also replace with `get_vars`

```{r}
get_vars(wlddev, 9:13) <- lapply(
  get_vars(wlddev, 9:13), log
)
get_vars(wlddev, 9:13) <- NULL
head(wlddev)
rm(wlddev)
```

collapse offers a set of functions to efficiently select and replace data by data type: `num_vars`, `cat_vars` (for categorical = non-numeric columns), `char_vars`, `fact_vars`, `logi_vars` and `date_vars` (for date and date-time columns).


```{r}
num_vars(wlddev) %>% head(2)
cat_vars(wlddev) %>% head(2)
fact_vars(wlddev) %>% head(2)
```

Replacing

```{r}
fact_vars(wlddev) <- fact_vars(wlddev)
```

## Subsetting

`fsubset` allows multiple comma-separated select arguments after the subset argument, and it also preserves all attributes of subsetted columns:


```{r}
fsubset(GGDC10S, Variable == "VA" & Year > 1990,
        Country, Year, AGR:GOV) %>% head(2)
```

```{r}
fsubset(GGDC10S, Variable == "VA" & Year > 1990,
        -(Regioncode:Variable), -(OTH:SUM)) %>% head(2)
```

```{r}
ss(GGDC10S, 1:2, 6:16)
ss(GGDC10S, -(1:2), c("AGR","MIN")) %>% head(2)
```


```{r}
microbenchmark(
  base = subset(GGDC10S, Variable == "VA" & Year > 1990, AGR:SUM),
  collapse = fsubset(GGDC10S, Variable == "VA" & Year > 1990, AGR:SUM)
)
microbenchmark(GGDC10S[1:10, 1:10], ss(GGDC10S, 1:10, 1:10))
```

## Reordering Rows and Columns

Replaces `arrange`. Negative variable names indicate descending sort.

```{r}
roworder(GGDC10S, -Variable, Country) %>% 
  ss(1:2, 1:8)
```


```{r}
microbenchmark(collapse = collapse::roworder(GGDC10S, -Variable, Country),
               dplyr = dplyr::arrange(GGDC10S, desc(Variable), Country))
```

The function roworderv is a standard evaluation analogue to roworder:

```{r}
roworderv(GGDC10S, c("Variable", "Country"), 
          decreasing = c(TRUE, FALSE)) %>% 
  ss(1:2, 1:8)
```

Move or exchange rows with `roworderv`

```{r}
GGDC10S %>% roworderv(
  neworder = which(GGDC10S$Country == "GHA")
) %>% 
  ss(1:2, 1:8)
GGDC10S %>% roworderv(
  neworder = which(GGDC10S$Country == "BWA"), pos = "end"
) %>% 
  ss(1:2, 1:8)
```

`pos = "exchange"` arranges selected rows in the order they are passed, without affecting other rows

```{r}
GGDC10S %>% roworderv(
  neworder = with(GGDC10S, c(which(Country == "GHA"),
                             which(Country == "BWA"))),
  pos = "exchange"
) %>% 
  ss(1:2, 1:8)
```

## Transforming and Computing New Columns

`ftransform` always returns the entire data frame.

```{r}
GGDC10S %>% 
  ftransform(AGR_perc = AGR / SUM * 100,
             Year = as.integer(Year),
             AGR = NULL) %>% 
  tail(2)
```

```{r}
GGDC10S %>% 
  ftransform(MIN_mean = fmean(MIN),
             Intercept = 1) %>% 
  tail(2)
```

The modification `ftransformv` exists to transform specific columns using a function:


```{r}
GGDC10S %>% 
  ftransformv(6:16, log) %>% 
  tail(2)
GGDC10S %>% 
  ftransformv(6:16, `*`, 100 / SUM) %>% 
  tail(2)
GGDC10S %>% 
  ftransformv(is.numeric, log) %>% 
  tail(2)
```

```{r}
GGDC10S %>%
  ftransform(num_vars(.) %>%
    lapply(log) %>%
    replace_Inf()) %>%
  tail(2)
```

The function settransform(v) can be used to change the input data frame by reference: 


```{r}
settransform(GGDC10S, FIRE_MAN = FIRE / MAN,
                      Regioncode = NULL, Region = NULL)
tail(GGDC10S, 2)
```

```{r}
rm(GGDC10S)
settransformv(GGDC10S, 6:16, `*`, 100 / SUM)
tail(GGDC10S)
```

`fcompute` can be used to compute new columns, returning the computed columns in a new data frame.

```{r}
fcompute(GGDC10S, AGR_perc = AGR / SUM * 100, FIRE_MAN = FIRE / MAN) %>% 
  tail(2)
```

## Adding and Binding Columns

`add_vars`

```{r}
rm(wlddev)
add_vars(wlddev) <- 
  get_vars(wlddev, 9:13) %>% 
  lapply(log10) %>% 
  add_stub("log10.")
head(wlddev, 2)
```

Replace at other positions

```{r}
rm(wlddev)
add_vars(wlddev, "front") <- 
  get_vars(wlddev, 9:13) %>% 
  lapply(log10) %>% 
  add_stub("log10.")
head(wlddev, 2)
```

```{r}
rm(wlddev)

add_vars(wlddev, c(10L, 12L, 14L, 16L, 18L)) <- 
  get_vars(wlddev, 9:13) %>%
  lapply(log10) %>%
  add_stub("log10.")
head(wlddev, 2)
```

Using `add_vars` without replacement.

```{r}
rm(wlddev)
add_vars(
  wlddev, 
  get_vars(wlddev, 9:13) %>% lapply(log) %>% add_stub("log."),
  get_vars(wlddev, 9:13) %>% lapply(log10) %>% add_stub("log10.")
) %>% head(2)
```

```{r}
add_vars(wlddev,
  get_vars(wlddev, 9:13) %>% lapply(log) %>% add_stub("log."),
  get_vars(wlddev, 9:13) %>% lapply(log10) %>% add_stub("log10."),
  pos = c(10L, 13L, 16L, 19L, 22L, 11L, 14L, 17L, 20L, 23L)
) %>% head(2)
```


```{r}
identical(cbind(wlddev, wlddev), add_vars(wlddev, wlddev))

microbenchmark(cbind(wlddev, wlddev), add_vars(wlddev, wlddev))
```

## Renaming Columns

```{r}
GGDC10S %>% 
  frename(AGR = Agriculture, MIN = Mining) %>% 
  head(2)
```

```{r}
frename(GGDC10S, tolower) %>% head(2)

frename(GGDC10S, tolower, cols = .c(AGR, MIN)) %>% head(2)
```

Rename by reference

```{r}
setrename(GGDC10S, AGR = Agriculture, MIN = Mining)
head(GGDC10S, 2)
rm(GGDC10S)
```

## Shortcuts

Only recommended for personal scripts

- `fselect` -> `slt`
- `fsubset` -> `sbt`
- `ftransform(v)` -> `tfm(v)`
- `settransform(v)` -> `settfm(v)`
- `get_vars -> gv`
- `num_vars -> nv`
- `add_vars` -> `av`. 

## Missing Values/Rows

```{r}
microbenchmark(
  na_omit(wlddev, na.attr = TRUE), 
  na.omit(wlddev))
```

Act on certain columns only

```{r}
na_omit(wlddev, cols = .c(PCGDP, LIFEEX)) %>% 
  head(2)
na_omit(wlddev, cols = is.numeric) %>% head(2)
```


For atomic vectors the function `na_rm` also exists which is 2x faster than `x[!is.na(x)]`. Both `na_omit` and `na_rm` return their argument if no missing cases were found.

The existence of missing cases can be checked using `missing_cases`, which is also considerably faster than `complete.cases` for data frames.

There is also a function `na_insert` to randomly insert missing values into vectors, matrices and data frames. The default is `na_insert(X, prop = 0.1)` so that 10% of values are randomly set to missing.

Finally, a function `allNA` provides the much needed opposite of anyNA for atomic vectors.

## Unique Values / Rows

```{r}
funique(GGDC10S$Variable, sort = T)
```

If all values/rows are unique, the original data is returned

```{r}
identical(funique(GGDC10S), GGDC10S)
```

Remove duplicate rows

```{r}
funique(GGDC10S, cols = .c(Country, Variable)) %>% 
  ss(1:2, 1:8)
```

```{r}
funique(GGDC10S, cols = c("Country", "Variable")) %>% 
  ss(1:2, 1:8)
```

```{r}
funique(GGDC10S, cols = .c(Country, Variable), sort = T) %>% 
  ss(1:2, 1:8)
```

## Recoding and Replacing

- `recode_num`
- `recode_char`
- `replace_NA`
- `replace_Inf`
- `replace_outliers`

```{r}
microbenchmark(replace_NA(GGDC10S, 0))
```

```{r}
add_vars(GGDC10S, 6:16*2-5) <- 
  fselect(GGDC10S, AGR:SUM) %>% 
  lapply(log) %>% 
  replace_Inf() %>% 
  add_stub("log.")
head(GGDC10S, 2)
rm(GGDC10S)
```

```{r}
recode_char(month.name,
            ber = "C", "^J" = "A", default = "B",
            regex = TRUE)
```

`replace_outliers` replaces values falling outside a 1- or 2-sided numeric threshold or outside a certain number of column- standard deviations with a value (default is NA).

Replace all values below 2 and above 100 with NA

```{r}
replace_outliers(mtcars, c(2, 100)) %>% 
  head(3)
```

Replace only values smaller than 2

```{r}
replace_outliers(mtcars, 2, single.limit = "min") %>% 
  head(3)
```

Replace only values larger than 100

```{r}
replace_outliers(mtcars, 100, single.limit = "max") %>% 
  head(3)
```

Replace values above or below 3 column-standard-deviations with NA

```{r}
replace_outliers(mtcars, 3) %>% tail(3)
```
