---
title: "Efficient Coding with Collapse"
format: gfm
---


[Source](https://fastverse.org/collapse/articles/developing_with_collapse.html)

```{r}
#| message: false
library(collapse)
library(magrittr)
library(microbenchmark)
options(paged.print = FALSE)
```

# Be Minimalistic in Computations

A key for very efficient code is to use the minimal required operations/objects to get the job done. 

Eg, if a grouping vector `g` is used only once, use internal grouping instead.

```{r}
fmean(mtcars$mpg, mtcars$cyl)
```

```{r}
fmean(mtcars$mpg, mtcars$cyl, TRA = "fill")
```

For aggregation, use unsorted grouping, eg.

`fsum(x, qF(g, sort = FALSE))` or `fsum(x, qG(g, sort = FALSE), use.g.names = FALSE)` if group names aren't needed. `na.exclude = FALSE` should be used with `qF()` and `qG()`.

If `g` is a plain vector of the first-appearance order should be preserved, use `group(g)` instead of `qG(g, sort = FALSE, na.exclude = FALSE)`. Set `use.g.names = FALSE` and `na.rm = FALSE` when possible.

```{r}
x <- rnorm(1e7) # 10 million random obs
g <- sample.int(1e6, 1e7, TRUE) # 1 Million random groups
oldopts <- set_collapse(na.rm = FALSE)
microbenchmark(
  internal = fsum(x, g),
  internal_expand = fsum(x, g, TRA = "fill"),
  qF1 = fsum(x, qF(g, sort = FALSE)),
  qF2 = fsum(x, qF(g, sort = FALSE, na.exclude = FALSE)),
  qG1 = fsum(x, qG(g, sort = FALSE), use = FALSE),
  qG2 = fsum(x, qG(g, sort = FALSE, na.exclude = FALSE), use = FALSE),
  group = fsum(x, group(g), use = FALSE), # Same as above basically
  GRP1 = fsum(x, GRP(g)), 
  GRP2 = fsum(x, GRP(g, sort = FALSE)), 
  GRP3 = fsum(x, GRP(g, sort = FALSE, return.groups = FALSE), use = FALSE)
)
```

Factors and ‘qG’ objects are efficient inputs to all statistical/transformation functions except for
- `fmedian()`, `fnth()`, `fmode()`, `fndistinct()`
- split-apply-combine operations using `BY()`/`gsplit()`

For these, great a `GRP` object. Set `sort = FALSE` and `return.groups = FALSE` if not needed.

```{r}
f <- qF(g); f2 <- qF(g, na.exclude = F)
gg <- group(g) # Same as qG(g, sort = FALSE, na.exclude = FALSE)
grp <- GRP(g)
microbenchmark(
  factor = fsum(x, f),
  factor_nona = fsum(x, f2),
  qG_nona = fsum(x, gg),
  qG_nona_nonam = fsum(x, gg, use = FALSE),
  GRP = fsum(x, grp),
  GRP_nonam = fsum(x, grp, use = FALSE)
)
```

`qF()`/`qG()` are a bit smarter than `group()` when it comes to handling input factors/‘qG’ objects because `group()` hashes every vector:

```{r}
microbenchmark(
  factor_factor = qF(f),
  # This checks NA's and adds 'na.included' class -> full deep copy
  factor_factor2 = qF(f, na.exclude = FALSE), 
  # NA checking costs.. incurred in fsum() and friends
  check_na = collapse:::is.nmfactor(f), 
  check_na2 = collapse:::is.nmfactor(f2),
  factor_qG = qF(gg),
  qG_factor = qG(f),
  qG_qG = qG(gg),
  group_factor = group(f),
  group_qG = group(gg)
)
```

Only in rare cases are grouped/indexed data frames created with fgroup_by()/findex_by() needed in package code. Likewise, functions like fsummarise()/fmutate() are essentially wrappers.

```{r}
mtcars %>% 
  fgroup_by(cyl, vs, am) %>% 
  fsummarise(mpg = fsum(mpg),
             across(c(carb, hp, qsec), fmean))
```

is the same as

```{r}
g <- GRP(mtcars, c("cyl", "vs", "am"))

add_vars(g$groups,
         get_vars(mtcars, "mpg") %>% 
           fsum(g, use = FALSE),
         get_vars(mtcars, c("carb", "hp", "qsec")) %>% 
           fmean(g, use = FALSE))
```

```{r}
microbenchmark(
  fgroup = mtcars %>%
    fgroup_by(cyl, vs, am) %>%
    fsummarise(
      mpg = fsum(mpg),
      across(c(carb, hp, qsec), fmean)
    ),
  addvar = add_vars(
    g$groups,
    get_vars(mtcars, "mpg") %>%
      fsum(g, use = FALSE),
    get_vars(mtcars, c("carb", "hp", "qsec")) %>%
      fmean(g, use = FALSE)
  )
)
```

# Memory

`x[x == 0] <- NA` creates a logical vector of 1 million elements. `setv(x, 0, NA)` is the efficient equivalent.

```{r}
x <- abs(round(rnorm(1e6)))
y <- abs(round(rnorm(1e6)))
microbenchmark(
  lvec = x[x == 0] <- NA,
  coll = setv(y, 0, NA)
)
```



```{r}
x1 <- abs(round(rnorm(1e6)))
x2 <- abs(round(rnorm(1e6)))
y <- rnorm(1e6)
microbenchmark(
  lvec = x1[is.na(x1)] <- y[is.na(x1)],
  coll = setv(x2, NA, y)
)
```

Logical vectors are inefficient compared to indices.

```{r}
xNA <- na_insert(x, prop = 0.4)
xmiss <- is.na(xNA)
ind <- which(xmiss)
microbenchmark(x[xmiss], x[ind])
```

With `collapse`, they can be created directly using `whichNA(xNA)` in this case, or `whichv(x, 0)` for `which(x == 0)` or any other number. Also here there exist an `invert = TRUE` argument covering the `!=` case. For convenience, infix operators `x %==% 0` and `x %!=% 0` wrap `whichv(x, 0)` and `whichv(x, 0, invert = TRUE)`, respectively.

Use %iin% instead of %in%.

```{r}
microbenchmark(
  `%in%` = fsubset(wlddev, iso3c %in% c("USA", "DEU", "ITA", "GBR")),
  `%iin%` = fsubset(wlddev, iso3c %iin% c("USA", "DEU", "ITA", "GBR"))
)
```

`anyNA()`, `allNA()`, `anyv()` and `allv()` are faster than, eg, `any(x == 0)` `na_rm(x)` is faster than `x[!is.na(x)]`.

With `ss()`, index checks can be avoided using `check = FALSE`.

```{r}
ind <- wlddev$iso3c %!iin% c("USA", "DEU", "ITA", "GBR")
microbenchmark::microbenchmark(
  withcheck = ss(wlddev, ind),
  nocheck = ss(wlddev, ind, check = FALSE)
)
```

Inefficiencies can come from copies produced in statistical operations.

```{r}
x <- rnorm(100); y <- rnorm(100); z <- rnorm(100)
```

This produces 2 copies

```{r}
res <- x + y + z
```

Instead, from `kit`, use `psum(x, y, z)` or

```{r}
res <- x + y
res %+=% z
```

For operations between vectors and rows, use `rowwise` argument.

```{r}
m <- qM(mtcars)
setop(m, "*", seq_col(m), rowwise = T)
head(m / qM(mtcars))
```

Some functions like `na_locf()`/`na_focb()` also have set = TRUE arguments to perform operations by reference. There is also `setTRA()` for (grouped) transformations by reference, wrapping `TRA(..., set = TRUE)`.

Replace Sepal.Length with group median.

```{r}
fmedian(iris$Sepal.Length, iris$Species,
        TRA = "fill", set = TRUE)
head(iris)
data(iris)
```

This is the same as

```{r}
setTRA(iris$Sepal.Length, 
       fmedian(iris$Sepal.Length, iris$Species), 
       "fill", iris$Species)
head(iris)
data(iris)
```

```{r}
head(iris)
```

`set` can be invoked anywhere. The following turns columns of the matrix into proportions.

```{r}
fsum(m, TRA = "/", set = TRUE)
fsum(m)
```

Example of an efficient function for univariate linear regression.

```{r}
greg <- function(y, x, g) {
  g <- group(g)
  dmx <- fmean(x, g, TRA = "-", na.rm = FALSE)
  (fsum(y, g, dmx, use = FALSE, na.rm = FALSE) %/=%
      fsum(dmx, g, dmx, use = FALSE, na.rm = FALSE))
}

y <- rnorm(1e7)
x <- rnorm(1e7)
g <- sample.int(1e6, 1e7, TRUE)
microbenchmark(greg(y, x, g), group(g))
```

The expression computed by `greg()` amounts to `sum(y * (x - mean(x)))/sum((x - mean(x))^2)` for each group, which is equivalent to `cov(x, y)/var(x)`, but very efficient, requiring exactly one full copy of x to create a group-demeaned vector, `dmx`, and then using the w (weights) argument to `fsum()` to sum the products (`y * dmx` and `dmx * dmx`) on the fly, including a division by reference avoiding an additional copy. 

# Prefer primitive R objects and functions

Vectors, matrices and lists are good, data frames and complex objects are bad.

```{r}
#| warning: false
l <- unclass(mtcars)
nam <- names(mtcars)
microbenchmark(
  names(mtcars),
  attr(mtcars, "names"),
  names(l),
  names(mtcars) <- nam, 
  attr(mtcars, "names") <- nam, 
  names(l) <- nam,
  mtcars[["mpg"]],
  .subset2(mtcars, "mpg"), 
  l[["mpg"]],
  mtcars[3:8], 
  .subset(mtcars, 3:8), 
  l[3:8],
  ncol(mtcars), 
  length(mtcars), 
  length(unclass(mtcars)), 
  length(l),
  nrow(mtcars), 
  length(.subset2(mtcars, 1L)), 
  length(l[[1L]])
)

```

```{r}
microbenchmark(
  nrow(mtcars),
  length(unclass(mtcars)[[1L]])
)
```

By means of further illustration, let’s recreate the pwnobs() function in collapse which counts pairwise missing values. The list method is written in R. A basic implementation is

```{r}
pwnobs_list <- function(X) {
  dg <- fnobs(X)
  n <- ncol(X)
  nr <- nrow(X)
  N.mat <- diag(dg)
  for (i in 1:(n - 1L)) {
    miss <- is.na(X[[i]])
    for (j in (i + 1L):n) N.mat[i, j] <- N.mat[j, i] <- 
        nr - sum(miss | is.na(X[[j]]))
  }
  rownames(N.mat) <- names(dg)
  colnames(N.mat) <- names(dg)
  N.mat
}
mtcNA <- na_insert(mtcars, prop = 0.2)
pwnobs_list(mtcNA)
```

Optimized version.

```{r}
pwnobs_list_opt <- function(X) {
  dg <- fnobs.data.frame(X)
  class(X) <- NULL
  n <- length(X)
  nr <- length(X[[1L]])
  N.mat <- diag(dg)
  for (i in 1:(n - 1L)) {
    miss <- is.na(X[[i]])
    for (j in (i + 1L):n) N.mat[i, j] <- N.mat[j, i] <- 
        nr - sum(miss | is.na(X[[j]]))
  }
  dimnames(N.mat) <- list(names(dg), names(dg))
  N.mat
}

identical(pwnobs_list(mtcNA), pwnobs_list_opt(mtcNA))
microbenchmark(
  pwnobs_list(mtcNA), pwnobs_list_opt(mtcNA)
)
```

Avoid `as.data.frame` and related.

```{r}
attr(l, "row.names") <- .set_row_names(length(l[[1L]]))
class(l) <- "data.frame"
head(l, 2)
```

```{r}
#| warning: false
library(data.table)
library(tibble)
microbenchmark(qDT(mtcars), as.data.table(mtcars),
                               qTBL(mtcars), as_tibble(mtcars))

```

```{r}
l <- unclass(mtcars)
microbenchmark(qDF(l), as.data.frame(l), 
               as.data.table(l), as_tibble(l))
```

Another efficient workflow for general data frame-like objects is to save the attributes `ax <- attributes(data)`, manipulate it as a list `attributes(data) <- NULL`, modify `ax$names` and `ax$row.names` as needed and then use `setattrib(data, ax)` before returning.



