---
title: "Collapse Advanced Data Aggregation"
format: gfm
---


[Source](https://fastverse.org/collapse/articles/collapse_intro.html)

```{r}
#| message: false
library(collapse)
library(magrittr)
library(microbenchmark)
options(paged.print = FALSE)
```

`collap`

```{r}
weights <- abs(rnorm(fnrow(mtcars)))
collap(mtcars,
       mpg + disp ~ cyl + vs + am,
       list(fmean, fsd, fmin, fmax),
       w = weights, keep.col.order = F)
```

Suppose we would like to aggregate `wlddev` data by country and decade, but keep all that categorical information. 

```{r}
collap(wlddev, ~ iso3c + decade) %>% head()
```

`collap` syntax.

```{r}
#| eval: false
collap(X, by, 
       FUN = fmean, catFUN = fmode, 
       cols = NULL, 
       w = NULL, wFUN = fsum,
       custom = NULL, 
       keep.by = TRUE, keep.w = TRUE, keep.col.order = TRUE,
       sort.row = TRUE, 
       parallel = FALSE, mc.cores = 1L,
       return = c("wide","list","long","long_dupl"), 
       give.names = "auto") # , ...
```

Suppose we only want to aggregate 4 series in this dataset.

```{r}
collap(wlddev,
       PCGDP + LIFEEX + GINI + ODA ~ iso3c + decade) %>% 
  head()
collap(wlddev, 
       ~ iso3c + decade, cols = 9:12) %>% 
  head()
```

Using multiple functions

```{r}
collap(wlddev,
       ~ iso3c + decade,
       list(fmean, fmedian, fsd),
       cols = 9:12) %>% 
  head()
```

Return long-format data

```{r}
collap(wlddev,
       ~ iso3c + decade,
       list(fmean, fmedian, fsd),
       cols = 9:12,
       return = "long") %>% 
  head()
```

The `custom` argument allows the user to circumvent the broad distinction into numeric and categorical data (and the associated `FUN` and `catFUN` arguments) and specify exactly which columns to aggregate using which functions.

```{r}
collap(wlddev,
       ~ iso3c + decade,
       custom = list(
         fmean = 9:10, fmedian = 11:12,
         ffirst = c("country", "region", "income"),
         flast = c("year", "date"),
         fmode = "OECD"
       )) %>% 
  head()
```

It is also possible to perform weighted aggregations and append functions with _uw to yield an unweighted computation.

```{r}
collap(wlddev,
       ~ region + year,
       w = ~ POP,
       custom = list(
         fmean = 9:10, fmedian_uw = 11:12,
         ffirst_uw = c("country", "region", "income"),
         flast_uw = c("year", "date"),
         fmode = "OECD"
       ),
       keep.w = FALSE) %>% 
  head()
```









