---
title: "Advanced and Fast Statisctical Computing"
format: gfm
---

[Source](https://www.youtube.com/watch?v=OwWT1-dSEts&t=6433s)

```{r}
library(collapse)
library(magrittr)
options(paged.print = FALSE)
```

```{r}
.FAST_STAT_FUN
```

```{r}
methods(fmean)
```

```{r}
class(AirPassengers)
fmean(AirPassengers)
```

```{r}
class(EuStockMarkets)
fmean(EuStockMarkets)
```

```{r}
class(airquality)
fmean(airquality)
```

```{r}
library(microbenchmark)
x <- rnorm(1e7)
microbenchmark(
  mean(x), fmean(x), fmean(x, nthreads = 4)
)
```

```{r}
microbenchmark(colMeans(EuStockMarkets), fmean(EuStockMarkets))
```

```{r}
microbenchmark(sapply(mtcars, mean), fmean(mtcars))
```

```{r}
fmean(mtcars)
```

```{r}
microbenchmark(
  base = sapply(GGDC10S[6:16], mean, na.rm = TRUE),
  collapse = fmean(GGDC10S[6:16])
)
```

Weighted mean

```{r}
w <- abs(rnorm(nrow(iris)))
all.equal(
  fmean(num_vars(iris), w = w),
  sapply(num_vars(iris), weighted.mean, w = w)
)
microbenchmark(
  coll = fmean(num_vars(iris), w = w),
  base = sapply(num_vars(iris), weighted.mean, w = w)
)
```

Missing weights are treated as 0, ie observation omitted.

```{r}
wNA <- na_insert(w, prop = 0.05)
fmean(num_vars(iris), w = wNA)
```

Grouped statistics

```{r}
microbenchmark(
  fmean = fmean(iris$Sepal.Length, iris$Species),
  tapply = tapply(iris$Sepal.Length, iris$Species, mean)
)
```

```{r}
library(magrittr)
microbenchmark(
  fmean = fmean(num_vars(iris), iris$Species),
  fgroup = iris %>% fgroup_by(Species) %>% fmean,
  dplyr = iris %>% dplyr::group_by(Species) %>% fmean
)
```

# Weighted group means

```{r}
fmean(num_vars(iris), iris$Species, w)
```

```{r}
iris %>% 
  add_vars(w) %>% 
  fgroup_by(Species) %>% 
  fmean
```

```{r}
microbenchmark(
  base = lapply(num_vars(iris), weighted.mean, w),
  clp_w = fmean(num_vars(iris), w = w),
  clp_g = fmean(num_vars(iris), iris$Species),
  clp_g_w = fmean(num_vars(iris), iris$Species, w)
)
```

Benchmark with 10M observations across 1M groups

```{r}
#| message: false
library(data.table)
g <- sample.int(1e6, 1e7, replace = T)
w <- abs(rnorm(1e7))
dt <- setkey(data.table(x, g))
g <- GRP(g)

microbenchmark(
  clp = fmean(x, na.rm = F),
  clp_g = fmean(x, g, use.g.names = F, na.rm = F),
  clp_t_w = fmean(x, g, w, use.g.names = F, na.rm = F),
  clp_t_w_th = fmean(x, g, w, use.g.names = F, na.rm = F, nthreads = 4),
  dt = dt[, mean(x), keyby = g]
)
```

Grouped weighted centering. These are equivalent.

```{r}
#| eval: false
fmean(x, g, w, "-")
TRA(x, fmean(x, g, w), "-", g)
```


```{r}
microbenchmark(
  base = iris$Sepal.Length - ave(iris$Sepal.Length, iris$Species),
  fmean = fmean(iris$Sepal.Length, iris$Species, TRA = "-"),
  fmean_TRA = TRA(iris$Sepal.Length, fmean(iris$Sepal.Length, iris$Species),
                  "-", iris$Species)
)
```

Modification in-place

```{r}
iris_cp1 <- iris
iris_cp2 <- iris
fmean(iris_cp1$Sepal.Length, iris_cp1$Species, TRA = "-", set = TRUE)
setTRA(iris_cp2$Sepal.Length, fmean(iris_cp2$Sepal.Length, iris_cp2$Species),
       "-", iris_cp2$Species)
all_obj_equal(iris_cp1, iris_cp2)
iris_cp1 %>% head
```

```{r}
fmean(iris$Sepal.Length, iris$Species, TRA = "-") %>% head
```

Impute missing variables by reference

```{r}
num_vars(iris) %<>% na_insert(prop = 0.05)
num_vars(iris) %>% head
num_vars(iris) %>% 
  fmedian(iris$Species, TRA = "replace_NA", set = TRUE)
num_vars(iris) %>% head
```

Different grouped/weighted transformations at once.

```{r}
mtcars %>% 
  ftransform(
    A = fsum(mpg, TRA = "%"),
    B = mpg %>% fmedian(mpg, cyl, TRA = "replace_fill"),
    C = fmedian(mpg, list(vs, am), wt, "-"),
    D = fmean(mpg, vs,, 1L) > fmean(mpg, am,, 1L) # 1L is "replace_fill"
  ) %>% head
```

Row and column arithmetic

```{r}
mtcars %c/% mtcars %>% head(2)
```

Sweep median centered values

```{r}
mtcars %r-% fmean(mtcars) %>% head(3)
```

# Grouping objects

Main function `GRP()`, returns a list-like object. Allows for 0 cost grouped computations

```{r}
(g <- GRP(iris, by = ~ Species))
```

```{r}
str(g)
```


```{r}
fmean(num_vars(iris), g)
```

```{r}
ffirst(num_vars(iris), g, na.rm = F)
```

```{r}

```

# Efficient Inputs: Factors

```{r}
x <- na_insert(rnorm(1e7), prop = 0.01)
g <- sample.int(1e6, 1e7, TRUE)
system.time(gg <- GRP(g))
```

```{r}
system.time(f <- factor(g, exclude = NULL))
system.time(f <- qF(g, na.exclude = F))
```

```{r}
class(f)
```

The `na.included` class indicates no missing values.

```{r}
f2 <- `class<-`(f, "factor")
class(f2)
```


```{r}
microbenchmark(
  fmean(x, g),
  fmean(x, gg),
  fmean(x, gg, na.rm = FALSE),
  fmean(x, f),
  fmean(x, f2),
  ffirst(x, gg, na.rm = FALSE)
)
```

Depending on the specific caluclation, the group object can be faster.

```{r}
microbenchmark(
  fmedian(x, gg),
  fmedian(x, f)
)
```

Collapse provides a `qG` object with is a "light" factor class, or quick group.

```{r}
system.time(qg <- qG(g, na.exclude = F))
str(qg)
```

```{r}
fmean(x, qg)
```





