---
title: "Collapse Time Series and Panel Series"
format: gfm
---


[Source](https://fastverse.org/collapse/articles/collapse_intro.html#time-series-and-panel-series)

```{r}
#| message: false
library(collapse)
library(magrittr)
library(microbenchmark)
options(paged.print = FALSE)
```

# Panel Series to Array Conversions

To facilitate the exploration and access of panel data, `psmat` was created as an S3 generic to efficiently obtain matrices or 3D-arrays from panel data.

```{r}
mts <- psmat(wlddev, PCGDP ~ iso3c, ~ year)
str(mts)
```

```{r}
plot(log10(mts), main = paste("Log10", vlabels(wlddev$PCGDP)))
```

Passing a data frame of panel series to psmat generates a 3D array:

```{r}
psar <- psmat(wlddev, ~ iso3c, ~ year, cols = 9:12)
str(psar)
```

```{r}
plot(psar)
```

Plot array of Panel Series aggregated by region

```{r}
collap(wlddev, ~ region + year, cols = 9:12) %>% 
  psmat(~ region, ~ year) %>% 
  plot(legend = TRUE, labs = vlabels(wlddev)[9:12])
```

`psmat` can also output a list of panel series matrices, which can be used among other things to reshape the data with `unlist2d`.

A list of ps-matrices

```{r}
psml <- psmat(wlddev, ~ iso3c, ~ year, 9:12, array = FALSE)
str(psml, give.attr = F)
```

Generate a data frame from the list

```{r}
unlist2d(psml, idcols = "Variable", row.names = "Country") %>% 
  get_vars(1:10) %>% 
  head
```

# ACF, PACF and CCF

The correlation structure of panel data can also be explored with `psacf`, `pspacf` and `psccf`. They use `fscale` to group-scale panel data by the panel-id provided, and then compute the covariance of a sequence of panel-lags (generated with `flag` discussed below) with the group-scaled level-series, dividing by the variance of the group-scaled level series. The Partial-ACF is generated from the ACF using a Yule-Walker decomposition (as in `stats::pacf`).

Panel-ACF of GCP per Capita

```{r}
psacf(wlddev, PCGDP ~ iso3c, ~ year)
```

Panel-Partial-ACF

```{r}
pspacf(wlddev, PCGDP ~ iso3c, ~ year)
```

Panel- Cross-Correlation function of GDP and Life-Expectancy

```{r}
wlddev %$% psccf(PCGDP, LIFEEX, iso3c, year)
```

Multivariate Panel-auto and cross-correlation function of 3 variables:

```{r}
psacf(wlddev, PCGDP + LIFEEX + ODA ~ iso3c, ~ year)
```

# Fast Lags and Leads

`flag`, `L`, `F`

1 lag

```{r}
L(AirPassengers)
```









