---
title: "Collapse Advanced Statistical Programming"
format: gfm
---


[Source](https://fastverse.org/collapse/articles/collapse_intro.html)

```{r}
library(collapse)
library(magrittr)
library(microbenchmark)
```

# Fast statistical functions

`fsum`, `fprod`, `fmean`, `fmedian`, `fmode`, `fvar`, `fsd`, `fmin`, `fmax`, `fnth`, `ffirst`, `flast`, `fnobs`, `fndistinct`

For column-wise means

```{r}
fmean(mtcars$mpg)
(f1 <- fmean(mtcars))
```

```{r}
(f2 <- fmean(mtcars, drop = F))
```

```{r}
class(f1); class(f2)
```

```{r}
m <- qM(mtcars)
fmean(m)
fmean(m, drop = F)
```

Weighted statistics

```{r}
weights <- abs(rnorm(fnrow(mtcars)))
fmean(mtcars)
fmean(mtcars, w = weights)
```

Frequency-weighted standard deviation

```{r}
fsd(mtcars, w = weights)
```

Grouped statistics

```{r}
fmean(mtcars, mtcars$cyl)
```

```{r}
fmean(mtcars, fselect(mtcars, cyl, vs, am))
```

Getting column indices

```{r}
(ind <- fselect(mtcars, cyl, vs, am, return = "indices"))
fmean(get_vars(mtcars, -ind), get_vars(mtcars, ind))
```

# Factors, Grouping Objects and Grouped Data Frames

Passing factors or grouping objects to the `g` argument is faster than passing vectors. `na.exclude = FALSE` attaches a class `na.included`

Create a factor. 
```{r}
f <- qF(mtcars$cyl, na.exclude = F)
attributes(f)
```

Grouped standard deviation

```{r}
dat <- get_vars(mtcars, -ind)
fsd(dat, f)
```

Without option na.exclude = FALSE, anyNA needs to be called on the factor (noticeable on larger data).

```{r}
f2 <- qF(mtcars$cyl)
microbenchmark(fsd(dat, f), fsd(dat, f2))
```

Group (GRP) objects are prefereable over factors.

Create a `GRP` object

```{r}
(g <- GRP(mtcars, ~ cyl + vs + am))
str(g)
```

```{r collapse-adv-stat-prog-1}
plot(g)
```

```{r}
microbenchmark(fsd(dat, g), 
               fsd(dat, get_vars(mtcars, ind)))
```

Or, create grouped data frame.

```{r}
gmtcars <- fgroup_by(mtcars, cyl, vs, am)
fmedian(gmtcars)
head(fgroup_vars(gmtcars))
```

```{r}
fmedian(gmtcars, keep.group_vars = FALSE)
```

Now suppose we wanted to create a new dataset which contains the mean, sd, min and max of the variables mpg and disp grouped by cyl, vs and am:

Standard evaluation
```{r}
dat <- get_vars(mtcars, c("mpg", "disp"))
add_vars(g[["groups"]],
         add_stub(fmean(dat, g, use.g.names = F), "mean_"),
         add_stub(fsd(dat, g, use.g.names = F), "sd_"),
         add_stub(fmin(dat, g, use.g.names = F), "min_"),
         add_stub(fmax(dat, g, use.g.names = F), "max_"))
```

Non-standard evaluation

```{r}
fgroup_by(mtcars, cyl, vs, am) %>% 
  fselect(mpg, disp) %>% {
    add_vars(fgroup_vars(., "unique"),
             fmean(., keep.group_vars = F) %>% add_stub("mean_"),
             fsd(., keep.group_vars = F) %>% add_stub("sd_"),
             fmin(., keep.group_vars = F) %>% add_stub("min_"),
             fmax(., keep.group_vars = F) %>% add_stub("max_"))
  }
```

# Grouped and Weighted Computations

Using a weight vector

```{r}
add_vars(g[["groups"]],
         add_stub(fmean(dat, g, weights, use.g.names = F), "w_mean_"),
         add_stub(fsd(dat, g, weights, use.g.names = F), "w_sd_"),
         add_stub(fmin(dat, g, use.g.names = F), "min_"),
         add_stub(fmax(dat, g, use.g.names = F), "max_"))
```

With reordering

```{r}
add_vars(g[["groups"]],
         add_stub(fmean(dat, g, weights, use.g.names = F), "w_mean_"),
         add_stub(fsd(dat, g, weights, use.g.names = F), "w_sd_"),
         add_stub(fmin(dat, g, use.g.names = F), "min_"),
         add_stub(fmax(dat, g, use.g.names = F), "max_"),
         pos = c(4,8,5,9,6,10,7,11))
```

```{r}
microbenchmark(call = add_vars(g[["groups"]],
         add_stub(fmean(dat, g, weights, use.g.names = FALSE), "w_mean_"),
         add_stub(fsd(dat, g, weights, use.g.names = FALSE), "w_sd_"),
         add_stub(fmin(dat, g, use.g.names = FALSE), "min_"),
         add_stub(fmax(dat, g, use.g.names = FALSE), "max_")))
```

# Transformations with `TRA`



- calculate weighted group means and use them to demean the data
- calculate weighted group sd's and use them to scale the data
- replace all observations by their group-minimum
- replace all observations by their group-maximum

```{r}
add_vars(get_vars(mtcars, ind),
         add_stub(fmean(dat, g, weights, "-"), "w_demean_"),
         add_stub(fsd(dat, g, weights, "/"), "w_scalse_"),
         add_stub(fmin(dat, g, "replace"), "min_"),
         add_stub(fmax(dat, g, "replace"), "max_")) %>% 
  head()
```

Modify `mtcars`

```{r}
pos <- as.integer(c(2,8,3,9,4,10,5,11))

add_vars(mtcars, pos) <- c(
  add_stub(fmean(dat, g, weights, "-"), "w_demean_"),
  add_stub(fsd(dat, g, weights, "/"), "w_scale_"),
  add_stub(fmin(dat, g, "replace"), "min_"),
  add_stub(fmax(dat, g, "replace"), "max_")
)
head(mtcars)
rm(mtcars)
```

Using `ftransform`

```{r}
settransform(mtcars,
             carb_dwmed_cyl = fmedian(carb, cyl, weights, "-"),
             carb_wsd_vs_am = fsd(carb, list(vs, am), weights, "replace"))


rm(mtcars)
```

Multivariate

```{r}
settransform(mtcars, c(
  fmedian(list(carb_dwmed_cyl = carb, mpg_dwmed_cyl = mpg), 
          cyl, weights, "-"),
  fsd(list(carb_wsd_vs_am = carb, mpg_wsd_vs_am = mpg), 
      list(vs, am), weights, "replace")
))
```

Nested (Computing the weighted 3rd quartile of mpg, grouped by cyl and carb being greater than it's weighted median, grouped by vs)

```{r}
settransform(mtcars,
  mpg_gwQ3_cyl = fnth(
    mpg, 0.75,
    list(cyl, carb > fmedian(carb, vs, weights, 1L)),
    weights, 1L
  )
)

head(mtcars)
rm(mtcars)
```


