---
title: "Collapse Advanced Statistical Programming"
format: gfm
---

[Source](https://fastverse.org/collapse/articles/collapse_intro.html)

```{r}
library(collapse)
library(magrittr)
```

## Fast statistical functions

`fsum`, `fprod`, `fmean`, `fmedian`, `fmode`, `fvar`, `fsd`, `fmin`, `fmax`, `fnth`, `ffirst`, `flast`, `fnobs`, `fndistinct`

For column-wise means

```{r}
fmean(mtcars$mpg)
(f1 <- fmean(mtcars))
```

```{r}
(f2 <- fmean(mtcars, drop = F))
```

```{r}
class(f1); class(f2)
```

```{r}
m <- qM(mtcars)
fmean(m)
fmean(m, drop = F)
```

Weighted statistics

```{r}
weights <- abs(rnorm(fnrow(mtcars)))
fmean(mtcars)
fmean(mtcars, w = weights)
```

Frequency-weighted standard deviation

```{r}
fsd(mtcars, w = weights)
```

Grouped statistics

```{r}
fmean(mtcars, mtcars$cyl)
```

```{r}
fmean(mtcars, fselect(mtcars, cyl, vs, am))
```

Getting column indices

```{r}
(ind <- fselect(mtcars, cyl, vs, am, return = "indices"))
fmean(get_vars(mtcars, -ind), get_vars(mtcars, ind))
```

## Factors, Grouping Objects and Grouped Data Frames

Passing factors or grouping objects to the `g` argument is faster than passing vectors. `na.exclude = FALSE` attaches a class `na.included`

Create a factor. 
```{r}
f <- qF(mtcars$cyl, na.exclude = F)
attributes(f)
```

Grouped standard deviation

```{r}
dat <- get_vars(mtcars, -ind)
fsd(dat, f)
```

Without option na.exclude = FALSE, anyNA needs to be called on the factor (noticeable on larger data).

```{r}
f2 <- qF(mtcars$cyl)
microbenchmark(fsd(dat, f), fsd(dat, f2))
```

Group (GRP) objects are prefereable over factors.

Create a `GRP` object

```{r}
(g <- GRP(mtcars, ~ cyl + vs + am))
str(g)
```

```{r}
plot(g)
```

```{r}
microbenchmark(fsd(dat, g), 
               fsd(dat, get_vars(mtcars, ind)))
```

Or, create grouped data frame.

```{r}
gmtcars <- fgroup_by(mtcars, cyl, vs, am)
fmedian(gmtcars)
head(fgroup_vars(gmtcars))
```

```{r}
fmedian(gmtcars, keep.group_vars = FALSE)
```

Now suppose we wanted to create a new dataset which contains the mean, sd, min and max of the variables mpg and disp grouped by cyl, vs and am:








