---
title: "Collapse and dyplr"
format: gfm
---


[Source](https://fastverse.org/collapse/articles/collapse_intro.html)

```{r}
#| message: false
library(collapse)
library(magrittr)
library(dplyr)
library(microbenchmark)
options(paged.print = FALSE)
```

# Fast Aggregations

- `fsum`
- `fprod`
- `fmean`
- `fmedian`
- `fmode`
- `fvar`
- `fsd`
- `fmin`
- `fmax`
- `fnth`
- `ffirst`
- `flast`
- `fnobs` - Number of Observations
- `fndistinct` - Number of distinct values

The grouped tibble method has the following arguments:

```
#| eval: false
FUN.grouped_df(x, [w = NULL,] TRA = NULL, [na.rm = TRUE,]
               use.g.names = FALSE, keep.group_vars = TRUE, 
               [keep.w = TRUE,] ...)
```

## Simple Aggregations


Efficient conversion to `tibble`

```{r}
GGDC10S <- qTBL(GGDC10S)
```

Column-wise computations

Number of Observations

```{r}
GGDC10S %>% fnobs
```

Number of distinct values

```{r}
GGDC10S %>% fndistinct
```



```{r}
GGDC10S %>% select_at(6:16) %>% fmedian
```

```{r}
GGDC10S %>% fmode
```

Keep data structure intact

```{r}
GGDC10S %>% fmode(drop = FALSE)
```

Moving on to grouped statistics, we can compute the average value added and employment by sector and country using:

```{r}
GGDC10S %>% 
  group_by(Variable, Country) %>%
  select_at(6:16) %>% fmean
```

It is important to not use `dplyr`’s `summarize` together with these functions since that would eliminate their speed gain. These functions are fast because they are executed only once and carry out the grouped computations in C++, whereas summarize will apply the function to each group in the grouped tibble.

The method GRP.grouped_df takes a dplyr grouping object from a grouped tibble and efficiently converts it to a collapse grouping object:

```{r}
GGDC10S %>% 
  group_by(Variable, Country) %>% 
  GRP %>% str
```

## Using `collapse` Verbs

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  get_vars(6:16) %>% 
  fmedian
```

```{r}
microbenchmark(
  collapse = GGDC10S %>% fgroup_by(Variable, Country) %>%
    get_vars(6:16) %>% fmedian(),
  hybrid = GGDC10S %>% group_by(Variable, Country) %>%
    select_at(6:16) %>%
    fmedian(),
  dplyr = GGDC10S %>% group_by(Variable, Country) %>%
    select_at(6:16) %>%
    summarise_all(median, na.rm = TRUE)
)
```

```{r}
class(group_by(GGDC10S, Variable, Country))
class(fgroup_by(GGDC10S, Variable, Country))
```

```{r}
fgroup_by(GGDC10S, Variable, Country)
```

Note further that fselect and get_vars are not full drop-in replacements for select because they do not have a grouped_df method:

```{r}
GGDC10S %>% group_by(Variable, Country) %>% select_at(6:16) %>% tail(3)
GGDC10S %>% group_by(Variable, Country) %>% get_vars(6:16) %>% tail(3)
```

Since by default keep.group_vars = TRUE in the Fast Statistical Functions, the end result is nevertheless the same:

```{r}
GGDC10S %>% group_by(Variable, Country) %>% 
  select_at(6:16) %>% fmean %>% tail(3)
GGDC10S %>% group_by(Variable, Country) %>% 
  get_vars(6:16) %>% fmean %>% tail(3)
```

`fgroup_vars` can be used to efficiently obtain the grouping columns or grouping variables from a grouped tibble.

`fgroup_by` fully supports grouped tibbles created with `group_by` or `fgroup_by`: 

```{r}
GGDC10S %>% group_by(Variable, Country) %>% 
  fgroup_vars %>% 
  head(3)
GGDC10S %>% fgroup_by(Variable, Country) %>% 
  fgroup_vars %>% 
  head(3)
```

```{r}
GGDC10S %>% group_by(Variable, Country) %>% 
  fgroup_vars("unique") %>% 
  head(3)
```

`fsubset` is a faster alternative to `dplyr::filter` which also provides an option to flexibly subset columns after the select argument:

Two equivalent calls, the first is substantially faster


```{r}
microbenchmark(
  filter = GGDC10S %>% 
    filter(Variable == "VA" & Year > 1990) %>%
    select(Country, Year, AGR:GOV) %>% head(3),
  fsubset = GGDC10S %>% 
    fsubset(Variable == "VA" & Year > 1990, 
            Country, Year, AGR:GOV) %>% head(3)
)
```

## Multifunction Aggregations

For such operations it is often necessary to use curly braces `{` to prevent first argument injection so that `%>% cbind(FUN1(.), FUN2(.))` does not evaluate as `%>% cbind(., FUN1(.), FUN2(.))`

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  get_vars(6:16) %>%  {
    cbind(fmedian(.),
          add_stub(fmean(., keep.group_vars = F), "mean_"))
  } %>% head(3)
```
 # Regular expression matching column names
```{r}
GGDC10S %>%
  fgroup_by(Variable, Country) %>%
  {
    add_vars(
      get_vars(., "Reg", regex = TRUE) %>% # Gets both Region and Regioncode
        ffirst(),
      num_vars(.) %>%
        fmean(keep.group_vars = FALSE) %>%
        add_stub("mean_"),
      fselect(., PU:TRA) %>%
        fmedian(keep.group_vars = FALSE) %>%
        add_stub("median_"),
      fselect(., PU:CON) %>%
        fmin(keep.group_vars = FALSE) %>%
        add_stub("min_")
    )
  } %>%
  head(3)
```

Specify position of new columns

```{r}
names(GGDC10S)
```

```{r}
GGDC10S %>%
  fsubset(Variable == "VA", Country, AGR, SUM) %>% 
  fgroup_by(Country) %>% {
   add_vars(fgroup_vars(.,"unique"),
            fmean(., keep.group_vars = FALSE) %>% add_stub("mean_"),
            fsd(., keep.group_vars = FALSE) %>% add_stub("sd_"), 
            pos = c(2,4,3,5))
  } %>% head(3)
```

```{r}
GGDC10S %>%
  fsubset(Variable == "VA", Country, AGR, SUM) %>% 
  fgroup_by(Country) %>% {
   add_vars(fgroup_vars(.,"unique"),
            fmean(., keep.group_vars = FALSE) %>% add_stub("mean_"),
            fsd(., keep.group_vars = FALSE) %>% add_stub("sd_"))
  } %>% head(3)
```

A much more compact solution to multi-function and multi-type aggregation is offered by the function collapg, which by default uses `fmean` for numeric and `fmode` for categorical columns.

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  collapg %>% 
  head(3)
```

Changing the default

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  collapg(fmedian, flast) %>% head(3)
```

One can apply multiple functions to both numeric and/or categorical data:

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  collapg(list(fmean, fmedian), list(ffirst, fmode, flast)) %>% 
  head(3)
```

Applying multiple functions to only numeric (or only categorical) data allows return in a long format:

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>%
  collapg(list(fmean, fmedian), 
          cols = is.numeric, return = "long") %>% 
  head(3)
```

Apply aggregator functions to select columns

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  collapg(
    custom = list(fmean = 6:8,
                  fmedian = 10:12)
  ) %>% head(3)
```

## Weighted Aggregations

```{r}
GGDC10S %>%
  fgroup_by(Variable, Country) %>%
  fselect(AGR:SUM) %>% 
  fsd(SUM) %>% head(3)
GGDC10S %>%
  fgroup_by(Variable, Country) %>%
  fselect(AGR:SUM) %>% 
  fmode(SUM) %>% head(3)
```

Weighted aggregations may also be performed with collapg. By default fsum is used to compute a sum of the weights, but it is also possible here to aggregate the weights with other functions:

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  collapg(w = SUM, wFUN = list(fsum, fmax)) %>%  head(3)
```

# Fast Transformations

`ftransform` instead of `mutate`.

```{r}
GGDC10S %>% 
  fsubset(Variable == "VA", Country, Year, AGR, SUM) %>% 
  ftransform(
    AGR_perc = AGR / SUM * 100,
    AGR_mean = fmean(AGR),
    AGR = NULL, SUM = NULL # Deleting columns AGR and SUM
  ) %>% head
```

```{r}
microbenchmark(
  ftransform = GGDC10S %>%
    fsubset(Variable == "VA", Country, Year, AGR, SUM) %>%
    ftransform(
      AGR_perc = AGR / SUM * 100,
      AGR_mean = fmean(AGR),
      AGR = NULL, SUM = NULL # Deleting columns AGR and SUM
    ) %>% head(),
  mutate = GGDC10S %>%
    filter(Variable == "VA") %>%
    select(Country, Year, AGR, SUM) %>%
    mutate(
      AGR_perc = AGR / SUM * 100,
      AGR_mean = mean(AGR)
    ) %>%
    select(-AGR, -SUM) %>%
    head()
)
```

The modification brought by `ftransformv` enables transformations of groups of columns like `dplyr::mutate_at` and `dplyr::mutate_if`:


This replaces variables mpg, carb and wt by their log (.c turns expressions into character vectors)

```{r}
mtcars %>% 
  ftransformv(.c(mpg, carb, wt), log) %>% 
  head
```

Logging numeric variables

```{r}
iris %>% ftransformv(is.numeric, log) %>% head
```

Can pass a list of transformed variables instead of using `column = value`.

```{r}
mtcars %>% 
  ftransform(
    fselect(., mpg, cyl, vs:gear) %>% 
      lapply(log) %>% replace_Inf
  ) %>% 
  head
```

If only the computed columns need to be returned, `fcompute` provides an efficient alternative:

```{r}
GGDC10S %>% 
  fsubset(Variable == "VA", Country, Year, AGR, SUM) %>%
  fcompute(AGR_perc = AGR / SUM * 100,
           AGR_mean = fmean(AGR)) %>% head
```

## Replacing and sweeping out statistics

- “replace_fill” : replace and overwrite missing values (same as mutate)
- “replace” : replace but preserve missing values
- “-” : subtract (center)
- “-+” : subtract group-statistics but add average of group statistics
- “/” : divide (scale)
- “%” : compute percentages (divide and multiply by 100)
- “+” : add
- “*” : multiply
- “%%” : modulus
- “-%%” : subtract modulus

This subtracts the median value from all data points i.e. centers on the median

```{r}
GGDC10S %>% 
  num_vars %>% 
  fmedian(TRA = "-") %>% head()
```

This replaces all data points with the mode

```{r}
GGDC10S %>% 
  char_vars %>% 
  fmode(TRA = "replace") %>% head
```


For grouped transformations:

Replacing data with the 2nd quartile (25%)

```{r}
GGDC10S %>% 
  fselect(Variable, Country, AGR:SUM) %>% 
  fgroup_by(Variable, Country) %>% 
  fnth(0.25, TRA = "replace_fill") %>% head(3)
```

Scaling sectoral data by Variable and Country

```{r}
GGDC10S %>% 
  fselect(Variable, Country, AGR:SUM) %>% 
  fgroup_by(Variable, Country) %>% 
  fsd(TRA = "/") %>% head
```

Grouped mutations without creating a grouped tibble

AGR_gmed = TRUE if AGR is greater than it's median value, grouped by Variable and Country. Note: This calls fmedian.default

```{r}
settransform(GGDC10S,
             AGR_gmed = AGR > fmedian(AGR, list(Variable, Country),
                                      TRA = "replace")
)
tail(GGDC10S, 3)
```

Dividing (scaling) the sectoral data (columns 6 through 16) by their grouped standard deviation

```{r}
settransformv(
  GGDC10S,
  6:16,
  fsd,
  list(Variable, Country),
  TRA = "/",
  apply = FALSE
)
tail(GGDC10S)
rm(GGDC10S)
```

Adding weights to grouped transformations

```{r}
GGDC10S %>% 
  fselect(Variable, Country, AGR:SUM) %>% 
  fgroup_by(Variable, Country) %>% 
  fmean(SUM, "-") %>% 
  head
```

Sequentially scale then subtract median.

```{r}
GGDC10S %>% 
  fselect(Variable, Country, AGR:SUM) %>% 
  fgroup_by(Variable, Country) %>% 
  fsd(TRA = "/") %>% 
  fmedian(TRA = "-") %>% 
  head(10)
```

Of course it is also possible to combine multiple functions as in the aggregation section, or to add variables to existing data:

This adds a groupwise observation count next to each column

```{r}
add_vars(GGDC10S, seq(7, 27, 2)) <- GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  fselect(AGR:SUM) %>% 
  fnobs("replace_fill") %>% 
  add_stub("N_")
head(GGDC10S)
rm(GGDC10S)
```

## The `TRA` function

Fundamentally, `TRA` is a generalization of `base::sweep` for column-wise grouped operations. Direct calls to `TRA` enable more control over inputs and outputs.

The two operations below are equivalent, although the first is slightly more efficient as it only requires one method dispatch and one check of the inputs:

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  get_vars(6:16) %>% 
  fprod(TRA = "/") %>% 
  head
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  get_vars(6:16) %>% 
  TRA(fprod(., keep.group_vars = FALSE), "/") %>% 
  head
```

This only demeans Agriculture (AGR) and Mining (MIN)

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  TRA(fselect(., AGR, MIN) %>% 
        fmean(keep.group_vars = FALSE), "-") %>% 
  head
```

`TRA` is best used in computations where grouped statistics are computed using some other function.

Same as above, with one line of code using fmean.data.frame and ftransform...

```{r}
GGDC10S %>% 
  ftransform(fmean(list(AGR = AGR, MIN = MIN),
                   list(Variable, Country),
                   TRA = "-")) %>% 
  head
```

Another potential use of TRA is to do computations in two- or more steps, for example if both aggregated and transformed data are needed, or if computations are more complex and involve other manipulations in-between the aggregating and sweeping part.

Get grouped tibble and then aggregate data

```{r}
gGGDC <- GGDC10S %>% 
  fgroup_by(Variable, Country)
gsumGGDC <- gGGDC %>% 
  fselect(AGR:SUM) %>% 
  fsum
head(gsumGGDC)
```

Get transformed (scaled) data

```{r}
head(TRA(gGGDC, gsumGGDC, "/"))
```

## Faster Centering, Averaging and Standardizing

The functions `fbetween` and `fwithin` are slightly more memory efficient implementations of `fmean` invoked with different `TRA` options

```{r}
rm(GGDC10S)
GGDC10S <- qTBL(GGDC10S)
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  get_vars(6:16) %>% 
  fbetween %>% 
  tail(2)
```

Same as.

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  get_vars(6:16) %>% 
  fmean(TRA = "replace") %>% 
  tail(2)
```

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  get_vars(6:16) %>% 
  fbetween(fill = TRUE) %>% 
  tail(2)
```

Same as

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  get_vars(6:16) %>% 
  fmean(TRA = "replace_fill") %>% 
  tail(2)
```

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  get_vars(6:16) %>% 
  fwithin %>% 
  tail(2)
```

Same as

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  get_vars(6:16) %>% 
  fmean(TRA = "-") %>% 
  tail(2)
```

Apart from higher speed, fwithin has a mean argument to assign an arbitrary mean to centered data, the default being mean = 0. A very common choice for such an added mean is just the overall mean of the data, which can be added in by invoking mean = "overall.mean":

```{r}
GGDC10S %>%
  fgroup_by(Variable, Country) %>%
  fselect(Country, Variable, AGR:SUM) %>%
  fwithin(mean = "overall.mean") %>%
  tail(3)
```

Use the `SUM` column as weights, for each variable and each group, subtract weighted mean, then add overall weighted column mean back to centered columns. `SUM` is kept as-is.

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  fselect(Country, Variable, AGR:SUM) %>% 
  fwithin(SUM, mean = "overall.mean") %>% 
  tail(3)
```

The `theta` parameter allows  partial- or quasi-demeaning operations, e.g. fwithin(gdata, theta = theta) is equal to gdata - theta * fbetween(gdata). This is particularly useful to prepare data for variance components (also known as ‘random-effects’) estimation.

`fscale` can be used to avoid sequential calls such as `... %>% fsd(TRA = "/") %>% fmean(TRA = "-")`

```{r}
GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  fselect(Country, Variable, AGR:SUM) %>% 
  fscale
```

The `mean` and `sd` arguments allow scaling to an arbitrary mean and standard deviation. `mean = FALSE` scales but preserves the means.

```{r}
gGGDC <- GGDC10S %>% 
  fgroup_by(Variable, Country) %>% 
  fselect(Country, Variable, AGR:SUM)
head(fmean(gGGDC))
```

Mean preserving scaling:

```{r}
gGGDC %>% 
  fscale(mean = FALSE) %>% 
  fmean %>% 
  head
```

```{r}
gGGDC %>% 
  fscale(mean = FALSE) %>% 
  fsd %>% 
  head
```

One can also set `mean = "overall.mean"`, which group-centers columns on the overall mean as illustrated with `fwithin` or `sd = "within.sd"`, which group-scales data such that every group has a standard deviation equal to the within-standard deviation of the data:

```{r}
gGGDC <- GGDC10S %>%
  fsubset(Variable == "VA", Country, AGR:SUM) %>% 
      fgroup_by(Country)
```

This calculates the within- standard deviation for all columns

```{r}
fwithin(gGGDC) %>% 
  ungroup %>% 
  num_vars %>% 
  fsd
```

This scales all groups to take on the within- standard deviation while preserving group means 

```{r}
fscale(gGGDC, mean = FALSE, sd = "within.sd") %>% 
  fsd
```

## Lags/Leads, Differences, Growth Rates

- `flag`
- `fdiff`
- `fgrowth`

The following code computes 1 fully-identified panel-lag and 1 fully identified panel-lead of each variable in the data.

```{r}
GGDC10S %>% 
  fselect(-Region, -Regioncode) %>% 
  fgroup_by(Variable, Country) %>% 
  flag(-1:1, Year)
```

It is also possible to omit the time-variable if one is certain that the data is sorted:

```{r}
GGDC10S %>%
  fselect(Variable, Country,AGR:SUM) %>%
  fgroup_by(Variable, Country) %>% 
  flag
```

`fdiff` computes sequences of lagged-leaded and iterated differences as well as quasi-differences and log-differences on time series and panel data. The code below computes the 1 and 10 year first and second differences of each variable in the data:

```{r}
GGDC10S %>% 
  fselect(-Region, -Regioncode) %>% 
  fgroup_by(Variable, Country) %>% 
  fdiff(c(1, 10), 1:2, Year)
```

Log-differences of the form $\log(x_t)−\log(x_{t−s})$ are also easily computed.

```{r}
GGDC10S %>% 
  fselect(-Region, -Regioncode) %>% 
  fgroup_by(Variable, Country) %>% 
  fdiff(c(1, 10), 1, Year, log = TRUE)
```

Finally, it is also possible to compute quasi-differences and quasi-log-differences of the form $x_t−\rho x_{t−s}$ or $\log(x_t)−\rho \log(x_{t−s})$.

```{r}
GGDC10S %>%
  fselect(-Region, -Regioncode) %>% 
  fgroup_by(Variable, Country) %>% 
  fdiff(t = Year, rho = 0.95)
```

Finally, `fgrowth` computes growth rates in the same way. By default exact growth rates are computed in percentage terms using $(x_t−x_{t−s})/x_{t−s} \times 100$ (the default argument is `scale = 100`). The user can also request growth rates obtained by log-differencing using $\log(x_t/x_{t−s})×100$.

```{r}
GGDC10S %>% 
  fselect(-Region, -Regioncode) %>% 
  fgroup_by(Variable, Country) %>% 
  fgrowth(c(1, 10), 1, Year)
```

This computes the 1 and 10-year growth rates, for the current period and lagged by one period

```{r}
GGDC10S %>% 
  fselect(-Region, -Regioncode) %>% 
  fgroup_by(Variable, Country) %>% 
  fgrowth(c(1, 10), 1, Year) %>% 
  flag(0:1, Year)
```

# Benchmarking

## Data

```{r}
rm(GGDC10S)
GRP(GGDC10S, ~ Variable + Country)
```

```{r}
data <- replicate(200, GGDC10S, simplify = FALSE)
uniquify <- function(x, i) ftransform(x, 
                                      lapply(unclass(x)[c(1,4)], paste0, i))
data <- unlist2d(Map(uniquify, data,
                     as.list(1:200)),
                 idcols = FALSE)
fdim(data)
GRP(data, ~ Variable + Country)
gc()
```

## Selecting, Subsetting, Ordering and Grouping

```{r}
microbenchmark(dplyr = select(GGDC10S, Country, Variable, AGR:SUM),
               collapse = fselect(GGDC10S, Country, Variable, AGR:SUM))
```

```{r}
microbenchmark(dplyr = select(data, Country, Variable, AGR:SUM),
               collapse = fselect(data, Country, Variable, AGR:SUM))
```

```{r}
microbenchmark(dplyr = filter(GGDC10S, Variable == "VA"),
               collapse = fsubset(GGDC10S, Variable == "VA"))
microbenchmark(dplyr = filter(data, Variable == "VA"),
               collapse = fsubset(data, Variable == "VA"))
```

```{r}
microbenchmark(dplyr = arrange(GGDC10S, desc(Country), Variable, Year),
               collapse = roworder(GGDC10S, -Country, Variable, Year))
microbenchmark(dplyr = arrange(data, desc(Country), Variable, Year),
               collapse = roworder(data, -Country, Variable, Year), times = 2)
```

```{r}
microbenchmark(dplyr = group_by(GGDC10S, Country, Variable),
               collapse = fgroup_by(GGDC10S, Country, Variable))
microbenchmark(dplyr = group_by(data, Country, Variable),
               collapse = fgroup_by(data, Country, Variable), times = 10)
```




